%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Missing Semester MIT Adaptation
% Author: Dr.-Ing. Mark Schutera
% Based on Tufte-style layout
\makeindex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass{../sharedAssets/tufte}

%%
% Book metadata
\title{Full Stack Handwerk}
\author[\defaultauthor]{\defaultauthor}
\publisher[\defaultpublisher]{\defaultpublisher}

% Load optional fonts
\loadoptionalfonts

% Additional packages
\usepackage{pifont}
\usepackage{amsmath}
\usepackage{soul}
\usepackage{textcase}
\usepackage{enumitem}
\usepackage{tikz}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usepackage{float}
\usepackage{algorithm}
\usepackage{algpseudocode}

\begin{document}



\makestandardfrontmatter
% Dedication
% Alternative quotes:
\makededication{\openepigraph{“Seine Wort' und Werke
Merkt ich und den Brauch,\\
und mit Geistesstärke
tu ich Wunder auch.”}{Der Zauberlehrling}}
\tableofcontents


% Introduction
\makeintroduction{%
  \newthought{This is an adaptation} of the MIT Missing Semester course material. And oh boy, we need it here in Germany as well! In the Länd of the 
  "Ingenieure", "Handwerk", and "Lehrlings" - we forgot the importance of practice and mastering our tooling skills. 
  \marginnote{Original course URL \url{https://missing.csail.mit.edu/}}
  }
  
  \newthought{As a mechanical engineer by training}, I have always had this nagging feeling that something important was missing from my education.
  The more I transitioned into software engineering roles, the more this feeling intensified.
  Looking at your curriculum, there is no shortage of advanced concepts and theories to learn. This one fills the gap
  that many computer science programs overlook: practical proficiency with essential tools and workflows that every software engineer relies on.
  And truth be told, Professors rely on students having this basic proficiency already acquired, or to figure this stuff out on their own - which is fair in academia.
  For me, software engineering is also a practical discipline, and ought to be taught as such. I need you to know this stuff, so it is also fair to teach it.
   
  \newthought{This course will teach you} everything you need to know to become proficient with the command line, version control systems like Git,
  text editors, shell scripting, and other essential tools that are the backbone of modern software development - in the end you will be able to develop and deploy 
  your software solutions. This will make you a better software engineer, and it will make your life easier latest when you need to go full ML Lifecycle.

%% Start the main matter
\mainmatter

% ----------------------------------------------------------------------
% 10-LECTURE STRUCTURE: FULL-STACK ML LIFECYCLE
% Goal: students can develop and deploy an ML product across VMs
% Each lecture: ~45 min general full-stack tooling + ~45 min ML lifecycle application
% Structure: Introduction (Why, Preview, Objectives) → Theory → Exercises & Reflection
% ----------------------------------------------------------------------

% LECTURE 0: Projektbericht — Course Work Requirements
\input{chapters/00_projektbericht.tex}

% LECTURE 1: Set Up the Lab
\input{chapters/01_servers_vms_linux.tex}

% LECTURE 2: Track the Code
\input{chapters/02_git_github.tex}

% LECTURE 3: Package the Stack
% \input{chapters/03_dependencies_structure.tex}

% LECTURE 4: Integrate the Model
%\input{chapters/04_model_integration.tex}

% LECTURE 5: Build the Backend
%\input{chapters/05_apis_services.tex}

% LECTURE 6: Store the Data
%\input{chapters/06_databases_sql.tex}

% LECTURE 7: Expose to the World
%\input{chapters/07_frontend_nginx.tex}

% LECTURE 8: Ship It
%\input{chapters/08_cicd_automation.tex}

% LECTURE 9: Watch It Run
%\input{chapters/09_monitoring_logging.tex}

% LECTURE 10: Vibe Coding
%\input{chapters/10_ai_dev_vibecoding.tex}


\nocite{TufteVisuals2001}
\makestandardbackmatter{missingsemester}

\end{document}
