% !TEX root = ../notes_missingsemester.tex
% Chapter 2: Version Everything — Git & GitHub
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% POST IMPROVEMENTS 2026:
% 1. Add a figure of the commit graph showing a feature branch merging back into main.
% 2. Record a short walkthrough of writing, amending, and squashing a commit; link in a margin note.
% 3. Add a worked example of rotating a secret that was committed, for the .gitignore section.

\index{Git}\index{GitHub}\index{version control}
\chapter{Version Everything: Git \& GitHub}
\newthought{Git a life.}

% ==============================================================================================
% BLOCK 1: INTRODUCTION
% ==============================================================================================

\section{The Why}\index{version control}\index{Git}

Software does not stop changing. Requirements shift, collaborators join, bugs appear,
and experiments turn into features that later have to be unwound.
Without a deliberate record of what changed, when, and why,
the past becomes guesswork and rollback becomes impossible.
This chapter introduces the tool every working codebase uses
to keep the past reachable and the future safe to explore.

\marginnote{\newthought{PixelWise} continues to be the narrative thread.
Block~1 left two empty virtual machines on your laptop;
by the end of this block, one of them hosts a \texttt{pixelwise} repository
tracked by Git, published on GitHub, and cloned onto the server at \texttt{/opt/pixelwise}.}

\newthought{Across the ten lecture blocks we build PixelWise}\index{PixelWise},
the full stack of tools we introduce sits on top of version control from block~2 onward.
Every configuration file, every script, and every line of backend code
is committed, pushed, and recoverable.
The server setup we ran manually in block~1 becomes a script
that any fresh VM can replay end to end.
What was ephemeral knowledge in a single terminal turns into durable, auditable state
that moves between machines without anybody having to remember what they typed.

\subsection{Hands On Experience}

\newthought{You have been there.} A folder full of copies: \texttt{report.py}, \texttt{report\_v2.py},
\texttt{report\_final.py}, \texttt{report\_FINAL\_fixed.py}.
Which one is current? Which one works? What changed between them?
Nobody knows, including you, two weeks from now.

\marginnote{\newthought{Count the time} you have lost
to guessing which file was current, reconstructing a change you made yesterday,
or restoring a working version from a zip you made ``just in case''.
Version control removes the class of problem, not a single instance of it.}

\newthought{Version control solves this permanently.}
Every change is recorded, annotated, and reversible;
you can see exactly what changed, when, by whom, and why;
you can return to any previous state;
and you can work on multiple ideas in parallel without them interfering with each other.
Git is the version control system we adopt, and it is the de facto standard for the rest of the industry.

\newthought{This second block} focuses on Git and GitHub rather than on application code.
Before we write a line of backend logic, every artefact we produce from here on must be tracked,
reviewable, and reproducible on a new machine.
By the end of the block you will have
\begin{itemize}
    \item initialised a local Git repository for PixelWise on the \texttt{dev} VM,
    \item captured the manual server setup from block~1 in a reproducible shell script,
    \item published the repository on GitHub over SSH, and
    \item cloned the repository onto the \texttt{prod} VM,
          so both machines share a single source of truth.
\end{itemize}



% ==============================================================================================
% BLOCK 2: MAIN THEORY
% ==============================================================================================

\newpage
\section{The Git Data Model}\index{Git!data model}

\newthought{Worry not,} git gets easier once you get the basic idea that branches are homeomorphic
endofunctors mapping submanifolds of a Hilbert space, or so someone on Reddit said.
In practice the data model is a lot more approachable than the joke suggests,
and the rest of this section spells it out.

\newthought{Git is not a backup tool.}
It is a content-addressable filesystem with a version history layered on top,
and understanding that model makes every command intuitive
rather than a sequence of incantations.
Under the cover of \texttt{git add} and \texttt{git commit}
there are four kinds of objects and two pointers
that together express every state a repository has ever been in.

\marginnote{\newthought{Git}\index{Git} at \url{https://git-scm.com/}
was created by Linus Torvalds in 2005 to manage the Linux kernel source code
after the previous tool's licence terms changed.
It is now the de facto standard for version control across the industry.\\
Pro Git book, free: \url{https://git-scm.com/book/en/v2}.}

\begin{description}
    \item[Blob]\index{Git!blob} A file's contents, stored and addressed by the SHA-1 hash of those contents.
    The name is irrelevant to Git; only the bytes matter.
    \item[Tree]\index{Git!tree} A directory listing that maps filenames to blobs for files
    and to other trees for subdirectories.
    \item[Commit]\index{Git!commit} A snapshot of the entire project at a point in time,
    combining a pointer to the root tree, the author, a timestamp, a message,
    and a pointer to the parent commit or commits.
    \item[Branch]\index{Git!branch} A lightweight, movable pointer to a commit.
    \texttt{main} is a branch. Creating one is instantaneous because it only creates a new pointer.
    \item[HEAD]\index{Git!HEAD} A pointer to the branch you are currently on;
    when you commit, \texttt{HEAD} moves forward along with the branch it points to.
\end{description}

\newthought{Every commit is identified by a SHA-1 hash},
a 40-character hexadecimal string such as \texttt{a3f2b7c\ldots}
computed from the commit's contents.
If anything changes, a single character in a single file, the hash changes,
which is what makes the history tamper-proof
and why two repositories with the same hashes are known to contain the same objects.

\marginnote{\newthought{Content-addressable storage}\index{content-addressable storage}
is the idea that an object's identifier is a function of its bytes.
Git inherits this pattern from earlier systems such as Monotone and the Plan~9 Venti archival store;
Docker image layers and IPFS use the same trick.
If you know the identifier, you already know whether you have the right object.}



% ==============================================================================================
\newpage
\section{The Staging Workflow}\index{Git!staging}

\newthought{Git has three areas} in which your changes live
before they become part of the permanent history.

\begin{description}
    \item[Working directory] The files on disk, what your editor reads and writes.
    \item[Staging area]\index{Git!index} Sometimes called the index,
    the set of changes you have marked for the next commit.
    A deliberate selection, not a side effect of editing.
    \item[Repository] The committed history: permanent, immutable snapshots linked by parent pointers.
\end{description}

\newthought{A handful of commands} move changes between those three areas.

\begin{verbatim}
git status                 # what changed? what is staged?
git add file.py            # stage a file for the next commit
git add .                  # stage everything, use with care
git commit -m "message"    # commit the staged changes
git diff                   # unstaged changes vs last commit
git diff --staged          # staged changes vs last commit
git log                    # show commit history
git log --oneline          # compact view, one line per commit
\end{verbatim}

\marginnote{\newthought{Write commit messages in the imperative mood}:
``Add classifier module'' rather than ``Added classifier module.''
The message completes the sentence ``This commit will \ldots''.
Keep the first line under 72 characters
and leave a blank line before any longer explanation.}

\newthought{The separation between staging and committing} is what gives Git its fine-grained control.
A single working session might touch five files,
and you can group them into two or three commits that each tell a coherent story,
rather than one commit that mixes unrelated changes.
In practice this means the staging area is where you compose the history you want to keep,
not just a technical step on the way to \texttt{git commit}.



% ==============================================================================================
\newpage
\section{The \texttt{.gitignore} File}\index{.gitignore}

\newthought{Not everything belongs in version control.}
Compiled byte code is regenerated on every run,
virtual environments are rebuilt from a lockfile,
trained model artefacts are larger than a repository should carry,
and secrets must never be tracked at all.
The \texttt{.gitignore} file tells Git which paths to exclude from tracking:

\begin{verbatim}
# Python
__pycache__/
*.pyc
.venv/

# Secrets
.env

# Data and models, too large and reproducible
data/
models/*.pkl

# Editor files
.vscode/
*.swp
\end{verbatim}

\marginnote{\newthought{Git history is permanent.}\index{Git!history}
A secret committed once is leaked forever,
even if you delete the file in a later commit:
the old commit still contains it, and any clone carries the full history.
\texttt{.gitignore} is the first line of defence,
and it only works when the entry is in place before the first commit
that would otherwise include the file.}

\newthought{Treat a committed secret as compromised.}\index{secrets}
If a password, API key, or private key has entered a commit,
assume somebody now holds it and rotate it immediately.
Rewriting history with \texttt{git filter-branch}\index{git filter-branch}
or BFG Repo-Cleaner\index{BFG Repo-Cleaner} at \url{https://rtyley.github.io/bfg-repo-cleaner/}
removes the secret from future clones,
but it does not recall any clone that already exists.
The principle of least privilege\index{least privilege} from block~1 has its version control counterpart here:
the narrower the set of files Git tracks, the smaller the blast radius of a mistake.



% ==============================================================================================
\newpage
\section{Branching and Merging}\index{Git!branching}\index{Git!merging}

\newthought{Branches let you work on ideas in isolation.}
The \texttt{main} branch is the stable baseline that always builds and always runs.
A feature branch diverges from \texttt{main} at a specific commit,
evolves through any number of commits of its own,
and merges back once the work is ready.

\begin{verbatim}
git branch                    # list branches
git branch feature-x          # create a new branch
git checkout feature-x        # switch to it
git checkout -b feature-y     # create and switch in one step
\end{verbatim}

\newthought{When the feature is ready,} you merge it back into the baseline:

\begin{verbatim}
git checkout main
git merge feature-x
\end{verbatim}

\newthought{A merge conflict}\index{Git!merge conflict} occurs
when two branches changed the same line of the same file
and Git cannot decide which version to keep.
Rather than guessing, it marks the conflict in the file and asks you to resolve it:

\begin{verbatim}
<<<<<<< HEAD
result = model.predict(x)
=======
result = classifier.classify(x)
>>>>>>> feature-x
\end{verbatim}

Edit the file to keep the correct version, remove the markers,
then stage and commit the resolution:

\begin{verbatim}
git add file.py
git commit -m "Resolve merge conflict in file.py"
\end{verbatim}

\marginnote{\newthought{A conflict is not a failure.}
It is Git refusing to invent an answer you did not give it.
Treat conflicts as a prompt to re-read both sides,
understand which change should survive, and write the resolution
as a small, explicit commit of its own.}



% ==============================================================================================
\newpage
\section{Tags}\index{Git!tags}

\newthought{A tag marks a point in history that matters},
a release, a milestone, or a known-good state you may need to return to.
Unlike a branch, a tag does not move once it is created.

\begin{verbatim}
git tag v0.1                             # lightweight tag
git tag -a v0.1 -m "initial setup"       # annotated tag, preferred
git push --tags                          # push tags to remote
\end{verbatim}

\marginnote{\newthought{Lightweight vs annotated tags.}
A lightweight tag is only a name pointing to a commit.
An annotated tag stores the tagger, the date, and a message,
and it is what releases should use.
We tag \texttt{v0.1} in this block;
the tag becomes meaningful when you look back at it in block~8,
after seven more iterations have rolled past it.}

\newthought{Semantic versioning}\index{semantic versioning}
is the convention we follow:
versions take the form \texttt{MAJOR.MINOR.PATCH},
and each component carries meaning about the kind of change it represents.
A \texttt{v0.x} tag signals that the project is still pre-release
and that the public interface can still change without warning,
which is exactly where PixelWise sits today.



% ==============================================================================================
\newpage
\section{Remotes and GitHub}\index{Git!remotes}\index{GitHub}

\newthought{A remote is a copy of your repository on another machine,}
usually a hosting service such as GitHub.
Remotes are how a repository crosses machines, how collaborators share work,
and how code travels from \texttt{dev} to \texttt{prod} later in the course.

\begin{verbatim}
git remote add origin git@github.com:user/pixelwise.git
git push -u origin main        # push and set upstream
git pull                       # fetch plus merge from remote
git fetch                      # fetch without merging
git clone <url>                # copy a remote repo locally
\end{verbatim}

\marginnote{\newthought{GitHub is not Git.}\index{GitHub}
Git is the version control system;
GitHub is a hosting platform that adds pull requests\index{pull request},
issues, CI/CD, and collaboration features on top of Git.\\
GitHub: \url{https://github.com/}\\
GitHub Docs: \url{https://docs.github.com/}}

\newthought{A pull request}\index{pull request}
is a proposal to merge a branch into another branch,
typically a feature branch into \texttt{main}.
It is the standard mechanism for code review:
changes are visible, reviewable, and discussable
before they become part of the main codebase.
A pull request is a GitHub feature, not a Git one;
at the Git level, the merge itself is what happens once the review passes.

\marginnote{\newthought{SSH again.}\index{SSH}
The remote URL \texttt{git@github.com:user/pixelwise.git}
uses the same SSH protocol you configured in block~1.
GitHub holds your public key; your private key stays on \texttt{dev}.
The key-based authentication that replaced passwords on \texttt{prod}
replaces them here as well,
and the habit of adding your public key to each new service generalises
to every remote system the rest of the course touches.}



% ==============================================================================================
\newpage
\section{Beyond GitHub}\index{Git!self-hosting}\index{GitLab}\index{Gitea}\index{Forgejo}

\newthought{GitHub is not the only option.}
Git is decentralised by design, and any machine that can accept SSH connections can host a repository.
Picking a platform is a trade-off between convenience, cost, and sovereignty over your data.
A short tour of the landscape keeps the choice explicit rather than accidental.

\begin{description}
    \item[GitLab]\index{GitLab} The full-featured GitHub alternative.
    Strong built-in CI/CD, issue boards, and release tooling.
    Runs as a hosted service at \texttt{gitlab.com} or as GitLab Community Edition on your own server.
    Choose it if you want one integrated DevOps stack.
    \item[Bitbucket]\index{Bitbucket} Part of the Atlassian ecosystem.
    Integrates tightly with Jira and Confluence,
    a natural fit for teams that already live in Atlassian tools.
    \item[Gitea]\index{Gitea} A lightweight, self-hosted forge written in Go.
    Runs on a Raspberry Pi, installs in minutes,
    and looks and feels like a stripped-down GitHub.
    Choose it when you want simple self-hosting without the resource weight of GitLab.
    \item[Forgejo]\index{Forgejo} A community-driven soft fork of Gitea.
    Looks, acts, and installs almost exactly like Gitea,
    but is governed as a non-profit community project with no for-profit company behind it.
    Choose it if you want the Gitea experience with a free-software purist stance.
    \item[Bare SSH repository] The minimal option.
    A directory created with \texttt{git init --bare} on any SSH-reachable machine is enough to collaborate:
    \texttt{git clone user@host:/path/to/repo.git}.
    No web UI, no pull requests, just Git.
\end{description}

\marginnote{\newthought{Where to look.}
Self-hosted and alternative platforms:\\
GitLab: \url{https://about.gitlab.com/}\\
Gitea: \url{https://about.gitea.com/}\\
Forgejo: \url{https://forgejo.org/}\\
Bitbucket: \url{https://bitbucket.org/}\\
Bare repo guide: \url{https://git-scm.com/book/en/v2/Git-on-the-Server-Getting-Git-on-a-Server}}

\newthought{The choice is reversible.}
Because the Git data model is independent of any platform,
a repository can be pushed to GitHub today, mirrored to GitLab tomorrow,
and archived on a bare SSH host the year after.
We use GitHub in this course for familiarity,
but nothing we learn here is specific to it.



% ==============================================================================================
\newpage
\section{The Setup Script}\index{setup-server.sh}

\newthought{On the server VM,} you installed packages manually in block~1
by typing commands into a shell.
Those commands worked, and then they were gone.
A fresh VM, a new teammate, or a rebuild after a failure
would need the same sequence run again,
and remembering which commands you ran last week is not a plan.
A short shell script captures the work once and reruns it on demand.

\begin{verbatim}
#!/bin/bash
# setup-server.sh — provision a fresh server VM
set -euo pipefail

sudo apt update
sudo apt install -y git python3 python3-pip python3-venv curl
\end{verbatim}

\marginnote{\newthought{The \texttt{set -euo pipefail} line}
makes the script fail immediately on any error with \texttt{-e},
on any undefined variable with \texttt{-u},
and on pipe failures with \texttt{-o pipefail}.
Without it, a failing \texttt{apt install} prints a message and the script continues anyway,
which is how servers end up in half-configured states that nobody can debug later.}

\newthought{This is the first file that justifies the whole chapter.}
It is small, it is boring,
and it turns a bundle of remembered terminal commands
into an artefact you can commit, tag, and reproduce on any new machine.
Every later block adds to it: a Python environment in block~3,
a systemd unit in block~5, a reverse proxy in block~7.
By block~8 the script provisions an entire service stack from nothing.





% ==============================================================================================
% BLOCK 3: STUDENT ACTIVATION
% ==============================================================================================

\newpage
\section{Exercise: Initialising the Repository}\index{exercises}

\marginnote{\newthought{Exercises} are for practice and reinforcing concepts.
Work through them yourself first, break things, discuss; this is not a time trial.
If your VM ends up in a state you cannot recover,
restore the \texttt{B1-complete} snapshot from block~1 and start over.
At the end of the session take a snapshot named \texttt{B2-complete};
it is the starting state for block~3.}

\newthought{Block~2 introduced seven ideas} in the theory section:
the Git data model, the staging workflow, \texttt{.gitignore},
branching and merging, tags, remotes and GitHub, and the setup script.
Each exercise below mirrors one of them in the same order,
so you can read a section and immediately practise what it described.

\marginnote{\newthought{Install Git first.}
On both VMs run \texttt{sudo apt update \&\& sudo apt install -y git}
so the commands below have something to execute.
You will capture this install step properly in the final exercise,
together with the rest of the server's packages.}

\newthought{Configure your Git identity} on \texttt{dev}
so every commit carries your name and email address:

\begin{verbatim}
git config --global user.name "Your Name"
git config --global user.email "your@email.com"
\end{verbatim}

\newthought{Create the project directory} and initialise a repository:

\begin{verbatim}
mkdir ~/pixelwise && cd ~/pixelwise
git init
\end{verbatim}

\marginnote{\newthought{Look inside \texttt{.git/}.}
\texttt{ls -la .git} shows the on-disk form of the data model from block~2:
\texttt{objects/} for blobs and trees,
\texttt{refs/} for branches and tags,
\texttt{HEAD} as a plain file that names the current branch.
Every Git command you run from here on manipulates exactly these files.}

Run \texttt{git status}; Git should report that you are on branch \texttt{main}
with no commits yet.



% ==============================================================================================
\newpage
\section{Exercise: Staging and Committing}\index{exercises}

\newthought{Write a \texttt{README.md}} with a one-line description of what PixelWise is.
Stage it, inspect the status, commit, and read the resulting history:

\begin{verbatim}
git add README.md
git status
git commit -m "Add README with project description"
git log --oneline
\end{verbatim}

\marginnote{\newthought{Read the hash} Git prints after the commit.
That seven-character prefix is the start of the SHA-1 from block~2,
and any unambiguous prefix of a commit's hash refers to that commit.
Running \texttt{git log} later should always show this commit at the root of the history.}

\newthought{Watch the three areas.}
Edit \texttt{README.md}, add a second line,
and run \texttt{git diff} before staging and \texttt{git diff --staged} after:

\begin{verbatim}
git diff
git add README.md
git diff --staged
git commit -m "Expand README with a second line"
\end{verbatim}

Naming the difference between the two diff outputs aloud
is the fastest way to make the working directory, staging area,
and repository stop feeling abstract.



% ==============================================================================================
\newpage
\section{Exercise: Guarding Secrets}\index{exercises}

\newthought{Write a \texttt{.gitignore}} that covers the paths
you already know should stay out of version control:

\begin{verbatim}
# Python
__pycache__/
*.pyc
.venv/

# Secrets
.env

# Data and models
data/
models/*.pkl

# Editor files
.vscode/
*.swp
\end{verbatim}

\newthought{Test that it works.}
Create a dummy \texttt{.env} with a fake secret inside and run \texttt{git status}:

\begin{verbatim}
echo "API_KEY=not-a-real-secret" > .env
git status
\end{verbatim}

\marginnote{\newthought{What you expect to see.}
\texttt{.env} does not appear in the status output
because \texttt{.gitignore} already excluded it.
If it does appear, the \texttt{.gitignore} was written after the file was tracked,
and \texttt{git rm --cached .env} is the fix.
This is why \texttt{.gitignore} is the very first configuration file to commit in a new repository.}

Stage and commit \texttt{.gitignore} itself so the rule
is part of the repository from the start:

\begin{verbatim}
git add .gitignore
git commit -m "Add gitignore for Python, secrets, data, and editors"
\end{verbatim}



% ==============================================================================================
\newpage
\section{Exercise: A Feature Branch}\index{exercises}

\newthought{Warm up in a sandbox first.}\index{Learn Git Branching}\index{Oh My Git!}
Before touching the real repository, complete the introduction sequence
on the \texttt{Main} tab at Learn Git Branching: \url{https://learngitbranching.js.org/}.
The four interactive puzzles walk you through commits, branches, merges, and rebases
with a live commit graph next to the command line,
which is the most direct way to make the data model from block~2 concrete.
Finish at least \texttt{Introduction to Git Commits},
\texttt{Branching in Git},
\texttt{Merging in Git},
and \texttt{Rebase Introduction} before you continue;
each one takes a few minutes and resets with a click if you get stuck.

\marginnote{\newthought{Two more sandboxes}
worth keeping open while you work:\\
Git Exercises: \url{https://gitexercises.fracz.com/}\\
Oh My Git!: \url{https://ohmygit.org/}\\
Git Exercises poses command-line puzzles against a real Git clone,
while Oh My Git! is a card game that shows the graph changing as you play.
Pick the format that matches how you learn,
but do not skip this step; branching is the one Git topic
that fluency in a sandbox transfers directly to.}

\newthought{Now create a branch on \texttt{pixelwise}} to extend the README
without touching \texttt{main}:

\begin{verbatim}
cd ~/pixelwise
git checkout -b improve-readme
\end{verbatim}

Add a short section to \texttt{README.md}
describing the two-machine setup from block~1, then commit:

\begin{verbatim}
git add README.md
git commit -m "Describe two-machine setup in README"
\end{verbatim}

\newthought{Merge back into \texttt{main}:}

\begin{verbatim}
git checkout main
git merge improve-readme
git log --oneline --graph --all
\end{verbatim}

\marginnote{\newthought{Read the graph.}
\texttt{git log --oneline --graph --all} draws the commit graph as ASCII art.
With one feature branch that merged cleanly the graph is a straight line,
but it is the same visualisation you saw in the Learn Git Branching puzzles,
and the same one you will use to untangle complicated merges later.}

\newthought{Provoke and resolve a conflict.}
Back in a browser, go to the \texttt{Remote} tab of Learn Git Branching
and complete \texttt{Push \& Pull -- Git Remotes!} and \texttt{To Origin And Beyond --
Advanced Git Remotes!}, then try the \texttt{Mixed Bag} puzzles that require rebasing a conflict.
The sandbox shows every intermediate state of the graph,
so when the same situation shows up on \texttt{pixelwise} you already know
which commit is which and what the resolution should look like.

\newthought{Delete the merged branch} once the work is on \texttt{main},
so the branch list stays a view of active work rather than a graveyard:

\begin{verbatim}
git branch -d improve-readme
\end{verbatim}



% ==============================================================================================
\newpage
\section{Exercise: Tagging \texttt{v0.1}}\index{exercises}

\newthought{Mark this point in history} with an annotated tag:

\begin{verbatim}
git tag -a v0.1 -m "Initial repository with README and gitignore"
git log --oneline --decorate
\end{verbatim}

The \texttt{--decorate} flag shows branches and tags alongside each commit,
so \texttt{v0.1} is visible on the commit it points to.

\marginnote{\newthought{Pushing tags is separate.}
A regular \texttt{git push} does not include tags by default.
You will use \texttt{git push --tags} in the next exercise
once the repository has a remote, which is deliberate:
tags name releases, and releases are a conscious act.}

\newthought{Inspect what the tag holds:}

\begin{verbatim}
git show v0.1
\end{verbatim}

The output names the tagger, the date, the annotation message,
and the commit the tag points at.
A lightweight tag would show only the commit;
the extra metadata is exactly what makes annotated tags appropriate for releases.



% ==============================================================================================
\newpage
\section{Exercise: Publishing to GitHub}\index{exercises}

\newthought{Create a GitHub account} if you do not have one,
and add your \texttt{dev} VM's SSH public key under
\texttt{Settings, SSH and GPG keys}.
The public key is the one in \texttt{\~{}/.ssh/id\_ed25519.pub} on \texttt{dev};
print it with \texttt{cat} and paste the contents into GitHub.

\marginnote{\newthought{Same key, new service.}\index{SSH}
You are reusing the SSH key generated in block~1 rather than creating a new one.
One key per machine is enough;
you add the public half to every service
that should accept connections from \texttt{dev}.
The private half never leaves the machine it was generated on.}

\newthought{Create an empty repository} on GitHub named \texttt{pixelwise}
with no README, no \texttt{.gitignore}, and no licence,
since you already have all three locally.
Then wire the local repository up and push:

\begin{verbatim}
git remote add origin git@github.com:your-user/pixelwise.git
git push -u origin main
git push --tags
\end{verbatim}

\newthought{Open a pull request} to experience the review flow on your own repository.
On GitHub, edit the README through the web interface on a new branch called \texttt{readme-tweak},
commit the change, and click \texttt{Compare \& pull request}.
Review the diff, merge the pull request, then pull the result back to \texttt{dev}:

\begin{verbatim}
git checkout main
git pull
\end{verbatim}

\marginnote{\newthought{What a pull request does.}\index{pull request}
The pull request renders the diff, collects reviewer comments,
and runs any configured checks before the merge button lights up.
For a solo push this is ceremony;
for a team repository it is where most of the engineering discussion happens,
and the habit of opening one is worth forming while the stakes are still low.}

\newthought{Contribute a pull request to an open-source project} before the next class.
Pick any project on GitHub, GitLab, or a self-hosted forge and submit a pull request.
It can be a typo fix in the README, a clarifying sentence in the documentation,
a missing example, or a small bug fix.
The goal is to practise the full collaboration loop end to end:
fork, clone, branch, commit, push, open the PR, and respond to review comments.
Projects you already use are the best starting point;
otherwise browse curated lists of beginner-friendly issues:

\begin{verbatim}
https://goodfirstissue.dev/
https://github.com/MunGell/awesome-for-beginners
\end{verbatim}

Bring the PR link to the next session, whether it has been merged or not.



% ==============================================================================================
\newpage
\section{Exercise: The Reproducible Setup Script}\index{exercises}

\newthought{SSH into the server VM} and install the tools the rest of the course needs:

\begin{verbatim}
sudo apt update
sudo apt install -y git python3 python3-pip \
    python3-venv curl
\end{verbatim}

Verify the install with \texttt{git --version} and \texttt{python3 --version}.

\newthought{Back on \texttt{dev}}, capture exactly what you just ran in a script.
Inside the \texttt{pixelwise} repository, create \texttt{scripts/setup-server.sh}:

\begin{verbatim}
#!/bin/bash
# setup-server.sh — provision a fresh server VM
set -euo pipefail

sudo apt update
sudo apt install -y git python3 python3-pip python3-venv curl
\end{verbatim}

\marginnote{\newthought{Why a script, not a note.}
A note tells a human what to do; a script does it.
As PixelWise grows the script grows with it,
and the fact that it always ran to completion
is the evidence that a fresh VM really does reach the state
the rest of the course assumes.}

\newthought{Make it executable, commit, and push:}

\begin{verbatim}
chmod +x scripts/setup-server.sh
git add scripts/setup-server.sh
git commit -m "Add reproducible server setup script"
git push
\end{verbatim}

\newthought{Clone the repository onto the server}
so both machines share the same source:

\begin{verbatim}
sudo mkdir -p /opt/pixelwise
sudo chown produser:produser /opt/pixelwise
git clone git@github.com:your-user/pixelwise.git /opt/pixelwise
\end{verbatim}

The server now carries the repository at \texttt{/opt/pixelwise},
and any \texttt{git pull} from there brings in changes you push from \texttt{dev}.

\newthought{Take a snapshot} of both VMs and name them \texttt{B2-complete}.
From here on, every new artefact lives inside the repository,
and the snapshot is your fallback if something in block~3 goes sideways.



% ============================= CODE ===============================
\newpage

\marginnote{\newthought{A note on what comes next.}
The repository is in place, but it only holds configuration so far.
Block~3 adds a Python environment, a package layout,
and the first PixelWise source file,
all of which will be staged, committed, and pushed
through the workflow you just practised.}



\section{Self-Reflection and Recap}\index{reflection}\index{recap}

\newthought{Self-Reflection} Questions which can guide your thoughts during and after the exercises:
\begin{itemize}
    \item What is the difference between \texttt{git add} and \texttt{git commit},
          and why is the separation useful rather than an extra step?
    \item Why does a SHA-1 hash change when any byte of a file changes,
          and what guarantee does that give you about the history?
    \item Why should you never commit \texttt{.env} files or API keys,
          and what happens to a secret that was committed and then deleted in a later commit?
    \item When would you use a branch instead of committing directly to \texttt{main}?
    \item What is the difference between a lightweight tag and an annotated tag,
          and which one belongs on a release?
    \item How does a pull request differ from \texttt{git merge},
          and which one does which part of the job?
    \item When would you reach for GitLab, Gitea, Forgejo, or a bare SSH repository
          instead of GitHub, and what changes about the workflow if you do?
    \item Why do we capture the server setup in a script
          rather than running the same commands manually on the next fresh VM?
\end{itemize}

\newthought{Recap} of Key Concepts:
\begin{itemize}
    \item Git stores content as blobs, grouped into trees,
          pointed at by commits, with branches as movable pointers
          and \texttt{HEAD} as the current one.
    \item The staging area is where you compose the next commit,
          rather than a technical hurdle before it.
    \item \texttt{.gitignore} keeps byproducts and secrets out of the history,
          and Git history is permanent enough
          that rotation, not deletion, is the right response to a leak.
    \item Branches isolate parallel work,
          and merges, pull requests, and conflicts
          are the means of bringing that work back together.
    \item Tags name points in history, and annotated tags are the right tool for releases.
    \item Remotes put your repository on another machine,
          and GitHub, GitLab, Gitea, Forgejo, and bare SSH hosts
          are interchangeable options for where that machine lives.
    \item A setup script turns remembered manual work into a reproducible artefact,
          and it is the first load-bearing file in PixelWise.
\end{itemize}

\marginnote{\newthought{Without shipping anything yet,}
PixelWise already has a single source of truth that lives on both \texttt{dev} and \texttt{prod},
and a reproducible path from an empty VM to a configured one.}

\marginnote{\newthought{Teaser.}
The repository holds configuration and a script, but no application code.
How do we turn an empty Python interpreter on \texttt{dev}
into a project with pinned dependencies, a package layout,
and a first module that the repository can track end to end?}

\newthought{So far, we own the history.}
The next step is to start filling it with code.
In the following chapter we create a Python environment on \texttt{dev},
lay out the PixelWise package, and commit the first real module,
so the repository starts doing work rather than just holding configuration.
