% !TEX root = ../notes_missingsemester.tex
% Chapter 4: Integrate the Model
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% POST IMPROVEMENTS 2026:
% 1. Replace the inline ASCII pipeline with a TikZ figure of the inference path.
% 2. Provide a slim notebook that visualises binarization side by side.

\index{model integration}\index{classifier}\index{MNIST}\index{inference}\index{joblib}
\chapter{Integrate the Model}
\newthought{Your ML course teaches you to train models. This block teaches you to ship one.}

% ==============================================================================================
% BLOCK 1: INTRODUCTION
% ==============================================================================================

\section{The Why}\index{model artefact}\index{train/serve skew}

Block~3 left us with a reproducible environment on both machines.
The virtual environment is ready, the dependencies are pinned,
the project structure is in place.
What is missing is the part of the system that does any actual work:
PixelWise still cannot classify anything, because the \texttt{app/} folder
is empty.

\marginnote{\newthought{PixelWise} now has a scaffold, pinned dependencies,
a provisioned server, and a \texttt{.env.example} that documents its
configuration contract.
But the \texttt{app/} directory is empty.
Before we can build an API or a frontend, we need the core capability:
a trained model that can classify a handwritten digit.}

\newthought{Across ten lecture blocks we build PixelWise}\index{PixelWise}.
This block is the first that produces application logic.
Everything before it was foundation: a place to put code, a way to track it,
a reproducible environment to run it in.
Everything after it depends on the inference interface produced here.

\newthought{The gap between a trained model and a working system}
is larger than most students expect.
A \texttt{.pkl}\index{.pkl} file sitting in a folder is not a product.
It needs to be loaded safely, called with the exact input format it was
trained on, and wrapped in an interface the rest of the application can
depend on.
Any of these can be wrong without producing an error: the model returns a
prediction, the prediction is plausible, the prediction is wrong.

\marginnote{\newthought{Silent failure}\index{silent failure} is the signature
bug of machine-learning systems.
Code that crashes is recoverable; you read the trace and fix the line.
Code that returns a wrong answer with the same confidence as a right one
goes unnoticed until somebody notices the impact downstream.
Most of the engineering in this block exists to turn silent failures into
loud ones at the earliest possible moment.}

\newthought{This block bridges that gap.}
We take a pre-trained digit classifier, integrate it into the PixelWise
codebase, and verify end to end that it works before any API or frontend
code exists.
The pipeline you build here, model file in, prediction dictionary out,
is the contract every later block consumes without modification.

\begin{figure}
\centering
\begin{tikzpicture}[
    every node/.style={font=\footnotesize, align=center},
    box/.style={rectangle, draw=black, thick, rounded corners=2pt, minimum width=1.9cm, minimum height=0.9cm, inner sep=4pt},
    arr/.style={-latex, thick, black!70}
]
\node[box] (img)  at (0, 0)    {input\\\texttt{(28,28)} uint8};
\node[box] (bin)  at (2.4, 0)  {binarize\\\texttt{>128}};
\node[box] (flat) at (4.8, 0)  {flatten\\\texttt{(1,784)}};
\node[box] (mod)  at (7.4, 0)  {pipeline\\\texttt{predict\_proba}};
\node[box] (out)  at (10.0, 0) {result\\\texttt{dict}};
\draw[arr] (img.east)  -- (bin.west);
\draw[arr] (bin.east)  -- (flat.west);
\draw[arr] (flat.east) -- (mod.west);
\draw[arr] (mod.east)  -- (out.west);
\end{tikzpicture}
\caption{The inference pipeline for one image. The same pipeline scales to a batch by stacking inputs into an $(N, 28, 28)$ array; \texttt{predict\_proba} is vectorised, so the cost per image stays nearly constant as $N$ grows.}
\label{fig:inference_pipeline}
\end{figure}

\subsection{Hands On Experience}

\newthought{Consider this scenario:} you download a model file from a
colleague, drop it into your project, and write a quick script that calls
\texttt{model.predict(image)}.
It runs without errors.
The predictions look reasonable.
You ship it.
Two weeks later a user reports that the app never predicts the digit 0.
You check: the model was trained on digits 1 to 9 only.
Nobody documented the omission.
Nobody checked.

\marginnote{\newthought{A silent failure} is one the program never reports.
Unlike a crash, it has no log line, no stack trace, no obvious moment of
failure.
The only way to find it is to compare the output against an expected value,
and the only way to do that systematically is to write the expectation down
somewhere the program can check.}

\newthought{The model contract} is the agreement between the training script
that produced the file and the inference code that loads it.
The contract has three parts: which classes the model predicts,
what shape and dtype the inputs must take, and what preprocessing has
already been baked in.
Honouring the contract is the difference between a service that classifies
digits and a service that returns plausible nonsense.

\marginnote{\newthought{The inference interface} you build today is consumed,
unchanged, by Block~5's API, Block~6's database writer, Block~7's frontend,
Block~8's CI pipeline, and Block~9's analytics.
Five later blocks read from the function you write this afternoon.
Designing it well is leverage.}

\newthought{This fourth block} integrates a trained model into PixelWise
and builds the inference layer.
By the end you will have
\begin{itemize}
    \item understood what a model file is and why it is not source code,
    \item loaded the model safely and verified its class contract at startup,
    \item implemented a batch-first inference interface in \texttt{app/classifier.py},
    \item written a smoke test that proves end-to-end correctness on real samples,
    \item and documented the model's capabilities and limitations in a model card.
\end{itemize}



% ==============================================================================================
% BLOCK 2: MAIN THEORY
% ==============================================================================================

\newpage
\section{What a Model File Actually Is}\index{pickle}\index{joblib}\index{.pkl}

\newthought{A \texttt{.pkl} file is a build artefact,} not source code.
You distribute it; you do not commit it.
Just as a compiled binary is produced by a compiler, a model file is produced
by a training script.
The training script is the source of truth: it takes data, runs an algorithm,
and writes weights, hyperparameters, and any preprocessing into a single
serialised object.

\marginnote{\newthought{The course provides}
\texttt{models/digit\_classifier\_v1.pkl}\index{LogisticRegression},
a LogisticRegression pipeline trained on MNIST digits 1 to 9
(9 classes, ${\sim}54{,}000$ training samples, public domain).
Class 0 is intentionally held back; students can add it later without
collecting any new data.}

\newthought{Inspect the artefact} in a Python shell to see what is inside:

\begin{verbatim}
import joblib
model = joblib.load("models/digit_classifier_v1.pkl")
print(type(model))
# <class 'sklearn.pipeline.Pipeline'>
print(model.classes_)
# ['1' '2' '3' '4' '5' '6' '7' '8' '9']
\end{verbatim}

The object that comes back is a full sklearn Pipeline\index{sklearn!Pipeline},
not just a weights tensor.
It carries the preprocessing, the classifier, and the class labels in one
package, so loading the file is enough to start predicting.

\marginnote{\newthought{The training script}\index{train.py}
\texttt{train.py} is provided as supplementary material.
Students who want to understand how the artefact was produced can run it:
MNIST downloads automatically via \texttt{sklearn.datasets.fetch\_openml},
and training takes ${\sim}30$ seconds on CPU.
Further reading on training practices:\\
Karpathy, A Recipe for Training Neural Networks: \\
\url{https://karpathy.github.io/2019/04/25/recipe/}\\
Google, Rules of Machine Learning: \\
\url{https://developers.google.com/machine-learning/guides/rules-of-ml}}

\newthought{Loading a \texttt{.pkl} from an untrusted source can execute
arbitrary code.}\index{pickle}
\texttt{pickle} and \texttt{joblib} deserialise objects by reconstructing
them, including any code embedded in the file.
This is convenient when you produced the file yourself; it is dangerous
when the file came from somewhere you do not control.
Treat \texttt{.pkl} files the way you treat executables: only run what you
trust.

\marginnote{\newthought{Safer formats exist.}\index{ONNX}
ONNX stores only weights and graph structure, no executable code.
SafeTensors does the same for PyTorch tensors.
For this course \texttt{joblib} is sufficient because we control the
training pipeline end to end; for any model received from outside,
prefer the safer formats.\\
ONNX: \url{https://onnx.ai/}\\
SafeTensors: \url{https://huggingface.co/docs/safetensors}}



% ==============================================================================================
\newpage
\section{Honouring the Model Contract}\index{train/serve skew}\index{model contract}

\newthought{Every model was trained on inputs in a specific format.}
The format is part of the contract: a 28-by-28 greyscale image, pixel
values in $[0, 1]$, flattened to 784 features.
Violate the format and the model predicts silently wrong: no error, no
warning, just a different number than the right one.
This bug class is called train/serve skew\index{train/serve skew}
and it is the most common production ML defect.

\marginnote{\newthought{Why silent?}
The model takes a vector of floats and returns a class index.
At no point does it know whether the floats came from the training
distribution or from some unrelated source.
A photograph, a random scribble, or a correctly drawn digit with the wrong
dynamic range all produce the same kind of output: a class label and a
confidence score.
The label might be wrong, but the system has no way to tell.}

\newthought{The MNIST model expects:}
\begin{itemize}
    \item a $28 \times 28$ greyscale image,
    \item pixel values normalised to $[0, 1]$,
    \item flattened to a 784-element vector.
\end{itemize}

\newthought{The canvas in the browser produces uint8 values}
in the range $[0, 255]$.
Converting to binary (0 or 1) reduces payload size, removes anti-aliasing
noise, and matches the training-time preprocessing exactly:

\begin{verbatim}
import numpy as np
raw = np.array(...)                  # (28, 28) uint8
binary = (raw > 128).astype(float)   # 0s and 1s
flat = binary.reshape(1, -1)         # (1, 784)
\end{verbatim}

\marginnote{\newthought{Binarization}\index{binarization} is aggressive but
effective for MNIST.
Anti-aliased grey edges become sharp black and white.
The model was trained on binarized inputs, so serving raw greyscale would
introduce train/serve skew immediately, with no exception to mark the
moment it happened.}

\newthought{The sklearn Pipeline} bundles preprocessing and model into one
serialised object.
Loading the pipeline is loading the preprocessing.
There is no separate normalisation step that a future maintainer might
forget; the same object that predicts also normalises, in the order the
training script established.

\marginnote{scikit-learn\index{scikit-learn}: \url{https://scikit-learn.org/}\\
joblib\index{joblib}: \url{https://joblib.readthedocs.io/}\\
NumPy\index{NumPy}: \url{https://numpy.org/}}



% ==============================================================================================
\newpage
\section{Designing the Inference Interface}\index{inference interface}\index{batch inference}

\newthought{Before writing any code,} design the contract.
Start with the class roster, then the function signatures, then the
loading strategy.
Each is a decision that propagates: every later block reads from this
interface, and every change to it ripples through the API, the database
schema, and the frontend.

\newthought{A named constant} makes the valid output set explicit and
independent of the model file:

\begin{verbatim}
CLASSES = ["1", "2", "3", "4", "5", "6", "7", "8", "9"]
# must match model.classes_, verified at startup
# note: "0" is intentionally absent in v1
\end{verbatim}

This list belongs in the code, not only in the model.
The frontend, the database schema, and the API documentation all reference it.
If the model is swapped for one trained on different classes, the assertion
described below fires loudly at startup rather than returning silently
wrong labels for the rest of the service's lifetime.

\marginnote{\newthought{Batch-first design.}\index{batch inference}
What happens when 50 students submit a drawing in the same second?
If the server processes each request individually, the model is called 50
times with a $(1, 784)$ array.
If it batches them, the model is called once with a $(50, 784)$ array.
sklearn's \texttt{predict\_proba} is natively vectorised, so the cost is
nearly identical.
Design around batch; single-user is a special case.}

\newthought{The function signatures} fall out of the contract:

\begin{verbatim}
def classify_batch(images: np.ndarray) -> list[dict]:
    """
    Args:
        images: (N, 28, 28) uint8, values 0-255
    Returns:
        List of N dicts:
        {"prediction": str, "confidence": float,
         "scores": dict[str, float]}
    """
    if images.ndim != 3 or images.shape[1:] != (28, 28):
        raise ValueError(
            f"Expected (N,28,28), got {images.shape}")
    arr = (images > 128).astype(float).reshape(
        len(images), -1)
    probs = _pipeline.predict_proba(arr)
    return [
        {"prediction": CLASSES[p.argmax()],
         "confidence": float(p.max()),
         "scores": dict(zip(CLASSES, p.tolist()))}
        for p in probs
    ]

def classify(image: np.ndarray) -> dict:
    """Single-image convenience wrapper."""
    return classify_batch(image[np.newaxis])[0]
\end{verbatim}

\marginnote{\newthought{The wrapper is for tests only.}
The API in Block~5 always calls \texttt{classify\_batch}, even for a single
drawing: it passes a $(1, 28, 28)$ array.
When a request queue is added later to amortise inference cost, no signature
changes are needed; the queue assembles a batch and the function consumes it.}

\newthought{Load \texttt{\_pipeline} once at module level:}

\begin{verbatim}
import joblib, os
_pipeline = joblib.load(os.getenv("MODEL_PATH"))
assert list(_pipeline.classes_) == CLASSES, \
    f"Model/CLASSES mismatch: {_pipeline.classes_}"
\end{verbatim}

Deserialising on every request kills throughput; the model stays in memory
for the lifetime of the process.
The assertion is the key teaching moment: it fires at startup if the model's
classes do not match the code constant.
A swapped model with the wrong classes is caught the moment the service
starts, not the next time a user gets a wrong digit.

\marginnote{\newthought{\texttt{MODEL\_PATH} from \texttt{.env}.}\index{MODEL\_PATH}
The path is configuration, not code.
Swapping the model in production becomes an env-var change plus a service
restart, not a code change and a deploy.
This is the same pattern as the database URL in Block~6 and the API key
from Block~3.}



% ==============================================================================================
\newpage
\section{The Model Card}\index{model card}\index{MODELCARD.md}

\newthought{A model card} is the documentation artefact that links a model
to its provenance and limitations.
It answers four questions any consumer of the model should know the answer
to before they trust it: where did the data come from, what can the model
do, what does it fail at, and what is it intended for.

\marginnote{Model cards as a concept were proposed by Mitchell et al.\
in 2018 and are now standard at Google, HuggingFace, and increasingly
across the industry.\\
Model Cards (Google):\\
\url{https://modelcards.withgoogle.com/}\\
HuggingFace Model Cards: \\
\url{https://huggingface.co/docs/hub/model-cards}}

\newthought{For PixelWise} the model card lives next to the artefact
in \texttt{models/MODELCARD.md} and is committed to Git, so it travels
with the repository even though the \texttt{.pkl} itself does not:

\begin{verbatim}
# MODELCARD.md: digit_classifier_v1

## Training Data
MNIST digits 1-9, ~54k train / ~9k test, public domain.
Class 0 withheld intentionally.

## Capabilities
Predict handwritten digits 1-9.
Expected accuracy: ~92% (LogisticRegression baseline).

## Known Failures
6/9 confusion, 3/8 confusion, messy or rotated digits.

## Intended Use
28x28 canvas drawings.
Out of scope: photos, non-digit characters.
\end{verbatim}

\marginnote{\newthought{The card is a contract,} like any other.
When v2 is deployed via the Block~8 feature flag, the card is updated
alongside it; the version number, the training data section, and the
capabilities all change in lockstep.
Traceability across versions is the entire reason the card is a separate
file.}

\newthought{What the card is not} is a substitute for the runtime checks
already built into the loader.
Documentation tells humans what to expect; the assertion at module load
tells the program what to expect.
Both are needed.
Documentation drifts; assertions cannot.



% ==============================================================================================
% BLOCK 3: STUDENT ACTIVATION
% ==============================================================================================

\newpage
\section{Exercise: Inspecting the Model Artefact}\index{exercises}

\marginnote{\newthought{Exercises} mirror the four theory sections in order.
Read a concept, then practise it; the structure of this block matches the
structure of Block~2 one to one.
If your repository has drifted from the end of Block~3, the catch-up below
brings it back to the known good state.}

\newthought{Catching up via the reference tag.}\index{Git!tag}
The course maintains a public reference repository at
\url{https://github.com/schutera/pixelwise} with a tag at the end of every block.
\texttt{v0.2} marks the end of Block~3; \texttt{v0.3} will mark the end of
this block.
If you joined late or your repository has drifted, reset to the previous
tag before starting:

\begin{verbatim}
git fetch origin
git reset --hard v0.2
\end{verbatim}

This rewrites your working tree to match the end of Block~3 exactly.
The state on disk is now identical to the published reference, including
\texttt{setup-server.sh}, \texttt{requirements.txt}, and \texttt{.env.example}.

\newthought{Block~2 introduced the model artefact} as something you load,
not something you commit.
Copy \texttt{digit\_classifier\_v1.pkl} (provided) into \texttt{models/}.
Confirm that \texttt{.gitignore} excludes \texttt{*.pkl} so the artefact
stays out of version control:

\begin{verbatim}
ls models/
git status
\end{verbatim}

\marginnote{\newthought{If the file appears in \texttt{git status},}
your \texttt{.gitignore} from Block~3 needs an update.
Add \texttt{models/*.pkl} on its own line and rerun \texttt{git status}.
A model file accidentally committed once is in the history forever, even
if the next commit deletes it.}

\newthought{Inspect the artefact} in a Python shell:

\begin{verbatim}
source .venv/bin/activate
python3
>>> import joblib
>>> model = joblib.load("models/digit_classifier_v1.pkl")
>>> print(type(model))
>>> print(model.classes_)
\end{verbatim}

What type is it?
What classes does it predict?
Does class 0 appear?
The answer to the third question determines what \texttt{CLASSES} should
contain in the next exercise.



% ==============================================================================================
\newpage
\section{Exercise: Preprocessing and Binarization}\index{exercises}

\newthought{Block~2 introduced binarization} as the preprocessing step that
matches the training contract.
Load a sample MNIST digit and compare the raw and binarized representations
side by side:

\begin{verbatim}
from sklearn.datasets import fetch_openml
import numpy as np

X, y = fetch_openml("mnist_784", version=1,
    return_X_y=True, as_frame=False)
sample = X[0].reshape(28, 28).astype(np.uint8)
binary = (sample > 128).astype(float)
print(sample[:5, :5])    # raw values
print(binary[:5, :5])    # 0s and 1s
\end{verbatim}

\marginnote{\newthought{Notice the structure survives.}\index{binarization}
The raw image has anti-aliased grey edges; the binarized image is sharp
black and white.
The digit is still recognisable, the noise is gone, and the input now
matches what the model saw during training.
Serving raw greyscale would mean predicting against a distribution the
model has never seen.}

\newthought{Plot if you have a notebook open.}
A side-by-side image of the raw and binarized digit makes the effect
visible at a glance.
\texttt{matplotlib} is not in \texttt{requirements.txt} yet; install it
locally with \texttt{pip install matplotlib} for this exercise only,
then leave it out of the pinned requirements until a later block needs it.

\marginnote{\newthought{Why threshold at 128?}\index{threshold}
128 is the midpoint of the uint8 range, the simplest split between
``ink'' and ``paper''.
Otsu's method picks a smarter threshold per image, and for noisier
inputs that is sometimes worth doing.
For MNIST the simple split is enough and matches the training pipeline
exactly.}



% ==============================================================================================
\newpage
\section{Exercise: The Classifier Module}\index{exercises}

\newthought{Block~2 introduced the inference interface.}
Implement \texttt{app/classifier.py} with the class constant, module-level
loading, the startup assertion, and the two functions:

\begin{verbatim}
import os
import joblib
import numpy as np
from dotenv import load_dotenv

load_dotenv()

CLASSES = ["1","2","3","4","5","6","7","8","9"]

_pipeline = joblib.load(os.getenv("MODEL_PATH"))
assert list(_pipeline.classes_) == CLASSES, \
    f"Model/CLASSES mismatch: {_pipeline.classes_}"

def classify_batch(images: np.ndarray) -> list[dict]:
    if images.ndim != 3 or images.shape[1:] != (28, 28):
        raise ValueError(
            f"Expected (N,28,28), got {images.shape}")
    arr = (images > 128).astype(float).reshape(
        len(images), -1)
    probs = _pipeline.predict_proba(arr)
    return [
        {"prediction": CLASSES[p.argmax()],
         "confidence": float(p.max()),
         "scores": dict(zip(CLASSES, p.tolist()))}
        for p in probs
    ]

def classify(image: np.ndarray) -> dict:
    return classify_batch(image[np.newaxis])[0]
\end{verbatim}

\marginnote{\newthought{The assertion is the headline.}\index{assertion}
Comment it out, save, restart Python, and import the module.
Now mutate \texttt{CLASSES} to add a fake \texttt{"10"} entry, restore
the assertion, restart, and observe the assertion firing.
The startup error message is exactly what you want to see when a model
swap goes wrong.}

\newthought{Update \texttt{.env} and \texttt{.env.example}} so the loader
knows where to find the artefact:

\begin{verbatim}
MODEL_PATH=models/digit_classifier_v1.pkl
\end{verbatim}

\newthought{Write \texttt{predict.py} as a smoke test.}
Load five MNIST test samples, call \texttt{classify\_batch}, print the
predictions next to the ground truth:

\begin{verbatim}
from app.classifier import classify_batch
from sklearn.datasets import fetch_openml
import numpy as np

X, y = fetch_openml("mnist_784", version=1,
    return_X_y=True, as_frame=False)
images = X[:5].reshape(-1, 28, 28).astype(np.uint8)
truth  = y[:5]
results = classify_batch(images)
for r, t in zip(results, truth):
    print(f"Pred: {r['prediction']} "
          f"(conf {r['confidence']:.2f})  True: {t}")
\end{verbatim}

\marginnote{\newthought{Read \texttt{predict.py} as the API preview.}\index{predict.py}
This is what the API in Block~5 will do: take pixel data, call
\texttt{classify\_batch}, return the result.
Block~5 swaps stdin for HTTP and stdout for JSON; the inference logic
is unchanged.}



% ==============================================================================================
\newpage
\section{Exercise: The Model Card}\index{exercises}

\newthought{Block~2 introduced the model card} as the documentation
artefact that travels with the project even though the \texttt{.pkl} does
not.
Write \texttt{models/MODELCARD.md} following the structure from the theory
section: dataset, input format, performance per class, known failures,
version, and date.

\marginnote{\newthought{Use the values you measured.}\index{model card}
Replace ``${\sim}92\%$'' with whatever \texttt{predict.py} reports as you
expand it across the test set.
The numbers you cite should match the numbers your code produces; a model
card with optimistic round figures is the start of the documentation drift
the previous section taught us to avoid.}

\newthought{Commit the new files,} but not the \texttt{.pkl}:

\begin{verbatim}
git add app/classifier.py predict.py \
    models/MODELCARD.md .env.example
git commit -m "Add classifier module and model card"
git push
\end{verbatim}

The repository is now at the state \texttt{v0.3} marks: a working classifier,
a smoke test, a model card, and an updated configuration contract.
The \texttt{.pkl} stays on each developer's machine, the artefact path is
in \texttt{.env}, and the rest of the team can reproduce the environment
from the code in Git.

\marginnote{\newthought{Verify the smoke test on \texttt{prod}.}\index{provisioning}
SSH into the server, pull, copy the \texttt{.pkl} into
\texttt{/opt/pixelwise/models/}, set \texttt{MODEL\_PATH} in the production
\texttt{.env}, and run \texttt{python predict.py}.
Identical output on both machines is the deliverable of this block.}



% ==============================================================================================
\newpage
\section{Optional: Add the Missing 0}\index{model versioning}

\newthought{The digit class ``0'' was intentionally excluded from v1.}
To produce v2: add \texttt{"0"} to \texttt{CLASSES}, include MNIST's class-0
samples in the training split, re-run \texttt{train.py}, save as
\texttt{digit\_classifier\_v2.pkl}.
The assertion will break the moment v2 is loaded against v1's
\texttt{CLASSES}; fix it by updating the constant.
Update \texttt{MODELCARD.md} to document that v2 covers all ten digits and
note any accuracy change.

\marginnote{This challenge connects directly to Block~8: v2 is deployed
via the feature flag, and Block~9's analytics compare v1 and v2 performance.
No data collection is needed; MNIST has the 0s already.}

\newthought{Could we have spotted the missing class at runtime?}
\begin{itemize}
    \item After a few hundred predictions, a missing class is invisible:
    the model never predicts it, users never see an error, the database
    just never accumulates any 0s.
    \item When the model sees an input it was not trained on, it is often
    less certain.
    A systematic drop in average confidence could indicate
    out-of-distribution inputs: a weak signal, but a runtime signal that
    requires no labelled data.
    \item Returning \texttt{null} when \texttt{confidence < threshold} turns
    low confidence into a monitoring signal: the volume of rejected
    predictions is itself a metric.
\end{itemize}

\marginnote{\newthought{Systematic detection of silent failures}\index{MLOps}
is the domain of MLOps: uncertainty estimation, distribution shift,
missing classes, silent degradation.
This surface is intentionally shallow here.
A dedicated MLOps course covers it in depth.}

\newthought{A note on what comes next.}
\texttt{app/classifier.py} works, but it is trapped in a Python script.
A user with a browser cannot call it; a teammate on a different machine
cannot reach it.
The next block wraps \texttt{classify\_batch} in an HTTP endpoint, defines
the request and response contract with Pydantic, and runs the result as a
managed \texttt{systemd} service so the model becomes reachable without a
human in the loop.



% ==============================================================================================
\newpage
\section{Self-Reflection and Recap}\index{reflection}\index{recap}

\newthought{Self-Reflection} questions to guide your thoughts during and after the exercises:
\begin{itemize}
    \item What is train/serve skew, and why is it the most common production ML defect?
    \item Why do we load the model once at module level instead of on every request?
    \item What does the startup assertion protect against, and what would happen without it?
    \item Why is the \texttt{.pkl} in \texttt{.gitignore} but \texttt{MODELCARD.md} is committed?
    \item How would you detect at runtime that the model is missing a class it should predict?
    \item Why is batch-first design the right default, even when most requests carry a single image?
\end{itemize}

\newthought{Recap} of key concepts:
\begin{itemize}
    \item A \texttt{.pkl} file is a build artefact, not source code; never commit it to Git.
    \item Train/serve skew is silent by default; the startup assertion is the loudest defence.
    \item Batch-first design scales from one user to $N$ concurrent users with no code changes.
    \item Module-level loading keeps the model in memory for the lifetime of the process.
    \item The model card documents capabilities, limitations, and provenance, and updates with the model.
    \item \texttt{predict.py} proves end-to-end correctness before any API exists.
\end{itemize}

\marginnote{\newthought{Security thread.}\index{security}
Never load a \texttt{.pkl} from an untrusted source; \texttt{joblib} and
\texttt{pickle} deserialisation execute arbitrary code.
\texttt{MODEL\_PATH} comes from \texttt{.env}, never hard-coded, so swapping
the model in production requires only an env-var change, not a code change.
Training data and model artefacts are not committed to Git.}

\marginnote{\newthought{Teaser.} The classifier is callable from Python,
but who, outside your terminal, can reach it?}

\newthought{Milestone.} \texttt{app/classifier.py} exposes a batch-first
inference interface that scales from one student to $N$ concurrent users
with no code changes.
\texttt{predict.py} proves it works end to end on real MNIST samples.
\texttt{MODELCARD.md} documents what the model is and is not.
The next chapter wraps the inference function in a FastAPI service so the
model becomes reachable over HTTP.
