% !TEX root = ../notes_missingsemester.tex
% Chapter 6: Data Lives Here — Databases & SQL
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% POST IMPROVEMENTS 2026:
% 1. Add a small TikZ schema diagram of the predictions table.
% 2. Provide a pre-baked v0.4 reference repo so catch-up takes a single clone.

\index{SQLite}\index{SQL}\index{SQLAlchemy}\index{database}\index{Alembic}\index{ORM}\index{PostgreSQL}
\chapter{Databases \& SQL}
\printchapterversion{06_databases_sql}
\newthought{In Comparison, Dumbledore's Pensieve is a joke.}

% ==============================================================================================
% BLOCK 1: INTRODUCTION
% ==============================================================================================

\section{The Why}\index{persistence}\index{relational database}

We have a service that classifies, but it does not remember anything.
The moment the response leaves the server, the prediction is gone.
This makes it hard to track the model's performance over time, compare versions, or
build any type of features and functionality that revolve around
past predictions.

\newthought{We need to wire the API onto a database.}\index{PixelWise}
This block is the first that adds persistent state.
Every later block depends on the database schema you design today.

\begin{figure}
\centering
\begin{tikzpicture}[
    every node/.style={font=\footnotesize, align=center},
    box/.style={rectangle, draw=black, thick, rounded corners=2pt, minimum width=2.2cm, minimum height=0.9cm, inner sep=4pt},
    arr/.style={-latex, thick, black!70}
]
\node[box] (api) at (0, 0)    {FastAPI\\\texttt{/classify}};
\node[box] (cb)  at (3.4, 0)  {\texttt{classify\_}\\\texttt{batch}};
\node[box] (pg)  at (7.4, 0)  {SQLite\\\texttt{predictions.db}};
\draw[arr] (api.east) -- (cb.west);
\draw[arr] (cb.east)  -- node[above]{\scriptsize SQLAlchemy} (pg.west);
\end{tikzpicture}
\caption{The persistence flow. \texttt{POST /classify} computes a prediction with \texttt{classify\_batch}, the handler maps the result through the \texttt{Prediction} SQLAlchemy model, and a row lands in the \texttt{predictions} table inside the \texttt{predictions.db} SQLite file before the response goes out.}
\label{fig:persistence_flow}
\end{figure}





\newthought{Why persist at all?}
A response that vanishes the moment it is sent is a response that cannot be
counted, compared, or queried.
A database makes the prediction permanent: Every drawing becomes a row, the
class label becomes a column, the confidence becomes another, and the model
version becomes a third.

\newthought{Beyond this course,}
the same table is where production teams build retraining datasets, let
users correct wrong predictions for labelled data, and detect distribution
drift.
We do not do any of those here, but it is trivial to see how every one of them starts with having
the database.



\newpage
\subsection{Hands On Experience}

\newthought{Consider this scenario:} A teammate asks how confident the
model has been on average across the last hundred predictions.
You realise nobody can answer that question.
The API returned the confidence to every caller, but nothing recorded it.
The only way back to those numbers is to ask the users to draw their digits
again, which is not an option. In an increasingly data-driven world, data defines your product, 
your moat, and your competitive advantage. Not having the data is like throwing away your most valuable asset, and 
until someone comes up with a solution for time-travel, it's a non-recoverable loss.

\newthought{The schema is a contract,} the same way the API contract is a
contract.
Get it right and every later question, distribution per digit, average
confidence per model version, error rate over time, becomes a query.
Get it wrong and you find out later, with a hundred thousand rows already
written, that the column you need does not exist. Schemas evolve through
migrations, which we cover at the end of the block as an optional path.

\marginnote{\newthought{Designing schemas} is its own discipline.
We design the \texttt{predictions} table once and keep it that way for the
mainline of the course. The optional Alembic exercise at the end lets you
feel the cycle of design, deploy, regret, migrate on a real database.}

\newthought{This sixth block} introduces persistence for PixelWise.
By the end you will have
\begin{itemize}
    \item written your first SQL queries against a SQLite database,
    \item modelled the \texttt{predictions} table as a SQLAlchemy class,
    \item wired \texttt{POST /classify} so every prediction becomes a row,
    \item replaced the \texttt{GET /results} stub with a real query,
    \item configured database credentials through \texttt{.env},
    \item and, optionally, run your first schema migration with Alembic.
\end{itemize}



% ==============================================================================================
\newpage
\section{Optional Exercise: Catch Up to \texttt{v0.4}}\index{exercises}\index{Git!tag}\index{recovery}

\marginnote{\newthought{Tag map.}\index{Git!tag}
\texttt{v0.4} marks the end of Block~5, \texttt{v0.5} will mark the end
of this block.
Browse the full set at
\url{https://github.com/schutera/pixelwise/tags}.}

\newthought{Catching up via tag.}\index{Git!tag}\index{recovery}
This catch-up extends the one in Block~5. If you have not done that one,
do it first: It lands you at \texttt{v0.3}, the end-of-Block~4 state.
The exercise here picks up from there and rewinds to \texttt{v0.4}, the
end-of-Block~5 state, with the FastAPI service exposing
\texttt{POST /classify} and \texttt{GET /health}, API key authentication,
basic rate limiting, and a \texttt{systemd}-supervised \texttt{uvicorn}
running on \texttt{prod}.

\newthought{Both VMs this time.} Block~5 ended with the service actually
running on \texttt{prod}, so unlike the previous catch-up, this one
brings both machines up to \texttt{v0.4}. The Git-tracked changes since
\texttt{v0.3} include \texttt{app/main.py},
\texttt{deploy/pixelwise.service}, an extended \texttt{setup-server.sh}
that installs the systemd unit on \texttt{prod}, an updated
\texttt{requirements.txt} adding \texttt{fastapi}, \texttt{uvicorn},
\texttt{slowapi}, and \texttt{pydantic}, and a refreshed
\texttt{.env.example} with \texttt{SECRET\_API\_KEY}. The local
\texttt{.env} on \texttt{prod} needs the new key written by hand;
\texttt{setup-server.sh} then pulls the model, copies the unit,
enables, and restarts the service in one shot.


\marginnote{\newthought{Sequence on \texttt{dev}.}\\[0.4em]
\texttt{cd \textasciitilde/pixelwise}\\
\texttt{git fetch origin -{}-tags}\\
\texttt{git reset -{}-hard v0.4}\\
\texttt{cp .env.example .env}\\
\texttt{nano .env}\\
\texttt{source .venv/bin/activate}\\
\texttt{pip install -r requirements.txt}\\
\texttt{bash setup-server.sh}\\[0.3em]
Edit \texttt{.env} to set \texttt{SECRET\_API\_KEY} to any string of
your choosing; \texttt{openssl rand -hex 32} prints a clean random
value if you want one.
If you already have a \texttt{.env} from earlier work, skip the
\texttt{cp} and just edit the file in place.
If \texttt{.venv/} is missing, run
\texttt{python3 -m venv .venv} before the \texttt{source} step.}

\newthought{Run the catch-up on \texttt{dev}.}
Rewind the repository to \texttt{v0.4}, refresh \texttt{.env} from the
updated example and edit it to set \texttt{SECRET\_API\_KEY}, reinstall
pinned dependencies if your virtual environment was wiped, then run
\texttt{setup-server.sh} to pull the model artefact. The exact commands
are in the margin.




\marginnote{\newthought{Sequence on \texttt{prod}.}\\[0.4em]
\texttt{ssh produser@192.168.56.11}\\
\texttt{cd /opt/pixelwise}\\
\texttt{git fetch origin -{}-tags}\\
\texttt{git reset -{}-hard v0.4}\\
\texttt{cp .env.example .env}\\
\texttt{nano .env}\\
\texttt{source .venv/bin/activate}\\
\texttt{pip install -r requirements.txt}\\
\texttt{bash setup-server.sh}\\[0.3em]
Use the same \texttt{SECRET\_API\_KEY} value as on \texttt{dev}.
If you already have a \texttt{.env} on \texttt{prod}, skip the
\texttt{cp} and edit the file in place.
Then from \texttt{dev}:
\texttt{curl http://192.168.56.11:8000/health}
should return \texttt{\{"status":"ok"\dots\}}.}

\newthought{Run the catch-up on \texttt{prod}.}
Same sequence on \texttt{prod}, where \texttt{setup-server.sh} also
copies the systemd unit into place and restarts the service. After the
script finishes, verify the API is reachable from \texttt{dev}.




\newthought{You are ready player one}.



% ==============================================================================================
% BLOCK 2: CONCEPTS AND PRACTICE
% ==============================================================================================

\newpage
\section{Relational Databases and SQL}\index{relational database}\index{SQL}

\newthought{A relational database stores data in tables.}
Each table has a fixed schema: a list of named columns, each with a type.
Each row is a single record.
The schema is the contract that every record must follow; the database refuses
inserts or queries that violate it.

\marginnote{\newthought{Vocabulary to keep straight.}\index{primary key}
A table is a named set of rows; a row is one record;
a column is a named field with a type;
a primary key is the column or columns that uniquely
identify each row.
Most tables use an auto-incrementing integer or a UUID as their key.

\vspace{0.5em}
{\footnotesize
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}llll}
\textbf{id} & \textbf{prediction} & \textbf{confidence} & \textbf{model\_version} \\
\hline
1 & 7 & 0.94 & v1 \\
2 & 3 & 0.81 & v1 \\
3 & 5 & 0.99 & v1 \\
\end{tabular*}}
\par\vspace{0.2em}
}

\newthought{Structured Query Language (SQL)\index{SQL}} is the language you use to talk to a relational database.
The database is a single file, there is no server to install, no users,
no network configuration. For a single-developer prototype or a desktop
application, this is exactly what you want.
Four operations cover almost everything:


\marginnote{\newthought{What you will meet later.}\index{SQL!JOIN}\index{SQL!GROUP BY}
\begin{itemize}
    \item \texttt{JOIN} combines rows from multiple tables.
    \item \texttt{UPDATE} adjusts existing values without changing the schema.
    \item \texttt{GROUP BY}, \texttt{COUNT}, and \texttt{AVG} aggregate.
    \item \texttt{ORDER BY} and \texttt{LIMIT} shape the output.
    \item \texttt{CREATE INDEX} speeds up queries that scan large tables.
\end{itemize}
The full treatment belongs in a dedicated database course; the four
operations above are enough to get PixelWise running.}


\begin{verbatim}
-- Insert a row
INSERT INTO predictions (prediction, confidence,
                         model_version)
VALUES ('7', 0.94, 'v1');

-- Read rows
SELECT prediction, confidence FROM predictions
    WHERE model_version = 'v1'
    ORDER BY created_at DESC
    LIMIT 10;

-- Delete a row
DELETE FROM predictions WHERE id = 42;
\end{verbatim}


\vfill
\subsection{PostgreSQL for Production}\index{PostgreSQL}\index{SQLite}

\marginnote{\newthought{Where SQLite stops short?}
SQLite has no user authentication, no password protection, and limited
concurrency - due to file locking.}

\newthought{PostgreSQL\index{PostgreSQL} is a production database.}
It runs as a separate process, accepts connections over the network, has
real users with real permissions, and supports concurrent reads and
writes without serialising the whole file. \texttt{psql}\index{psql} is
the command line client.

\marginnote{PostgreSQL: \url{https://www.postgresql.org/}\\
SQLite: \url{https://www.sqlite.org/}}

\newthought{Why a dedicated user? Least Privilege}\index{least privilege}
The \texttt{postgres} superuser can drop any database, create any user,
and read any table. A dedicated \texttt{pixelwise} user can only do what
its privileges allow, which by default is read and write its own tables.
The principle is the same as the non-root \texttt{produser} on your server.
You provision a \texttt{pixelwise} role and a \texttt{pixelwise} database
owned by that role inside PostgreSQL, then verify the credentials with
\texttt{psql} before any Python is involved.






% ==============================================================================================
\newpage
\section{Exercise: SQLite on the Server}\index{exercises}\index{SQLite}

\newthought{While you could set things up manually on \texttt{prod},} it
is a best practice to automate every step of the server configuration
through \texttt{setup-server.sh}; you edit it on \texttt{dev}, commit,
push, and \texttt{bash setup-server.sh} carries the work on both VMs.
The one manual step is writing the secret \texttt{DB\_PASSWORD} into the
local \texttt{.env} on each VM, since secrets never travel through git.
Provisioning on \texttt{dev} too means you can write \texttt{models.py},
run \texttt{init\_db.py}, and inspect rows with \texttt{psql} without
ever ssh'ing to \texttt{prod}.

\newthought{Extend \texttt{setup-server.sh} on \texttt{dev}.}\index{setup-server.sh}
Add \texttt{postgresql} to the \texttt{apt install} line, then append a
stanza that provisions the \texttt{pixelwise} role and database wherever
\texttt{.env} is present:

\begin{verbatim}
cd ~/pixelwise
nano setup-server.sh
\end{verbatim}

\marginnote{\newthought{If \texttt{psql} is missing}\index{postgresql-client-common}
after the script runs, the provisioning stanza fails. The
\texttt{postgresql} meta-package normally pulls \texttt{postgresql-client-common}
in transitively; on older Ubuntu releases, minimal cloud images, or
when the apt cache is stale, that chain breaks and the package has
to be installed by hand:\\[0.4em]
\texttt{sudo apt install postgresql-client-common}\\[0.3em]
\texttt{bash setup-server.sh}}

\marginnote{\newthought{The connection string}\index{DATABASE\_URL} is a
single URL: \\
\texttt{sqlite:////absolute/path/to/file.db}.
Note the four slashes after the colon: three for the URL scheme, one
for the absolute path. A relative path uses three slashes
(\texttt{sqlite:///./pixelwise.db}). The application reads this from
\texttt{.env}, so the same code works on \texttt{dev} and \texttt{prod}
with different paths.}

\marginnote{\newthought{Idempotent by design.}\index{idempotence}
Each \texttt{psql -tAc} probe asks PostgreSQL whether the role or
database already exists; the \texttt{||} short-circuits the create
when the answer is yes. Run the script ten times, the role and database
are still there exactly once. \texttt{SCRIPT\_DIR} resolves to whichever
checkout invoked the script, so the same stanza runs on \texttt{dev} and
\texttt{prod}; the systemd block from Block~5 stays gated behind
\texttt{id produser} since the service only ever runs on \texttt{prod}.}

\newthought{Resolve the script directory at the top.}
Add this line to \texttt{setup-server.sh} right after the shebang, so
every later stanza can locate the repo root regardless of the caller's
working directory:

\begin{verbatim}
SCRIPT_DIR="$(cd "$(dirname "${BASH_SOURCE[0]}")" \
  && pwd)"
\end{verbatim}

\newthought{Append the provisioning stanza} after the systemd block:

\begin{verbatim}
# Provision the pixelwise role and database on every VM
if command -v psql >/dev/null 2>&1 && \
   [ -f "$SCRIPT_DIR/.env" ]; then
    set -a; source "$SCRIPT_DIR/.env"; set +a
    sudo -u postgres psql -tAc \
        "SELECT 1 FROM pg_roles WHERE rolname='pixelwise'" \
        | grep -q 1 || \
    sudo -u postgres psql -c \
        "CREATE USER pixelwise \
         WITH PASSWORD '$DB_PASSWORD';"
    sudo -u postgres psql -tAc \
        "SELECT 1 FROM pg_database WHERE datname='pixelwise'" \
        | grep -q 1 || \
    sudo -u postgres createdb -O pixelwise pixelwise
fi
\end{verbatim}

\newthought{Document the new secret} in \texttt{.env.example} on
\texttt{dev}, then commit and push:

\begin{verbatim}
echo "DB_PASSWORD=" >> .env.example
git add setup-server.sh .env.example
git commit -m "Provision PostgreSQL"
git push
\end{verbatim}

\newthought{Apply on \texttt{dev} first.}
Write the password into your local \texttt{.env} and run the script:

\begin{verbatim}
echo "DB_PASSWORD=database" >> .env
bash setup-server.sh
\end{verbatim}

\marginnote{\newthought{Did it take on \texttt{dev}?}\\[0.4em]
\texttt{psql -U pixelwise -d pixelwise -h localhost}\\[0.3em]
Enter the password from \texttt{.env}. A \texttt{pixelwise=>} prompt
means the role exists, the database is yours, and the password matches.
If the connection is refused, re-read the script output before moving
on to \texttt{prod}.

\vspace{0.4em}
\newthought{Wipe and start fresh.}
The provisioning stanza skips creation when the role or database
already exists, so editing \texttt{DB\_PASSWORD} after a first run does
not propagate. If \texttt{psql} rejects the password or reports a
missing database, drop both and re-run:\\[0.4em]
\texttt{sudo -u postgres dropdb pixelwise}\\
\texttt{sudo -u postgres dropuser pixelwise}\\
\texttt{bash setup-server.sh}}

\newthought{Repeat on \texttt{prod}.}
Same shape, plus the \texttt{git pull} to bring in the committed script:

\begin{verbatim}
ssh produser@192.168.56.11
cd /opt/pixelwise
git pull
echo "DB_PASSWORD=database" >> .env
bash setup-server.sh
\end{verbatim}

The first run on each VM installs \texttt{postgresql} and provisions the
\texttt{pixelwise} role and database; subsequent runs change nothing
because the provisioning stanza is idempotent.

\newthought{Verify the connection} on \texttt{prod} with the application
user:

\begin{verbatim}
psql -U pixelwise -d pixelwise -h localhost
\end{verbatim}



\marginnote{\newthought{What each flag does.}\index{psql}
\begin{itemize}
\item \texttt{-U pixelwise} selects the role to log in as.
\item \texttt{-d pixelwise} selects the database to connect to.
\item \texttt{-h localhost} forces a TCP connection to the local
machine, which makes \texttt{psql} prompt for the password from
\texttt{.env} instead of trusting the Unix socket's peer authentication
on the \texttt{postgres} system user.
\end{itemize}}

\marginnote{\newthought{Role and database share a name.}
The setup script creates both:
\begin{itemize}
\item A role \texttt{pixelwise} with a password.
\item A database \texttt{pixelwise} owned by that role.
\end{itemize}
PostgreSQL keeps roles and databases as independent objects, so the
matching names are a readability convention, not a requirement; the
application could just as well live in a database called
\texttt{predictions} owned by a role called \texttt{api}.}



\texttt{pixelwise=>} prompt means the role and database are wired up.
If \texttt{psql} fails to connect, fix the credentials before going any
further.
A working \texttt{psql} is the foundation of everything the rest of the
block builds on. If it worked you know








% ==============================================================================================
\newpage
\section{SQLAlchemy: Tables as Python Classes}\index{SQLAlchemy}\index{ORM}

\newthought{SQLAlchemy is a Python library} that maps database tables to
Python classes. You define a \texttt{Prediction} class with fields like
\texttt{prediction}, \texttt{model\_version}, and \texttt{created\_at};
SQLAlchemy generates the SQL to create the table, insert rows, and query
data. This is called an ORM\index{ORM}, an object-relational mapper, and
is standard in Python web development.

\marginnote{SQLAlchemy: \\
\url{https://www.sqlalchemy.org/}}

\marginnote{SQLAlchemy Tutorial: \\
\url{https://docs.sqlalchemy.org/en/20/tutorial/}}


\newthought{Python is ergonomics.}\index{SQL injection}
Two reasons, why choosing to use an ORM: Writing Python is faster than writing
string SQL, and the editor knows the field names. The second is safety:
SQLAlchemy uses parameterised queries by default, so user input never ends
up concatenated into a SQL statement. SQL injection becomes a footgun the
framework refuses to hand you.

\newthought{An engine and a session} are the runtime objects that connect
the model to the database. \texttt{create\_engine} takes the connection
string \texttt{models.py} assembles from \texttt{DB\_PASSWORD} and opens a
pool of connections;
\texttt{sessionmaker} returns a \texttt{SessionLocal} factory the handlers
call once per request; \texttt{Base.metadata.create\_all(engine)} creates
the table on first run.

\marginnote{\newthought{The connection string}\index{DATABASE\_URL} is a single
URL: \texttt{postgresql://user:pass@host/dbname}.
It encodes everything the driver needs to reach the database.
\texttt{models.py} assembles it from the \texttt{DB\_PASSWORD} secret in
\texttt{.env}, so the secret stays in one place and a leaked connection
string is still a leaked database.}

% ==============================================================================================
\newpage
\section{Exercise: The \texttt{Prediction} Model}\index{exercises}

\newthought{Install the driver and the ORM on \texttt{dev}} into the
virtual environment, then pin the new packages to
\texttt{requirements.txt} so \texttt{prod} picks them up on the next
deploy.

\marginnote{\newthought{Install.}\\[0.4em]
\texttt{cd \textasciitilde/pixelwise}\\
\texttt{source .venv/bin/activate}\\
\texttt{pip install sqlalchemy psycopg2-binary}\\
\texttt{pip freeze > requirements.txt}}

\newthought{Write \texttt{app/models.py}} on \texttt{dev} with both the
schema for one prediction and the session factory the handlers will use:
The class label, the confidence, and the model version, plus an engine
that assembles its connection string from \texttt{DB\_PASSWORD} and a
\texttt{SessionLocal} bound to it.
\texttt{DB\_PASSWORD} is already in \texttt{.env} from the provisioning
exercise, and \texttt{create\_engine} only opens a connection when a
query runs, so defining the engine here is safe:

\marginnote{\newthought{Three columns plus bookkeeping.}\index{schema}
\texttt{prediction}, \texttt{confidence}, and \texttt{model\_version}
carry the answer; \texttt{id} and \texttt{created\_at} are bookkeeping the
database fills in for you. Every later question, distribution per digit,
average confidence per model version, error rate over time, lives in this
table.}

\begin{verbatim}
from sqlalchemy import (Column, Integer, String,
                        Float, DateTime, create_engine)
from sqlalchemy.orm import declarative_base, sessionmaker
from datetime import datetime
import os
from dotenv import load_dotenv

load_dotenv()

Base = declarative_base()

class Prediction(Base):
    __tablename__ = "predictions"
    id = Column(Integer, primary_key=True)
    prediction = Column(String, nullable=False)
    confidence = Column(Float, nullable=False)
    model_version = Column(String, nullable=False)
    created_at = Column(DateTime,
                        default=datetime.utcnow)

DB_PASSWORD = os.getenv("DB_PASSWORD")
DATABASE_URL = (
    "postgresql://pixelwise:"
    f"{DB_PASSWORD}@localhost/pixelwise")
engine = create_engine(DATABASE_URL)
SessionLocal = sessionmaker(bind=engine)
\end{verbatim}



\marginnote{\newthought{One engine, one factory.}\index{SQLAlchemy!session}
\texttt{create\_engine} holds the connection pool and \texttt{sessionmaker}
turns it into \texttt{SessionLocal}, the callable each handler invokes to
open a database session. The engine is lazy: It connects only when a
session first runs a query, so importing this module never needs a live
database.}

\marginnote{\newthought{Three parsers, one file.}\index{systemd!EnvironmentFile}
\texttt{.env} is read by bash in \texttt{setup-server.sh}, by
python-dotenv in the app, and by \texttt{systemd} through its
\texttt{EnvironmentFile}. The first two expand \texttt{\$\{...\}}
references; \texttt{systemd} does not. Keep every value a plain literal
and the three parsers have nothing to disagree about.}

\newthought{No \texttt{DATABASE\_URL} line goes in \texttt{.env}.}
The engine assembles its connection string from \texttt{DB\_PASSWORD},
which you already wrote into \texttt{.env} during the provisioning
exercise, so the password stays in exactly one place. Writing a
\texttt{DATABASE\_URL} that interpolates \texttt{\$\{DB\_PASSWORD\}}
inside \texttt{.env} looks tidier but is a trap: Bash and python-dotenv
expand the reference, yet \texttt{systemd} reads its
\texttt{EnvironmentFile} without expanding anything, so the service would
try to authenticate with the literal variable text and Postgres would
reject it. Assembling the URL in \texttt{models.py} keeps every reader of
\texttt{.env} in agreement, no matter who starts the process.

\newthought{Create the table once} from a small initialisation script
saved as \texttt{init\_db.py} at the repository root on \texttt{dev}.
The script imports the engine that \texttt{app/models.py} already
configured and calls \texttt{Base.metadata.create\_all} to materialise
every model class into a table:

\begin{verbatim}
from app.models import Base, engine

Base.metadata.create_all(engine)
\end{verbatim}

\marginnote{\newthought{Initialise the schema.}\index{schema!initialisation}
by running \texttt{python init\_db.py} on \texttt{dev}.}

\newthought{Commit and deploy.}
Stage the new and changed files on \texttt{dev}, push, and on each VM
install the pinned dependencies and run \texttt{python init\_db.py} to
create the schema. Then confirm with \texttt{psql} that the
\texttt{predictions} table is listed:

\begin{verbatim}
psql -U pixelwise -d pixelwise -h localhost -c "\dt"
\end{verbatim}

\marginnote{\newthought{Commit and deploy.}\\[0.4em]
On \texttt{dev}:\\
\texttt{git add app/models.py init\_db.py}\\
\texttt{\ \ requirements.txt .env.example}\\
\texttt{git commit -m "Add Prediction model"}\\
\texttt{git push}\\
\texttt{source .venv/bin/activate}\\
\texttt{python init\_db.py}\\[0.3em]
On \texttt{prod}:\\
\texttt{cd /opt/pixelwise \&\& git pull}\\
\texttt{source .venv/bin/activate}\\
\texttt{pip install -r requirements.txt}\\
\texttt{python init\_db.py}}

\marginnote{\newthought{Optional: One-shot deploy.}
Make \texttt{bash setup-server.sh} also install pinned dependencies and
create the schema by appending two stanzas after the Postgres
provisioning block:\\[0.4em]
\texttt{if [ -d "\$SCRIPT\_DIR/.venv" ] \&\&}\\
\texttt{\ \ [ -f "\$SCRIPT\_DIR/requirements.txt" ]; then}\\
\texttt{\ \ source "\$SCRIPT\_DIR/.venv/bin/activate"}\\
\texttt{\ \ pip install -r}\\
\texttt{\ \ \ \ "\$SCRIPT\_DIR/requirements.txt"}\\
\texttt{fi}\\[0.4em]
\texttt{if [ -f "\$SCRIPT\_DIR/init\_db.py" ]; then}\\
\texttt{\ \ (cd "\$SCRIPT\_DIR" \&\& python init\_db.py)}\\
\texttt{fi}\\[0.3em]
Keep the order: The init stanza relies on the venv the first stanza
activates.}



% ==============================================================================================
\newpage
\section{Exercise: Wiring the API to the Database}\index{exercises}

\newthought{The database is ready.} Let's make the API populate it with real data.

\newthought{Edit \texttt{app/main.py} on \texttt{dev}} so
\texttt{POST /classify} writes a row before returning the response, and
\texttt{GET /results} returns real rows instead of the current scaffold.
The change lands in three pieces: Two new imports at the top of the
file, an addition to the classify handler, and a real body for results.

\marginnote{\newthought{Two new imports.}
\texttt{Prediction} is the row class and \texttt{SessionLocal} the
session factory, both defined in \texttt{app/models.py}. Add the line at
the top of \texttt{main.py} next to the existing imports.}

\begin{verbatim}
from app.models import Prediction, SessionLocal
\end{verbatim}

\marginnote{\newthought{Classify now writes a row.}\index{SQLAlchemy!session}
The handler keeps its mainline signature from Block~5. If you ran the
optional API key and rate limiting exercise, leave your
\texttt{Depends(verify\_api\_key)} dependency, the \texttt{@limiter.limit}
decorator, and the \texttt{request: Request} parameter in place; only the
body changes either way.
The body picks up one new intermediate, \texttt{result},
so the same dict can be persisted and returned, then a five-line
session block that opens a \texttt{SessionLocal}, adds the new
\texttt{Prediction}, commits, and closes. 

A session is the unit of
database work: Open it at the start of the handler, close it at the
end. FastAPI documents a more elegant variant through \texttt{Depends};
the manual version here is the simplest one that shows what is
happening. The returned \texttt{result} dict is unchanged, so the
response body stays identical to the contract we defined in Block~5.}

\begin{verbatim}
@app.post("/classify",
          response_model=ClassifyResponse)
def classify(req: ClassifyRequest):
    arr = np.array(req.pixels,
                   dtype=np.uint8)[np.newaxis]
    result = classify_batch(arr)[0]

    db = SessionLocal()
    db.add(Prediction(
        prediction=result["prediction"],
        confidence=result["confidence"],
        model_version="v1"))
    db.commit()
    db.close()

    return result
\end{verbatim}

\marginnote{\newthought{Results runs one query.}
It pulls the twenty newest rows from \texttt{predictions}, ordered by
\texttt{created\_at} descending, and shapes each row into the JSON
object the frontend will expect in the next block. The Block~5 stub
that returned an empty list is gone.}

\begin{verbatim}
@app.get("/results")
def results():
    db = SessionLocal()
    rows = (db.query(Prediction)
              .order_by(Prediction.created_at.desc())
              .limit(20).all())
    db.close()
    return {"results": [
        {"id": r.id,
         "prediction": r.prediction,
         "confidence": r.confidence,
         "model_version": r.model_version,
         "created_at": r.created_at.isoformat()}
        for r in rows]}
\end{verbatim}

\newthought{Restart the service.}
On \texttt{dev} the running \texttt{uvicorn -{}-reload} picks up the new
code on save. On \texttt{prod} restart through systemd so the unit
re-reads \texttt{app/main.py}:

\begin{verbatim}
sudo systemctl restart pixelwise   # prod only
\end{verbatim}

\newthought{Smoke test the wiring} with one unauthenticated
\texttt{GET /results} call. An empty table is exactly what you want
here: A successful response with an empty list proves the engine
opened, the session works, the query ran, and the JSON shape is right. Any of those broken and the call
returns a 500, not an empty list:

\begin{verbatim}
# on dev, against the local uvicorn
curl -s http://localhost:8000/results \
  | python3 -m json.tool

# on dev, against prod via the API gateway
curl -s http://192.168.56.11:8000/results \
  | python3 -m json.tool
\end{verbatim}

\marginnote{\newthought{Commit and tag \texttt{v0.5}.}\\[0.4em]
\texttt{git add app/main.py app/models.py}\\
\texttt{\ \ init\_db.py requirements.txt}\\
\texttt{\ \ .env.example}\\
\texttt{git commit -m "Add PostgreSQL persistence"}\\
\texttt{git tag v0.5}\\
\texttt{git push -{}-follow-tags}\\[0.3em]
Stage on \texttt{dev}, commit with a message that names the change,
tag the result, and push both the commit and the tag in one shot.}

\marginnote{\newthought{Optional: Feed it payload.}\index{curl}
So far our flow has never been hit by real digit drawings, and on
that note testing really came short. Build a 28-by-28 zero payload
once, perhaps wrapped in a \texttt{probe\_init.py}, then use it to
test the API and the database.}




% ==============================================================================================
\newpage
\section{Optional Exercise: First Migration with Alembic}\index{Alembic}\index{migration}\index{exercises}

\newthought{Schemas always change.}
You design the table, deploy it, store 500 predictions. Then you realise
you forgot a column. You cannot drop the table and recreate it; those 500
rows are real data. You need a tool that alters the live schema without
losing them. This exercise is optional: The mainline of the course does
not depend on it, but running it once gives you the feel of the migration
cycle on a real database. The example below adds a \texttt{client\_ip}
column for debugging, but any new column works.

\marginnote{Alembic: \url{https://alembic.sqlalchemy.org/}\\
\newthought{The metaphor.}\index{migration}
A migration is a recipe that describes a change to the schema and how to
roll it back.
The migration tool keeps an internal log of which migrations have run,
applies new ones in order, and refuses to run the same one twice.
The result is that the schema in the database always matches the
migrations folder in Git.}

\newthought{Initialise Alembic on dev.}
Alembic is the migration tool for SQLAlchemy; the cycle is \texttt{init},
\texttt{revision -{}-autogenerate}, \texttt{upgrade head}, run in that
order. All file edits in this exercise happen on \texttt{dev}.
\texttt{alembic init} writes its scaffold into the current directory, so
run everything from the repository root:

\begin{verbatim}
cd ~/pixelwise
source .venv/bin/activate
pip install alembic
pip freeze > requirements.txt
alembic init alembic
\end{verbatim}

The \texttt{pip freeze} step captures the new package in
\texttt{requirements.txt} so \texttt{prod} picks it up on the next
deploy.

Edit \texttt{alembic/env.py} so autogenerate sees your tables and runs
against the same connection string the application builds, rather than the
\texttt{sqlalchemy.url} placeholder in \texttt{alembic.ini}. Two edits do
this: point \texttt{target\_metadata} at \texttt{Base.metadata}, and make
\texttt{run\_migrations\_online} connect through the application
\texttt{engine}.

\begin{verbatim}
# alembic/env.py
from app.models import Base, engine
target_metadata = Base.metadata

def run_migrations_online() -> None:
    with engine.connect() as connection:
        context.configure(
            connection=connection,
            target_metadata=target_metadata)
        with context.begin_transaction():
            context.run_migrations()
\end{verbatim}

Without these two edits autogenerate connects to the placeholder URL and
sees no metadata, so it emits an empty migration with no
\texttt{add\_column} operation.

\marginnote{\newthought{Read the generated config.}\index{alembic.ini}
\texttt{alembic init alembic} writes a folder of starter files.
Skim them once before you change anything; the structure is small and the
mental model becomes obvious by reading the templates.}

\newthought{Add a new column} to the model. In
\texttt{\textasciitilde{}/pixelwise/app/models.py} on \texttt{dev}, add
\texttt{client\_ip = Column(String)} between the \texttt{confidence} and
\texttt{model\_version} lines.

\newthought{Generate the migration} from the repository root on
\texttt{dev}:

\begin{verbatim}
cd ~/pixelwise
alembic revision --autogenerate \
    -m "add client_ip column"
\end{verbatim}

\marginnote{\newthought{Read the migration before you run it.}\index{Alembic!autogenerate}
The generated script lives in \texttt{alembic/versions/}.
Open it, confirm the only operation is
\texttt{add\_column("client\_ip")}, and only then run \texttt{upgrade
head}. A wrong migration on a populated table is much harder to undo than
to prevent.}

\newthought{Wire \texttt{client\_ip} into the handler} in
\texttt{\textasciitilde{}/pixelwise/app/main.py} on \texttt{dev}.
The column reads the caller address off the request, so \texttt{classify}
needs the FastAPI \texttt{Request} object. If you ran the optional rate
limiter, \texttt{request: Request} is already a parameter and
\texttt{Request} is already imported; otherwise add
\texttt{from fastapi import Request} and a \texttt{request: Request}
parameter to \texttt{classify} first. Then the only further change is one
keyword argument in the \texttt{Prediction(...)} call:

\begin{verbatim}
db.add(Prediction(
    prediction=result["prediction"],
    confidence=result["confidence"],
    client_ip=request.client.host,
    model_version="v1"))
\end{verbatim}

\newthought{Commit and deploy.}
On \texttt{prod}, order matters: Install the new pinned dependency and
apply the migration first, then run \texttt{setup-server.sh}.
\texttt{setup-server.sh} restarts the service with code that references
the new column, so if the migration has not run yet, the handler will try
to INSERT a column that does not exist in the database. From
\texttt{dev}, stage the changes and push:

\begin{verbatim}
cd ~/pixelwise
git add app/main.py app/models.py alembic/ \
    alembic.ini requirements.txt
git commit -m "Add client_ip column via Alembic"
git push
\end{verbatim}

Then on \texttt{prod}, pull, install the new dependency, run the
migration, and only then restart the service:

\begin{verbatim}
cd /opt/pixelwise
git pull
source .venv/bin/activate
pip install -r requirements.txt
alembic upgrade head
bash setup-server.sh
\end{verbatim}

\newthought{Verify on \texttt{prod} with \texttt{psql}:}

\begin{verbatim}
psql -U pixelwise -d pixelwise -h localhost \
  -c "SELECT prediction, client_ip
      FROM predictions
      ORDER BY created_at DESC LIMIT 5;"
\end{verbatim}

Old rows show \texttt{NULL} for \texttt{client\_ip}; new predictions
store the caller's address now that the field is wired in. That gap is
the visible signature of a migration: The column exists everywhere, but
its values populate only from the moment the new code shipped.

% ==============================================================================================
% BLOCK 3: CONSOLIDATION
% ==============================================================================================

\newpage
\section{Self-Reflection and Recap}\index{reflection}\index{recap}

\newthought{Self-Reflection} questions to guide your thoughts during and after the exercises:
\begin{itemize}
    \item Why do we store predictions in a database rather than just returning them in the API response?
    \item What is the difference between dropping and recreating a table versus running a migration?
    \item If you ran the optional Alembic exercise, why do old rows show \texttt{NULL} for the new column while later rows carry a value?
    \item What happens if you store the database password directly in your Python code instead of in \texttt{.env}?
    \item Why is a dedicated database user with minimal privileges better than connecting as the superuser?
    \item How does the ORM make SQL injection harder, and where does the protection break down?
    \item Why does it pay to think the schema through up front rather than adding columns later?
    \item If you did the optional Alembic section: what is the difference between dropping and recreating a table versus running a migration, and why did the new column show \texttt{NULL} on every existing row?
\end{itemize}

\newthought{Recap} of key concepts:
\begin{itemize}
    \item Relational databases store data in typed tables; SQL is the language that queries and changes them.
    \item SQLite is the database for this course: a single file, no server, no users, no passwords. PostgreSQL is the production next step we stop short of here.
    \item SQLAlchemy maps tables to Python classes and uses parameterised queries by default, so the model code looks the same against either backend.
    \item The one database secret, \texttt{DB\_PASSWORD}, lives in \texttt{.env}; \texttt{models.py} builds the connection string from it, so the same code runs on \texttt{dev}, on \texttt{prod}, and under \texttt{systemd} alike.
    \item Optionally, Alembic generates and applies migrations so schemas can evolve without losing rows.
\end{itemize}

\marginnote{\newthought{Teaser.} We classify, we store, we query, but still no user can touch our backend.}

\newthought{Bridge.} The service classifies, stores, and
queries, but the only way in is \texttt{curl} with an API key on a
non-standard port. No browser, no URL, no human-friendly surface. This needs to change
next we put \texttt{nginx} in front of the FastAPI process, serves a small
drawing frontend over plain HTTP and HTTPS, and route browser traffic
into the same \texttt{POST /classify} and \texttt{GET /results} handlers
you just wired up. The database stays where it is; the prediction it
stores will start arriving from a canvas in a browser rather than a
shell command.
