% !TEX root = ../notes_missingsemester.tex
% Chapter 8: Automate Everything: CI/CD, Cron, Webhooks & Backups
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% POST IMPROVEMENTS 2026:
% 1. Pre-bake a v0.6 reference repo so the catch-up takes a single clone.
% 2. Add a TikZ figure of the cron timeline alongside the pipeline figure.

\index{CI/CD}\index{GitHub Actions}\index{cron}\index{backup}\index{feature flag}\index{webhook}\index{pg\_dump}\index{rsync}\index{YAML}\index{SSH}
\chapter{Automate Everything: CI/CD, Cron, Webhooks \& Backups}
\printchapterversion{08_cicd_automation}
\newthought{Manual deploys are technical debt. Automate or regret.}

% ==============================================================================================
% BLOCK 1: INTRODUCTION
% ==============================================================================================

\section{The Why}\index{automation}\index{deploy}

Block~7 left us with a service a human can use.
The browser at \texttt{https://192.168.56.11} renders a canvas, the
canvas posts to \texttt{/api/classify}, Nginx forwards the request to
FastAPI, FastAPI writes a row to PostgreSQL, and the page renders the
answer.
The contract is signed at every layer.

\newthought{What none of that delivers} is the day after the demo.
Every deploy is still a human ssh'ing into \texttt{prod}, running
\texttt{git pull}, restarting the service, and hoping nothing broke.
Every test is whatever the author thought to type into \texttt{curl}
before pushing.
The database has no backup, so the day the disk dies, every prediction
the service has ever stored dies with it.
Behaviour can only change by editing code and shipping a new version,
so the smallest tweak goes through the same heavyweight path as a real
feature.

\newthought{This Block} closes those gaps.
Every push runs a linter and a test suite on a clean machine before
anything lands on \texttt{prod}.
Every merge to \texttt{main} ships to \texttt{prod} without a human
touching a terminal.
A scheduled task on \texttt{prod} dumps the database every night, and
the dump moves to \texttt{dev} over Secure Shell, SSH, so the backup
survives a dead \texttt{prod}.
A flag in \texttt{.env} gates one new code path on the API, so we can
ship the path silent and switch it on without redeploying.

\begin{figure}
\centering
\begin{tikzpicture}[
    every node/.style={font=\footnotesize, align=center},
    box/.style={rectangle, draw=black, thick, rounded corners=2pt, minimum width=2.0cm, minimum height=0.9cm, inner sep=4pt},
    arr/.style={-latex, thick, black!70}
]
\node[box] (dev) at (-1.0, 0) {\texttt{git push}\\from \texttt{dev}};
\node[box] (gh)  at (2.4, 0)  {GitHub\\Actions};
\node[box] (lint) at (5.4, 0) {\texttt{ruff}\\lint};
\node[box] (test) at (8.0, 0) {\texttt{pytest}\\test};
\node[box] (dep) at (5.4, -1.8) {SSH deploy\\to \texttt{prod}};
\node[box] (prod) at (8.0, -1.8) {\texttt{systemctl}\\restart};
\draw[arr] (dev.east) -- (gh.west);
\draw[arr] (gh.east) -- (lint.west);
\draw[arr] (lint.east) -- (test.west);
\draw[arr] (test.south) -- ++(0,-0.45) -| (dep.north);
\draw[arr] (dep.east) -- (prod.west);
\node[font=\scriptsize, anchor=west] at (-1.0, -1.8) {\texttt{main} only};
\end{tikzpicture}
\caption{The Continuous Integration, CI, and Continuous Deployment, CD, pipeline. Every push triggers a GitHub Actions run that lints with \texttt{ruff} and tests with \texttt{pytest}. The deploy job fires only for pushes to \texttt{main}, opens an SSH session to \texttt{produser@192.168.56.11}, and runs \texttt{setup-server.sh} to reinstall dependencies, apply migrations, and restart the systemd unit. A red lint or red test breaks the chain before deploy.}
\label{fig:cicd_pipeline}
\end{figure}

\newthought{Why now and not earlier?}
Automation has prerequisites.
You cannot deploy automatically until the service deploys by hand,
which Block~5 fixed, and you cannot back up a database until there
is a database, which Block~6 fixed.
The reverse proxy from Block~7 means the production URL is stable,
which is what makes a webhook deployment safe.
Each layer below makes the layer above worth automating.

\newpage
\subsection{Hands On Experience}

\newthought{Consider this scenario:} It is Friday at 18:30.
A user reports that \texttt{/api/classify} returns the wrong label
on images that contain a stray ink mark.
You write a fix, run it locally on \texttt{dev}, the prediction
looks right, you push to \texttt{main}, ssh to \texttt{prod}, pull,
restart, and head home.
Saturday morning the service is down.
The fix passed your single manual check and shipped a regression
that broke the response shape on every other input.
No tests ran the existing endpoints, no guard sat between your push
and \texttt{prod}, and the only person who would have noticed before
the weekend ended is you.

\newthought{The lesson generalises.}
Every undocumented manual step is a step a tired human will skip,
and every skipped step is a regression waiting to ship.
Automation does not replace careful engineering: It encodes the
steps careful engineers already do, so they happen every time, in
the same order, without anyone remembering.

\marginnote{\newthought{Outside the classroom.}\index{operations}
Teams call this practice DevOps.
The artefacts are different at scale, with Jenkins, GitLab CI, or
Argo replacing GitHub Actions, and Kubernetes deployments replacing
\texttt{systemctl restart}, but the shape is identical: One human
push, many automated gates, one observable change to production.}

\newthought{This eighth block} introduces automation around the
PixelWise pipeline.
By the end you will have
\begin{itemize}
    \item written a GitHub Actions workflow that lints and tests on every push,
    \item written a smoke test for \texttt{classify\_batch} and an HTTP test for \texttt{POST /classify},
    \item shipped a deploy job that ssh's into \texttt{prod} on every merge to \texttt{main},
    \item scheduled a nightly database backup on \texttt{prod} through \texttt{cron},
    \item moved each backup off the \texttt{prod} VM with \texttt{rsync} over SSH,
    \item and gated one new API behaviour behind a feature flag read from \texttt{.env}.
\end{itemize}

\marginnote{\newthought{The pipeline is the product surface.}\index{pipeline}
Once a pipeline runs on every push, the test suite, the deploy script,
and the cron schedule become part of how the application behaves.
A broken pipeline is a broken product, even when the running service
is fine.}


% ==============================================================================================
\newpage
\section{Optional Exercise: Catch Up to \texttt{v0.6}}\index{exercises}\index{Git!tag}\index{recovery}

\marginnote{\newthought{Tag map.}\index{Git!tag}
\texttt{v0.6} marks the end of Block~7, \texttt{v0.7} will mark the
end of this block.
Browse the full set at
\url{https://github.com/schutera/pixelwise/tags}.}

\newthought{Catching up via tag.}\index{Git!tag}\index{recovery}
This catch-up extends the one in Block~7.
If you have not done that one, do it first: It lands you at
\texttt{v0.5}, the end-of-Block~6 state.
The exercise here picks up from there and rewinds to \texttt{v0.6},
the end-of-Block~7 state, with Nginx running on \texttt{prod} as a
reverse proxy in front of the FastAPI service, a static frontend
served from \texttt{/opt/pixelwise/frontend/}, a self-signed Transport
Layer Security, TLS, certificate in place, the site reachable at
\texttt{https://192.168.56.11}, Cross-Origin Resource Sharing, CORS,
configured on the API, and security headers set on every response.

\newthought{Both VMs this time.} Block~7 changed both the application
code and the public surface on \texttt{prod}, so this catch-up brings
both machines up to \texttt{v0.6}. The Git-tracked changes since
\texttt{v0.5} include the new \texttt{frontend/} directory with the
HyperText Markup Language, HTML, JavaScript, and Cascading Style
Sheets, CSS, that render the drawing canvas, an \texttt{nginx/}
directory carrying the reverse-proxy configuration committed to the
repository, an extended \texttt{setup-server.sh} that installs Nginx,
deploys the frontend, and provisions the self-signed TLS certificate,
updated CORS middleware in \texttt{app/main.py} together with the
security-header configuration, and a refreshed \texttt{.env.example}
with any new entries Block~7 introduced. The local \texttt{.env} on
each VM needs the new values written by hand;
\texttt{setup-server.sh} then reloads Nginx and restarts the service
in one shot.

\marginnote{\newthought{Sequence on \texttt{dev}.}\\[0.4em]
\texttt{cd \textasciitilde/pixelwise}\\
\texttt{git fetch origin -{}-tags}\\
\texttt{git reset -{}-hard v0.6}\\
\texttt{cp .env.example .env}\\
\texttt{nano .env}\\
\texttt{source .venv/bin/activate}\\
\texttt{pip install -r requirements.txt}\\
\texttt{bash setup-server.sh}\\[0.3em]
Edit \texttt{.env} to set any new secrets and local overrides. If
\texttt{.venv/} is missing, run
\texttt{python3 -m venv .venv} before the
\texttt{source} step.}

\newthought{Run the catch-up on \texttt{dev}.}
Rewind the repository to \texttt{v0.6}, refresh \texttt{.env} from the
updated example and edit it for any new secrets, reinstall pinned
dependencies if your virtual environment was wiped, then run
\texttt{setup-server.sh} to align the development tree with the
release. The exact commands are in the margin.

\marginnote{\newthought{Sequence on \texttt{prod}.}\\[0.4em]
\texttt{ssh produser@192.168.56.11}\\
\texttt{cd /opt/pixelwise}\\
\texttt{git fetch origin -{}-tags}\\
\texttt{git reset -{}-hard v0.6}\\
\texttt{cp .env.example .env}\\
\texttt{nano .env}\\
\texttt{source .venv/bin/activate}\\
\texttt{pip install -r requirements.txt}\\
\texttt{bash setup-server.sh}\\[0.3em]
Then from \texttt{dev}:
\texttt{curl -k https://192.168.56.11/api/health}
should return \texttt{\{"status":"ok"\dots\}}.}

\newthought{Run the catch-up on \texttt{prod}.}
Same sequence on \texttt{prod}, where \texttt{setup-server.sh} also
installs Nginx, deploys the \texttt{frontend/} files, provisions the
TLS certificate, reloads Nginx, and restarts the service. After the
script finishes, verify the API is reachable from \texttt{dev} over
the proxied path.

\marginnote{\newthought{Already have a fork?}\index{Git!fork}
If your own fork carries the \texttt{v0.6} tag, swap \texttt{origin}
on \texttt{dev} with
\texttt{git remote set-url origin git@github.com:<you>/pixelwise.git}
so push and pull continue to target your account.
The HTTPS clone on \texttt{prod} stays read-only.}

\newthought{You are now} at the state where Block~7 ended, with both
machines aligned, ready to start Block~8.



% ==============================================================================================
% BLOCK 2: CONCEPTS AND PRACTICE
% ==============================================================================================

\newpage
\section{Continuous Integration and Continuous Deployment}\index{CI/CD}\index{continuous integration}\index{continuous deployment}

\newthought{Continuous Integration, CI,} is the practice of merging
every change into a shared branch as soon as it passes an automated
gate.
The gate is a pipeline: A linter that rejects broken style, a test
suite that rejects broken behaviour, a build step that rejects code
that does not even compile.
The pipeline runs on a clean machine, not on the author's laptop, so
the result is reproducible.
Green is safe to merge; red does not enter \texttt{main}.

\marginnote{\newthought{Why a clean machine?}\index{reproducibility}
Your laptop has packages your colleague's laptop does not.
A pipeline that runs on a fresh \texttt{ubuntu-latest} image each
time catches the missing dependency the day the author pushes, not
the day the colleague clones.}

\newthought{Continuous Deployment, CD,} extends the chain by one
step.
The pipeline that passes the gate also pushes the green commit to
production, with no human between the merge and the restart.
The unit of release shrinks from "a quarter of features" to "every
merge to \texttt{main}", which makes each release small enough to
reason about and small enough to revert.
A regression that ships in a five-line commit is a regression you
can find in five lines.

\marginnote{\newthought{Continuous Delivery vs Continuous Deployment.}\index{continuous delivery}
Two practices share the abbreviation CD.
Continuous Delivery means every merge produces a release artefact
ready to deploy, with a human pressing the button.
Continuous Deployment removes the button.
PixelWise runs the latter because the surface is small and the cost
of a bad deploy is bounded by a one-line revert.}

\newthought{The PixelWise pipeline} has three gates and one effect.
Gate one is \texttt{ruff}, the Python linter.
Gate two is \texttt{pytest}, which runs the test suite against the
checked-out code.
Gate three is the branch filter on the deploy job: It runs only when
the push lands on \texttt{main}.
The effect is one SSH session from the runner into
\texttt{produser@192.168.56.11}, followed by
\texttt{bash setup-server.sh}.

\newthought{Each gate gates the next.}
A failing lint stops the test step.
A failing test stops the deploy.
Lint is cheap so it runs first, tests are slower so they only run on
code that lints, deploy is the only step that changes \texttt{prod}
so it only runs on code that lints and tests.
The shape is a pyramid with the expensive, irreversible work at the
top.

\marginnote{\newthought{Why \texttt{ruff} and \texttt{pytest}?}\index{ruff}\index{pytest}
\texttt{ruff} is a Python linter written in Rust that runs ten times
faster than \texttt{flake8} and replaces a dozen older tools with one
binary.
\texttt{pytest} is the de-facto test runner for Python: A function
named \texttt{test\_*} is a test, \texttt{assert} is the assertion,
no class scaffolding required.
Both are pinned in \texttt{requirements.txt} so the CI image and
\texttt{dev} agree on the version.\\
ruff: \url{https://docs.astral.sh/ruff/}\\
pytest: \url{https://docs.pytest.org/}}



% ==============================================================================================
\newpage
\section{GitHub Actions}\index{GitHub Actions}\index{YAML}

\newthought{GitHub Actions} runs workflows on machines GitHub
provides, in response to events on the repository.
The configuration lives in YAML, short for Yet Another Markup
Language, the indentation-sensitive serialisation format that maps
nicely onto nested job and step definitions.
Files go under \texttt{.github/workflows/}; every file in that folder
becomes one workflow.

\marginnote{GitHub Actions:
\url{https://docs.github.com/en/actions}\\
Workflow syntax:
\url{https://docs.github.com/en/actions/using-workflows/workflow-syntax-for-github-actions}}

\newthought{Four top-level keys} shape every workflow.
\texttt{name} is the human-readable label that shows up on the Actions
tab.
\texttt{on} declares the triggers, typically \texttt{push} for every
commit and \texttt{pull\_request} for proposed changes that have not
landed on \texttt{main} yet.
\texttt{jobs} is a map from job name to job definition.
Each job pins a runner image with \texttt{runs-on} and lists its
\texttt{steps}, which run in order.

\marginnote{\newthought{Runners.}\index{GitHub Actions!runner}
A runner is the VM that executes the workflow.
\texttt{ubuntu-latest} is a fresh Ubuntu image GitHub provisions on
every run, throws away when the job finishes, and bills against the
account's free-tier minutes.
Self-hosted runners exist for teams that need custom hardware or
private network access; the classroom does not.}

\newthought{Steps} come in two flavours.
\texttt{uses} pulls a prebuilt action, a packaged unit of work
published by GitHub or the community;
\texttt{actions/checkout@v4} checks out the repository at the
triggering commit, \texttt{actions/setup-python@v5} installs a
specific Python version.
A \texttt{run} step is a shell command, executed in the runner's
working directory with the repository already checked out.

\newthought{Secrets are values the workflow needs but the repository
must not store in plain text:} An SSH private key, a database
password, a third-party API token.
GitHub stores them encrypted under Settings, then Secrets and
variables, then Actions, and exposes them inside the workflow as
\texttt{\$\{\{ secrets.NAME \}\}}.
The runtime substitutes the value, the logs mask any line that
contains it, and the same value never lands in Git.

\marginnote{\newthought{Why mask in logs?}\index{secrets}
Action logs are world-readable on public repositories.
A workflow that prints a secret prints it forever.
The only safe rule is to never \texttt{echo} them.}

\newthought{Matrices and environments} are advanced features the
PixelWise pipeline does not use yet.
A matrix multiplies one job into many, useful for running the same
tests against three Python versions in parallel.
An environment is a named protection layer with required reviewers
and dedicated secrets, useful when the same workflow targets staging
and production.



% ==============================================================================================
\newpage
\section{Exercise: The Continuous Integration Workflow}\index{exercises}\index{GitHub Actions}

\newthought{Author \texttt{.github/workflows/ci.yml} on \texttt{dev}.}
The workflow runs on every push and every pull request, sets up
Python 3.11, installs pinned dependencies plus \texttt{ruff} and
\texttt{pytest}, lints, and runs the test suite.
Create the directory and the file at the repository root:

\begin{verbatim}
cd ~/pixelwise
mkdir -p .github/workflows
nano .github/workflows/ci.yml
\end{verbatim}

\marginnote{\newthought{The workflow file is the contract.}\index{GitHub Actions!workflow}
Everything CI does on every push is in this file.
A reviewer should be able to read it in one sitting and predict the
outcome of any push, without opening the Actions tab.}

Paste the workflow body.
The \texttt{name} surfaces in the Actions tab; the \texttt{on} block
fires the workflow on both branch pushes and pull requests; the
\texttt{lint-and-test} job runs on a fresh Ubuntu runner; each step
either uses a prebuilt action or runs a shell command:

\begin{verbatim}
# .github/workflows/ci.yml
name: CI
on: [push, pull_request]

jobs:
  lint-and-test:
    runs-on: ubuntu-latest
    steps:
      - uses: actions/checkout@v4

      - name: Set up Python
        uses: actions/setup-python@v5
        with:
          python-version: "3.11"

      - name: Install dependencies
        run: |
          python -m venv .venv
          source .venv/bin/activate
          pip install -r requirements.txt
          pip install ruff pytest

      - name: Lint
        run: ruff check .

      - name: Test
        run: pytest tests/
\end{verbatim}

\marginnote{\newthought{Why pin the Python version?}\index{Python!version}
\texttt{ubuntu-latest} ships a Python that drifts over time.
Pinning \texttt{3.11} means the runner installs the same minor version
on every run, which keeps the CI outcome reproducible across months.
The \texttt{dev} VM should match.}

\newthought{Commit and push to watch it run.}
The workflow registers itself the moment it lands in \texttt{main},
so the first push that includes the file also triggers the first run:

\begin{verbatim}
git add .github/workflows/ci.yml
git commit -m "Add CI workflow"
git push
\end{verbatim}

\marginnote{\newthought{Did it take?}\\[0.4em]
Open the repository's Actions tab on GitHub.
The latest run should be listed with a yellow dot while it runs and a
green check when it finishes.
A red cross means a step failed; expand the run and read the failing
step's log.}

\newthought{Expect a red run on the first push.}
The repository does not yet contain a \texttt{tests/} directory, so
\texttt{pytest} exits with code 5, "no tests collected".
The next section adds the first tests and turns the run green.



% ==============================================================================================
\newpage
\section{Writing the First Tests}\index{testing}\index{pytest}

\newthought{Testing is its own discipline.}
A full treatment of unit testing, integration testing, contract
testing, fuzzing, and property-based testing belongs in a dedicated
course.
What you need here is the smallest set of tests that prove the
service still works after a push.
Two tests is enough: One for the classifier function directly, one
for the HTTP layer on top of it.

\marginnote{\newthought{The test pyramid.}\index{testing!pyramid}
Many fast unit tests at the base, fewer integration tests in the
middle, a handful of end-to-end tests at the top.
The classifier test is a unit test, the HTTP test is an integration
test.
The end-to-end test, a browser driving the canvas, lands in a later
block.}

\newthought{Where tests live.}
The convention in Python is a \texttt{tests/} directory at the
repository root, with files named \texttt{test\_*.py} and functions
named \texttt{test\_*}.
\texttt{pytest} discovers them automatically; no registration, no
configuration.
The CI workflow above runs \texttt{pytest tests/}, which scans the
directory once per push.

\newthought{What a test looks like.}
A \texttt{pytest} test is a plain Python function that uses
\texttt{assert} to state a property the code must satisfy.
A pass is a function that runs to completion; a fail is any
exception, including \texttt{AssertionError}.

\newthought{FastAPI's \texttt{TestClient}}\index{FastAPI!TestClient}
wraps the application in an in-process HTTP client.
A call to \texttt{client.post("/classify", json=\{...\})} does the
same work \texttt{curl} does at the wire level, but without sockets,
without \texttt{uvicorn}, and without a port to free between tests.
The test runs in the same Python process as the app, so it is fast
enough that the CI gate stays cheap.



% ==============================================================================================
\newpage
\section{Exercise: A Smoke Test and an HTTP Test}\index{exercises}\index{pytest}

\newthought{Add \texttt{httpx} to the dev environment} so
\texttt{TestClient} has the asynchronous HTTP library it depends on,
then pin it in \texttt{requirements.txt} so the CI runner picks it
up:

\begin{verbatim}
cd ~/pixelwise
source .venv/bin/activate
pip install pytest httpx
pip freeze > requirements.txt
\end{verbatim}

\marginnote{\newthought{Why \texttt{httpx}?}\index{httpx}
FastAPI's \texttt{TestClient} delegates to \texttt{httpx} under the
hood since version 0.87.
The package is small, has no native build, and installs cleanly on
the CI runner.
The pin makes the test step reproducible.}

\newthought{Create \texttt{tests/test\_classifier.py}} on \texttt{dev}
with a smoke test for \texttt{classify\_batch}.
The test builds a batch of two blank images, passes them through the
classifier, and asserts the response shape.
A blank image is not a meaningful prediction, but it is meaningful
proof that the function runs, returns the right number of entries,
and exposes the expected keys:

\begin{verbatim}
mkdir -p tests
nano tests/test_classifier.py
\end{verbatim}

\marginnote{\newthought{Smoke test.}\index{testing!smoke}
A smoke test proves the wiring works without claiming anything about
correctness.
The name comes from electronics: Turn the power on and see whether
smoke comes out.
A smoke test that fails means the code does not even run; a smoke
test that passes does not mean the code is right, only that it
runs.}

\begin{verbatim}
# tests/test_classifier.py
import numpy as np
from app.classifier import classify_batch


def test_classify_batch_returns_one_per_image():
    images = np.zeros((2, 28, 28), dtype=np.uint8)
    results = classify_batch(images)
    assert len(results) == 2


def test_classify_batch_result_shape():
    images = np.zeros((1, 28, 28), dtype=np.uint8)
    result = classify_batch(images)[0]
    assert "prediction" in result
    assert "confidence" in result
    assert isinstance(result["confidence"], float)
\end{verbatim}

\newthought{Create \texttt{tests/test\_api.py}} for the HTTP layer.
\texttt{TestClient(app)} mounts the FastAPI application in-process,
so the test can issue requests without a running server.
The classify endpoint needs the API key header from Block~5, so the
test reads \texttt{SECRET\_API\_KEY} from the environment, which the
CI runner sets through a workflow secret in a later iteration; for
now, an explicit key keeps the test self-contained:

\marginnote{\newthought{Self-contained tests.}\index{testing}
A test that depends on a running \texttt{prod} server is not a test,
it is a probe.
\texttt{TestClient} pulls the application into the test process so
the test depends only on the code in the repository, which is what
makes the CI gate trustworthy.}

\begin{verbatim}
nano tests/test_api.py
\end{verbatim}

\begin{verbatim}
# tests/test_api.py
import os
from fastapi.testclient import TestClient
from app.main import app

os.environ.setdefault("SECRET_API_KEY", "test-key")
client = TestClient(app)


def test_health_is_ok():
    response = client.get("/health")
    assert response.status_code == 200
    assert response.json()["status"] == "ok"


def test_classify_round_trip():
    pixels = [[0] * 28 for _ in range(28)]
    response = client.post(
        "/classify",
        headers={"X-API-Key": "test-key"},
        json={"pixels": pixels},
    )
    assert response.status_code == 200
    body = response.json()
    assert "prediction" in body
    assert "confidence" in body
\end{verbatim}

\marginnote{\newthought{Why \texttt{setdefault}?}\index{environment variable}
\texttt{os.environ.setdefault} writes the key only when the
environment does not already define it.
A developer running \texttt{pytest} locally with a real \texttt{.env}
loaded keeps their value; a fresh CI runner with no
\texttt{SECRET\_API\_KEY} falls back to the test default.}

\newthought{Run the tests locally on \texttt{dev}} before pushing,
from the repository root so \texttt{pytest} resolves the
\texttt{app} package against the on-disk source tree:

\begin{verbatim}
cd ~/pixelwise
source .venv/bin/activate
pytest tests/
\end{verbatim}

\marginnote{\newthought{Did it take on \texttt{dev}?}\\[0.4em]
\texttt{pytest tests/} should print four passes.\\[0.3em]
\newthought{Wipe and start fresh.}
If \texttt{pytest} reports import errors, run
\texttt{pip install -r requirements.txt}
again to pull in \texttt{httpx} and the dependencies the test client
needs.
If the API test returns 401, double-check that
\texttt{SECRET\_API\_KEY} is set or that
\texttt{os.environ.setdefault} ran before
\texttt{from app.main import app}.}

\newthought{Commit and push.}
The CI workflow runs the same tests on the GitHub runner, against
the same \texttt{requirements.txt}.
A green run means lint and tests both passed; a red run means at
least one did not, and the Actions log says which:

\begin{verbatim}
git add tests/ requirements.txt
git commit -m "Add smoke tests for classifier and API"
git push
\end{verbatim}



% ==============================================================================================
\newpage
\section{Continuous Deployment}\index{continuous deployment}\index{deploy}

\newthought{The deploy job} is one extra job in the same workflow.
It depends on the CI job through the \texttt{needs} keyword, so it
runs only after lint and tests pass, and it filters on the branch
so it runs only for pushes to \texttt{main}.
A push to a feature branch triggers CI alone; a merge to
\texttt{main} triggers CI followed by deploy.

\marginnote{\newthought{Two jobs or two workflows?}\index{GitHub Actions!jobs}
One workflow with two jobs works because the second needs the first.
Splitting into two workflows is fine when the second is independent,
for example a nightly job.
The \texttt{needs} keyword is what makes the single-workflow shape
cheap.}

\newthought{The runner needs three things on \texttt{prod}.}
The IP, read from \texttt{secrets.SERVER\_IP} and set to
\texttt{192.168.56.11}.
The username, read from \texttt{secrets.SERVER\_USER} and set to
\texttt{produser}.
An SSH private key the runner can present to \texttt{prod}, generated
on \texttt{dev}, public half appended to \texttt{produser}'s
\texttt{authorized\_keys} on \texttt{prod}, private half stored in
\texttt{secrets.SSH\_PRIVATE\_KEY}.

\marginnote{\newthought{Why a deploy key?}\index{SSH!key}
A deploy key is a dedicated SSH key whose only job is to let the
runner reach \texttt{prod}.
It lives in GitHub Secrets and on \texttt{prod}'s
\texttt{authorized\_keys}, nowhere else.
Rotating the key revokes deploy access without touching any other
credential.}

\newthought{The deploy step} uses
\texttt{appleboy/ssh-action}\index{appleboy/ssh-action}, a community
action that opens an SSH session and runs a script on the remote
machine.
The script the classroom runs is the same
\texttt{bash setup-server.sh} a human would run by hand: It pulls
the latest \texttt{main}, installs pinned dependencies, applies any
schema migrations, copies the systemd unit, and restarts the
service.
The script is idempotent, so running it on every merge leaves the
server in a known state.

\newthought{The webhook alternative.}
The same shape works without GitHub Actions: A tiny HTTP server on
\texttt{prod} that listens for a POST and runs
\texttt{bash setup-server.sh} in response.
A listener on \texttt{prod} is one more attack surface, but it
removes the need to store an SSH key in GitHub.
The optional section near the end of this block sketches the
listener pattern; the mainline uses Actions because the runner is
trusted infrastructure GitHub already operates.



% ==============================================================================================
\newpage
\section{Exercise: The Deploy Job}\index{exercises}\index{SSH}\index{deploy}

\newthought{Generate the deploy key on \texttt{dev}} and copy the
public half to \texttt{produser}'s \texttt{authorized\_keys} on
\texttt{prod}:

\begin{verbatim}
ssh-keygen -t ed25519 \
    -f ~/.ssh/pixelwise_deploy \
    -C "pixelwise-deploy"
ssh-copy-id -i ~/.ssh/pixelwise_deploy.pub \
    produser@192.168.56.11
\end{verbatim}

\marginnote{\newthought{Why \texttt{ed25519}?}\index{SSH!key}
\texttt{ed25519} is the modern elliptic-curve key format: Smaller,
faster, and at least as strong as \texttt{rsa-4096}.
Every recent OpenSSH version supports it.}

\newthought{Register the secrets on GitHub.}
In the browser, open the repository on GitHub, then Settings, then
Secrets and variables, then Actions.
Add three repository secrets:
\texttt{SERVER\_IP} with value \texttt{192.168.56.11},
\texttt{SERVER\_USER} with value \texttt{produser}, and
\texttt{SSH\_PRIVATE\_KEY} with the entire contents of
\texttt{\textasciitilde/.ssh/pixelwise\_deploy}, including the
\texttt{-----BEGIN} and \texttt{-----END} lines.

\marginnote{\newthought{Paste the full key.}\index{secrets}
The private key is a multi-line text block.
Paste every line including the header and footer; an SSH client
rejects a truncated key with a confusing error like "invalid format"
that points nowhere useful.}

\newthought{Extend the workflow on \texttt{dev}.}
Open \texttt{.github/workflows/ci.yml} and append a second job after
\texttt{lint-and-test}.
The \texttt{needs} keyword chains the deploy job behind the lint and
test job, and the \texttt{if} guard restricts it to pushes that land
on \texttt{main}:

\begin{verbatim}
nano .github/workflows/ci.yml
\end{verbatim}

\marginnote{\newthought{The guard is the headline.}\index{GitHub Actions!if}
\texttt{if: github.ref == 'refs/heads/main'} is the line that turns
the workflow from "lint and test on every push" into "and only
deploy on \texttt{main}".
Without it, every feature branch would ship to \texttt{prod}.}

\begin{verbatim}
  deploy:
    needs: lint-and-test
    if: github.ref == 'refs/heads/main'
    runs-on: ubuntu-latest
    steps:
      - name: Deploy to prod
        uses: appleboy/ssh-action@v1
        with:
          host: ${{ secrets.SERVER_IP }}
          username: ${{ secrets.SERVER_USER }}
          key: ${{ secrets.SSH_PRIVATE_KEY }}
          script: |
            cd /opt/pixelwise
            git pull origin main
            source .venv/bin/activate
            pip install -r requirements.txt
            bash setup-server.sh
\end{verbatim}

\marginnote{\newthought{Why not call a script?}\index{deploy}
The inline \texttt{script} block is short enough to read on the
Actions log, which keeps the trail clear when a deploy fails.
A separate \texttt{scripts/deploy.sh} is fine once the steps grow;
for now, the inline version keeps everything in one file.}

\newthought{Commit and push to \texttt{main}.}
The first push triggers both jobs: Lint and test run first, the
deploy job follows on success, the runner opens an SSH session to
\texttt{produser@192.168.56.11}, and the server pulls the new commit
and restarts itself:

\begin{verbatim}
git add .github/workflows/ci.yml
git commit -m "Add deploy job for pushes to main"
git push
\end{verbatim}

\marginnote{\newthought{Did it deploy?}\\[0.4em]
From \texttt{dev}:\\
\texttt{curl -k https://192.168.56.11/api/health}\\[0.3em]
The response should include the current commit's behaviour, including
any test or workflow change you just pushed.
\texttt{sudo systemctl status pixelwise} on \texttt{prod} should show
a recent \texttt{Active: active (running)} timestamp matching the
deploy run.

\vspace{0.4em}
\newthought{Wipe and start fresh.}
If the deploy job hangs at the SSH step, the runner cannot reach
\texttt{192.168.56.11}.
The classroom VM lives on a host-only adapter, which means GitHub's
public runners cannot route to it.
The deploy job will only complete from a self-hosted runner or from
a runner with network access to the lab; the workflow is otherwise
correct and ships as written to any reachable production server.}


\marginnote{\newthought{Why no schema step.}\index{schema}
The \texttt{predictions} table is created once on \texttt{prod} by
\texttt{init\_db.py} during the Block~6 setup, and nothing later in the
course changes its shape. The deploy script therefore only ships code,
not schema. If a future change needs to alter the live schema, that is
the point at which the optional Block~6 Alembic detour pays off, and the
deploy script grows a single \texttt{alembic upgrade head} line above the
restart.}


% ==============================================================================================
\newpage
\section{Cron: Scheduled Tasks}\index{cron}\index{scheduling}

\newthought{Cron is the unix scheduler.}
A line in a crontab pairs a schedule expression with a command;
\texttt{cron} wakes up every minute, checks every line, and runs the
ones whose schedule matches.
The schedule is five fields: Minute, hour, day of month, month, day
of week.
A \texttt{*} matches every value, a number matches that value,
\texttt{*/N} matches every Nth value.

\marginnote{\newthought{Read the fields out loud.}\index{cron!syntax}
\texttt{0 2 * * *} reads as "minute 0, hour 2, every day, every
month, every weekday", which is two in the morning.
\texttt{*/5 * * * *} reads as "every five minutes".
\texttt{0 0 * * 0} reads as "Sunday at midnight".
The schedule field is the part most people misread; saying it out
loud usually fixes that.}

\newthought{Where the crontab lives.}
Every Linux user has a private crontab, edited with
\texttt{crontab -e} and listed with \texttt{crontab -l}.
The file form, \texttt{crontab path/to/file}, replaces the whole
crontab atomically.
The classroom uses the file form because the schedule is part of
how the service operates and belongs under version control next to
the systemd unit and the Nginx config.

\newthought{What runs on a schedule.}
Database backups, log rotation, certificate renewal, model artefact
refresh, cleanup of temporary files, health probes that alert when
they fail.
Anything that needs to happen reliably without a human remembering
is a cron candidate.
The PixelWise schedule runs one job: A nightly database dump that
moves off the \texttt{prod} VM before sunrise.

\marginnote{\newthought{Cron's failure modes.}\index{cron!debugging}
A cron job that fails is silent by default; \texttt{cron} mails the
output to the local user, which on a server with no mail relay goes
nowhere.
Redirect both streams to a known log file with
\texttt{>>/var/log/pixelwise-backup.log 2>\&1} so you have something
to read when the backup did not land.}

\newthought{Cron versus a webhook deploy.}
Both are automation, but they fail differently.
Cron runs on a fixed schedule, so a missed run is invisible until
someone checks the log.
A webhook deploy runs in response to a trigger, so a failed run
fails the trigger, and the operator sees the red Actions run
immediately.
Use cron for work that does not have a trigger; use a webhook deploy
for work that follows an event.



% ==============================================================================================
\newpage
\section{Database Backups}\index{backup}\index{pg\_dump}\index{rsync}

\newthought{What to back up.}
The database.
The code is on GitHub and reproducible from any tagged commit, the
model artefact is reproducible from \texttt{train.py} and the seed,
the configuration is reproducible from \texttt{setup-server.sh} and
\texttt{.env.example}.
The database is the only piece of state that exists only on
\texttt{prod} and cannot be recreated.

\marginnote{\newthought{Reproducible vs irreplaceable.}\index{state}
Reproducible state can be rebuilt from inputs you control.
Irreplaceable state is whatever a real user generated, and the
database fills with it after every deploy.
A backup strategy protects irreplaceable state first.}

\newthought{How to dump.}\index{pg\_dump}
\texttt{pg\_dump} writes the database to a SQL file, a plain-text
script of \texttt{INSERT} statements that any PostgreSQL of the same
or newer major version can replay.
The dump runs as the database owner and exits zero on success.
The cron line pipes through \texttt{gzip} so the compressed output
replaces the uncompressed one atomically when the dump succeeds.

\newthought{Where to put the dump.}
A backup on the same server as the database is a backup that dies
with the server.
The classroom has a natural off-site target: The \texttt{dev} VM,
which lives on the same host but is independent of \texttt{prod} in
every way that matters for this exercise.
\texttt{rsync} over SSH copies the dump to \texttt{dev}, where it
sits on a different disk away from any disaster that takes out
\texttt{prod}.

\marginnote{\newthought{Real off-site.}\index{backup}
In production, off-site means a different cloud region, provider, or
an air-gapped tape.
The classroom uses \texttt{dev} because the two-VM setup from
Block~1 makes it free; the pattern is identical, only the target
changes.}

\newthought{Retention and restore.}
A backup folder that grows forever fills the disk and obscures the
recent dumps.
A retention policy keeps the last N nights and deletes older files;
the classroom keeps the last seven.
A restore is the only test that proves the backup is real: A dump
that has never been restored is a dump that may or may not work, and
the night you find out is the night you cannot afford to.



% ==============================================================================================
\newpage
\section{Exercise: Nightly Backup on \texttt{prod}}\index{exercises}\index{backup}

\newthought{Author the backup script on \texttt{dev}.}
Create \texttt{deploy/backup.sh} at the repository root.
The script dumps the database, compresses the result with
\texttt{gzip}, writes the file under
\texttt{/var/backups/pixelwise/}, and prunes anything older than
seven days:

\begin{verbatim}
cd ~/pixelwise
mkdir -p deploy
nano deploy/backup.sh
\end{verbatim}

\marginnote{\newthought{Why a script and not an inline cron line?}\index{cron}
A script can be read, version-controlled, and reused.
A cron line that does the same work in one screen-wide command is
unreadable in the dark at 03:00 when the dump did not land.
The cron entry stays a one-liner that calls the script.}

\begin{verbatim}
#!/bin/bash
# deploy/backup.sh
# Nightly PostgreSQL dump for PixelWise.
set -euo pipefail

BACKUP_DIR="/var/backups/pixelwise"
STAMP="$(date +%Y%m%d)"
OUT="$BACKUP_DIR/pixelwise-$STAMP.sql.gz"

mkdir -p "$BACKUP_DIR"

PGPASSWORD="${DB_PASSWORD:-}" \
    pg_dump -U pixelwise -h localhost pixelwise \
    | gzip > "$OUT"

# Keep the last seven nights, prune the rest.
find "$BACKUP_DIR" -name 'pixelwise-*.sql.gz' \
    -mtime +7 -delete

echo "Backup complete: $OUT"
\end{verbatim}

\marginnote{\newthought{The \texttt{set -euo pipefail} line.}\index{bash!set}
\texttt{-e} exits on any non-zero command, \texttt{-u} errors on
unset variables, \texttt{-o pipefail} fails the whole pipeline if any
stage fails.
Without \texttt{pipefail}, a \texttt{pg\_dump} that crashes mid-stream
would still produce a valid empty gzip and a green script exit, which
is the worst failure mode a backup can have.}

\newthought{Mark the script executable and commit.}
The executable bit travels with the file in Git, so \texttt{prod}
gets it on pull without a second \texttt{chmod}:

\begin{verbatim}
chmod +x deploy/backup.sh
git add deploy/backup.sh
git commit -m "Add nightly backup script"
\end{verbatim}

\newthought{Author the cron file on \texttt{dev}.}
Two schedule lines: The backup script runs at 02:30, and
\texttt{rsync} copies whatever landed in
\texttt{/var/backups/pixelwise/} to the dev VM fifteen minutes later:

\begin{verbatim}
mkdir -p deploy/cron
nano deploy/cron/pixelwise.cron
\end{verbatim}

\marginnote{\newthought{Why 02:30?}\index{cron!schedule}
Local quiet hours, before any morning traffic, with enough margin
that the dump and the \texttt{rsync} both finish before anyone logs
in.
A real production schedule depends on the user base; 02:30 is a
classroom default.}

\begin{verbatim}
# deploy/cron/pixelwise.cron
# m h dom mon dow command
30 2 * * * /opt/pixelwise/deploy/backup.sh \
    >> /var/log/pixelwise-backup.log 2>&1

45 2 * * * rsync -avz \
    /var/backups/pixelwise/ \
    user@192.168.56.10:/backups/pixelwise/ \
    >> /var/log/pixelwise-backup.log 2>&1
\end{verbatim}

\marginnote{\newthought{Two lines, one effect.}\index{cron}
The first line writes the dump locally; the second copies it off-VM.
Splitting the two means a failed \texttt{rsync} does not lose the
local dump, and a failed dump does not trigger an \texttt{rsync} of
yesterday's file.
The output of both lines appends to the same log so you have one
place to look when something went wrong overnight.}

\newthought{Document \texttt{DB\_PASSWORD} for the backup user.}
The script reads \texttt{DB\_PASSWORD} from the environment because
PostgreSQL on \texttt{prod} expects the application user to present
a password.
The cron runner inherits a minimal environment, so either the
variable is exported from \texttt{.env} at the top of the script, or
the cron entry invokes the script under a wrapper that loads it.

\newthought{Teach \texttt{setup-server.sh} to install the schedule.}\index{setup-server.sh}
Append a prod-only stanza so a fresh \texttt{prod} VM picks up the
cron file, the backup directory, and the log file without any
manual steps.
On \texttt{dev}:

\begin{verbatim}
nano setup-server.sh
\end{verbatim}

\marginnote{\newthought{Idempotent by design.}\index{idempotence}
The stanza creates directories with \texttt{-p}, which is a no-op
when they already exist, and replaces the crontab with
\texttt{crontab <file>}, which is an atomic overwrite of the same
schedule on every run.
Run the script ten times; the schedule is in place exactly once.}

\begin{verbatim}
# Install the cron schedule on prod
if [ -f "$SCRIPT_DIR/deploy/cron/pixelwise.cron" ] \
   && command -v crontab >/dev/null 2>&1 \
   && id produser >/dev/null 2>&1; then
    sudo mkdir -p /var/backups/pixelwise
    sudo chown produser:produser \
        /var/backups/pixelwise
    sudo touch /var/log/pixelwise-backup.log
    sudo chown produser:produser \
        /var/log/pixelwise-backup.log
    crontab "$SCRIPT_DIR/deploy/cron/pixelwise.cron"
fi
\end{verbatim}

\newthought{Prepare the off-site target on \texttt{dev}.}
The \texttt{rsync} on \texttt{prod} writes into
\texttt{/backups/pixelwise/} on \texttt{user@192.168.56.10}.
Create the directory, give \texttt{user} ownership, then authorise
\texttt{produser} from \texttt{prod} to log in over SSH:

\begin{verbatim}
# on dev
sudo mkdir -p /backups/pixelwise
sudo chown user:user /backups/pixelwise

# on prod
sudo -u produser ssh-keygen -t ed25519 \
    -f /home/produser/.ssh/dev_backup -N ""
sudo cat /home/produser/.ssh/dev_backup.pub
\end{verbatim}

\marginnote{\newthought{Why a separate key?}\index{SSH!key}
The deploy key earlier lets the GitHub runner reach \texttt{prod}.
This backup key lets \texttt{prod} reach \texttt{dev}.
Two purposes, two keys; revoking one does not break the other.}

Paste the printed public key into \texttt{user@192.168.56.10}'s
\texttt{\textasciitilde/.ssh/authorized\_keys}.

\newthought{Commit, push, and deploy.}
Stage the cron file, the backup script, and the updated
\texttt{setup-server.sh}, then push.
The CI workflow runs; on green, the deploy job ships the new files
to \texttt{prod} and \texttt{setup-server.sh} installs the
schedule:

\begin{verbatim}
git add deploy/backup.sh \
    deploy/cron/pixelwise.cron \
    setup-server.sh
git commit -m "Add nightly database backup"
git push
\end{verbatim}

\marginnote{\newthought{Did the cron install?}\\[0.4em]
On \texttt{prod}:\\
\texttt{crontab -l}\\[0.3em]
Should print the two schedule lines exactly as they live in
\texttt{deploy/cron/pixelwise.cron}.
A missing entry means the prod-only guard in
\texttt{setup-server.sh} did not fire, usually because
\texttt{id produser} fails or the cron file is not in the expected
path.

\vspace{0.4em}
\newthought{Run it once by hand.}
Waiting until 02:30 to learn the script broke is bad style.
On \texttt{prod}:\\[0.3em]
\texttt{sudo -u produser bash /opt/pixelwise/deploy/backup.sh}\\[0.3em]
The script should print
\texttt{Backup complete: /var/backups/pixelwise/pixelwise-YYYYMMDD.sql.gz}
and the file should exist with non-zero size.}

\newthought{Verify the restore on \texttt{dev}.}
A backup that has never been restored is not a backup.
Create a throwaway database, replay the dump into it, and confirm
the row count matches:

\begin{verbatim}
sudo -u postgres createdb pixelwise_restore_test
gunzip -c /backups/pixelwise/pixelwise-*.sql.gz \
    | sudo -u postgres \
        psql -d pixelwise_restore_test

sudo -u postgres \
    psql -d pixelwise_restore_test \
    -c "SELECT COUNT(*) FROM predictions;"
\end{verbatim}

\marginnote{\newthought{Throwaway by name.}\index{restore}
\texttt{pixelwise\_restore\_test} signals to the next person that
the database exists only to validate a restore.
Drop it when you are done:\\[0.3em]
\texttt{sudo -u postgres dropdb pixelwise\_restore\_test}}



% ==============================================================================================
\newpage
\section{Feature Flags}\index{feature flag}

\newthought{A feature flag is a runtime switch.}
The code path exists in the binary, but a check at the start of the
path decides whether to take it.
The check usually reads a boolean from \texttt{.env}, an environment
variable, or a row in a config table.
Flipping the flag changes behaviour without changing code, without a
deploy, and without a build.

\marginnote{\newthought{The shape of the flag.}\index{feature flag}
A simple flag is a boolean.
A multi-arm flag is a string from a known set, like \texttt{v1} vs
\texttt{v2}.
A percentage flag rolls a feature out to a fraction of traffic.
The classroom uses the boolean form.}

\newthought{Why this beats commenting code in and out.}
A commented-out code path is invisible to the test suite, the type
checker, the linter, and any reviewer who searches for usages.
A flag-gated path is real code: It compiles, it runs the linter, it
runs in CI, and it lights up in production the moment the flag
flips.
The cost is one branch in the code and one line in
\texttt{.env}.

\newthought{What this enables.}
Progressive rollout, where you ship the new path off, watch the
service, flip the flag on, watch again, and roll back instantly if
anything breaks.
Dark launch, where you ship the path before the user-visible work is
ready and turn it on when the rest lands.
Operational kill switch, where a path that turns out wrong stays
gated until the fix ships, without a redeploy.

\marginnote{\newthought{Outside the classroom.}\index{feature flag}
Production teams use hosted flag systems like LaunchDarkly or
Unleash that store flags out of band and push changes to running
services in seconds.
The classroom uses \texttt{.env} because it is the lightest possible
implementation of the same pattern.}



% ==============================================================================================
\newpage
\section{Exercise: A Pagination Flag for \texttt{/results}}\index{exercises}\index{feature flag}

\newthought{Pick a behaviour to gate.}
\texttt{GET /results} from Block~6 returns the twenty newest
predictions.
A natural enhancement is to make the page size configurable through
a query parameter, defaulting to twenty, with a maximum of one
hundred.
The flag \texttt{FEATURE\_RESULTS\_PAGINATION} controls whether the
endpoint accepts the new parameter or ignores it.

\marginnote{\newthought{Small flag, real path.}\index{feature flag}
The endpoint reads the flag at request time, branches once, and the
rest of the handler stays unchanged.
Real flags wrap larger changes; the mechanics are identical.}

\newthought{Document the flag} in
\texttt{\textasciitilde/pixelwise/.env.example} on \texttt{dev}, and
add the same line to your local \texttt{.env} on both VMs.
A default of \texttt{false} keeps existing behaviour for any caller
that has not opted in:

\begin{verbatim}
nano .env.example
\end{verbatim}

\begin{verbatim}
FEATURE_RESULTS_PAGINATION=false
\end{verbatim}

\marginnote{\newthought{Why default off?}\index{feature flag}
A new flag that defaults to \texttt{true} ships the new behaviour on
every machine that picks up the new \texttt{.env.example}.
\texttt{false} keeps the existing contract and lets each environment
opt in explicitly.
The day you trust the new path, change the default; until then,
default off.}

\newthought{Read the flag in \texttt{app/main.py} on \texttt{dev}.}
The handler reads \texttt{FEATURE\_RESULTS\_PAGINATION} on every
call, so flipping the flag and restarting the service is enough to
change behaviour.
Open the file and replace the \texttt{results} handler:

\begin{verbatim}
nano app/main.py
\end{verbatim}

\marginnote{\newthought{Two new imports.}
\texttt{Query} marks the \texttt{limit} parameter as a query-string
field and gives FastAPI the type information it needs to validate
it.
\texttt{os} reads the flag from the environment.}

\begin{verbatim}
from fastapi import Query
import os
\end{verbatim}

\marginnote{\newthought{The branch is the headline.}\index{feature flag}
The first line of the handler reads the flag.
The branch is one \texttt{if}; everything else is shared.
A reviewer can see at a glance which path is gated, and a search for
\texttt{FEATURE\_RESULTS\_PAGINATION} finds every effect of the flag
across the codebase.}

\begin{verbatim}
@app.get("/results")
def results(limit: int = Query(20, ge=1, le=100)):
    pagination_on = (
        os.getenv("FEATURE_RESULTS_PAGINATION", "false")
        .lower() == "true"
    )
    page_size = limit if pagination_on else 20

    db = SessionLocal()
    rows = (db.query(Prediction)
              .order_by(Prediction.created_at.desc())
              .limit(page_size).all())
    db.close()
    return {"results": [
        {"id": r.id,
         "prediction": r.prediction,
         "confidence": r.confidence,
         "model_version": r.model_version,
         "created_at": r.created_at.isoformat()}
        for r in rows]}
\end{verbatim}

\newthought{Smoke test the flag off, then on.}
On \texttt{dev}, with \texttt{FEATURE\_RESULTS\_PAGINATION=false},
the \texttt{limit} parameter parses but is ignored and the response
holds at most twenty rows.
Flip the flag in \texttt{.env} to \texttt{true}, restart the running
\texttt{uvicorn} on \texttt{dev} or
\texttt{sudo systemctl restart pixelwise} on \texttt{prod}, and the
same \texttt{curl} honours \texttt{limit=50}:

\begin{verbatim}
curl -s "http://localhost:8000/results?limit=50" \
    | python3 -m json.tool | head
\end{verbatim}

\marginnote{\newthought{Did the flag take?}\\[0.4em]
With the flag off, the response length stays at twenty no matter
which \texttt{limit} value you pass.
With the flag on, the response length tracks \texttt{limit} up to
one hundred.
If the response is identical in both cases, the handler is reading
the flag once at import time instead of on every call; move the
\texttt{os.getenv} call inside the function body.}

\newthought{Commit the gated path.}
The flag, the example, and the handler change all land in one
commit so a future bisect can find them together:

\begin{verbatim}
git add app/main.py .env.example
git commit -m "Gate /results pagination behind feature flag"
git push
\end{verbatim}



% ==============================================================================================
\newpage
\section{Optional: A Webhook Listener on \texttt{prod}}\index{webhook}\index{Discord}

\newthought{The deploy job above relies on GitHub Actions.}
An alternative shape skips the runner entirely: A tiny HTTP listener
on \texttt{prod} accepts a POST from a webhook provider, validates a
shared secret, and runs \texttt{bash setup-server.sh} in response.
The trigger source can be GitHub, a Discord slash command, or any
other system that can POST to a URL.

\marginnote{\newthought{Trade-offs.}\index{webhook}
The listener removes the SSH key on GitHub at the cost of adding a
public endpoint on \texttt{prod}.
Both shapes work; the choice depends on which attack surface is
easier to defend in your environment.}

\newthought{The sketch.}
A FastAPI application with one POST endpoint, run by its own systemd
unit, mounted behind Nginx on a path the public does not know:

\begin{verbatim}
# deploy/webhook.py
import os, subprocess
from fastapi import FastAPI, Header, HTTPException

webhook_app = FastAPI()

@webhook_app.post("/deploy")
def deploy(x_token: str = Header(...)):
    if x_token != os.getenv("WEBHOOK_TOKEN"):
        raise HTTPException(status_code=401,
                            detail="Bad token")
    result = subprocess.run(
        ["bash", "/opt/pixelwise/setup-server.sh"],
        capture_output=True, text=True)
    if result.returncode != 0:
        raise HTTPException(status_code=500,
                            detail=result.stderr[-500:])
    return {"status": "deployed"}
\end{verbatim}

\marginnote{\newthought{The Discord variant.}\index{Discord}
Discord can fire a slash command at a webhook URL.
The handler shape is the same, with the secret in a Discord-specific
signature header instead of \texttt{X-Token}.
A team member types \texttt{/deploy} in a channel, the listener
runs, the chat shows the result.
This is genuinely how some teams ship.}

Wiring it up properly needs a second systemd unit, an Nginx
location, a hardened secret, and rate limiting on the endpoint, none
of which the mainline of this block needs.
The sketch is enough to recognise the shape when you see it in the
wild.



% ==============================================================================================
\newpage
\section{Tag \texttt{v0.7}}\index{Git!tag}

\newthought{The pipeline is the artefact.}
The end of every block is a tagged, runnable state.
Stage every file the block added or changed, commit, push, tag, and
push the tag.
Future catch-ups reset to this tag and arrive at the same place you
are now:

\marginnote{\newthought{What lands in \texttt{v0.7}.}\index{Git!tag}
\texttt{.github/workflows/ci.yml} with the lint, test, and deploy
jobs; \texttt{tests/} with the classifier and API tests;
\texttt{deploy/backup.sh} and \texttt{deploy/cron/pixelwise.cron}
for the nightly dump; an extended \texttt{setup-server.sh} that
installs the cron schedule; an updated \texttt{.env.example} with
\texttt{FEATURE\_RESULTS\_PAGINATION}; the gated
\texttt{/results} handler in \texttt{app/main.py}.}

\begin{verbatim}
cd ~/pixelwise
git add .github/workflows/ci.yml \
    tests/ \
    deploy/backup.sh \
    deploy/cron/pixelwise.cron \
    setup-server.sh \
    .env.example \
    app/main.py \
    requirements.txt
git commit -m "End of Block 8: CI/CD, cron, backups, flag"
git tag -a v0.7 \
    -m "End of Block 8: CI/CD, backups, feature flags"
git push --follow-tags
\end{verbatim}

\marginnote{\newthought{Watch the pipeline land its own tag.}\index{CI/CD}
The push that creates \texttt{v0.7} is a push to \texttt{main}, so
the deploy job fires on it.
The runner pulls the tagged commit on \texttt{prod}, installs the
cron schedule, restarts the service, and writes back its exit
status.
The first time a chapter ships its own deploy is a moment worth
pausing for.}

\newthought{The pipeline now runs without you.}
Every push is linted and tested.
Every merge to \texttt{main} ships.
Every night the database is dumped to \texttt{prod} and rsynced to
\texttt{dev}.
One code path is gated behind a flag, ready to switch on or off in
\texttt{.env} alone.
The service is automated.



% ==============================================================================================
% BLOCK 3: CONSOLIDATION
% ==============================================================================================

\newpage
\section{Self-Reflection and Recap}\index{reflection}\index{recap}

\newthought{Self-Reflection} questions to guide your thoughts during
and after the exercises:
\begin{itemize}
    \item If you automate an undocumented manual deploy, what guarantees the automation does not bake in a bug the manual process used to skip past?
    \item A test catches a regression on a feature branch; a user catches it in \texttt{prod}. What does each cost, and where in the pipeline is the cheapest place to catch the next one?
    \item A feature flag stays \texttt{false} for three months and the code path never runs. How is that different from code commented out, and what makes the flag-gated version safer?
    \item Cron runs at 02:30 and the backup fails silently. How would you find out before the next disaster, and what is the smallest change to your schedule that surfaces the failure?
    \item A webhook deploy fails because \texttt{setup-server.sh} hit a non-zero exit. How does that failure mode differ from a cron job that hits a non-zero exit, and why does the operator notice one but not the other?
    \item The deploy key sits in two places: GitHub Secrets and \texttt{authorized\_keys} on \texttt{prod}. What does rotating the key require, and what breaks if you forget either side?
    \item A backup that has never been restored is not a backup. Which of the steps in the restore exercise is the one most teams skip in real production, and why?
\end{itemize}

\newthought{Recap} of key concepts:
\begin{itemize}
    \item Continuous Integration runs lint and tests on every push from a clean machine; Continuous Deployment ships every green merge to \texttt{main} without a human in the loop.
    \item A GitHub Actions workflow is YAML under \texttt{.github/workflows/}; jobs run on runners, steps run in order, secrets live in repository settings and never in the file.
    \item Tests live in \texttt{tests/}, run with \texttt{pytest}, and double as the gate that CI uses to decide whether a push is safe to deploy.
    \item Cron schedules recurring work on a unix machine; the schedule lives under version control alongside the service it supports.
    \item Database backups dump to a file with \texttt{pg\_dump}, move off-VM with \texttt{rsync}, are pruned by age, and are only real after a successful restore.
    \item Feature flags gate code that lives in the binary, so flipping behaviour is a config change and a restart instead of a deploy.
\end{itemize}

\marginnote{\newthought{Security thread.}\index{security}
Three new credentials enter the system in this block.
The GitHub Actions deploy key on the runner reaches
\texttt{produser@192.168.56.11}; the backup SSH key on \texttt{prod}
reaches \texttt{user@192.168.56.10}; the database password the cron
job inherits to drive \texttt{pg\_dump}.
Each key has one purpose, each lives in one place, and each rotates
independently.
The backup files themselves carry every prediction the service has
ever stored; file permissions on \texttt{/var/backups/pixelwise/}
should restrict reads to \texttt{produser} alone, and any real
off-site target encrypts the dump at rest.}

\marginnote{\newthought{Teaser.}
The pipeline runs, the deploy lands, the service is alive.
How do you know it is healthy?}

\newthought{Bridge to Block~9.}
The service is now automated end to end, and the operator is blind.
There are no metrics on the request rate, no structured logs on the
classify path, no alert when a deploy lands and the error rate
spikes ten minutes later.
A green CI run says the code compiles, lints, and passes its tests;
a green deploy says \texttt{systemctl} agreed to restart; neither
says the service is doing the right thing for real users right
now.
Block~9 closes that gap.
Monitoring sees the service's pulse, structured logs make the
classify path searchable, and a Discord webhook turns a silent
overnight failure into a chat message that wakes the right person
up.
