% !TEX root = ../notes_unsupervisedlearning.tex
% Chapter 2: Centroid Clustering
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\index{Centroid Clustering}\index{K-Means}
\chapter{Centroid Clustering}
\newthought{From Points to Clusters}

% ==============================================================================================
% BLOCK 1: INTRODUCTION
% ==============================================================================================

\section{The Why}\index{centroid clustering}\index{K-Means}

In the previous chapter, we established that distance is a choice\cite{Bishop2006}:
it encodes what similar means in a given feature space.
Given that means to encode similarity, we can ask a more complex question,
one that goes beyond pairs of points: which points belong together?
Or in other words, is there an overarching structure inherent in the data?

\marginnote{\newthought{A partition} assigns every point to exactly one group, with no group left empty.
K-Means produces a \emph{hard} partition: membership is binary, not graded.
This is a strong assumption that later clustering methods will soften.}

Raw pairwise distances do not yet tell us which observations belong together,
nor how many groups exist.
We need a procedure that assigns each point to a cluster; together, all clusters form a partition of the data.

\newthought{Centroid-based clustering} offers one answer: represent each group by a single point, its centroid, and assign every observation to the nearest one.
This is a methodologically important commitment.
Rather than asking about every pairwise relationship, we compress each group into a single representative and reason from there.
The method thereby makes an assumption explicit: a group is well-described by its mean.
When that assumption holds, the algorithm is remarkably efficient; when it does not, the failure is instructive.

\newthought{The procedure was formalized} as K-Means by MacQueen\cite{MacQueen1967}; its standard algorithmic form, alternating assignment and centroid update steps, was described by Lloyd\cite{Lloyd1982}.
It remains one of the most cited algorithms in machine learning\cite{Jain2010}, in part because it is easy to implement, easy to interpret, and fast enough to run on large datasets.
The main practical weakness, sensitivity to initialization, was addressed by Arthur and Vassilvitskii\cite{ArthurVassilvitskii2007} with K-Means++, which provides a probabilistic seeding strategy with provable guarantees on solution quality.

\marginnote{\newthought{A centroid} is the mean position of all points assigned to a cluster.
It need not be an actual data point; it is the center of mass of the group.
For the update step to be well-defined, the centroid must exist in the same space as the data, which is why K-Means is tied to Euclidean geometry and arithmetic means.
With a different distance, a different notion of center is needed.}

\newthought{This chapter marks a conceptual turning point} in the course.
In Chapter~1, the question was local and pairwise: given two points, how far apart are they?
Here the question is global and combinatorial: given $n$ points, how do we partition them?
The answer K-Means gives, a hard partition into $K$ spherical clusters, comes with two assumptions that later chapters will systematically relax.
Hierarchical clustering will remove the requirement to fix $K$ in advance, revealing nested structure through a dendrogram.
Density-based methods will remove the sphericity assumption, finding clusters of arbitrary shape.

\newthought{Working through centroid clustering} requires us to,
\begin{itemize}
    \item define an objective that formalizes what a good partition means.
    \item understand how the assignment-update cycle minimizes that objective and when it stops.
    \item recognize the geometric assumptions behind K-Means and the failure modes they produce.
\end{itemize}

\newthought{The fundamental question} this chapter answers:
given a distance measure, how do we find the clusters?

\subsection{Hands On Experience}

\newthought{Consider} twelve data points in two dimensions.
Which points belong together? How many groups do you see?

\begin{figure}
\centering
\begin{tikzpicture}[scale=1]
    \begin{axis}[
        xmin=0, xmax=10,
        ymin=0, ymax=10,
        width=0.5*\textwidth,
        height=0.5*\textwidth,
        axis line style={draw=none},
        tick style={black, thick},
        xtick={1,2,3,4,5,6,7,8,9},
        ytick={1,2,3,4,5,6,7,8,9},
        xticklabel style={font=\small},
        yticklabel style={font=\small},
        tick align=outside,
        tick pos=left,
    ]
    \addplot[only marks, mark=*, mark size=2pt, black] coordinates {
        (1,2) (2,1) (2,3) (3,2)
        (4,5) (5,4) (5,6) (6,5)
        (7,8) (8,7) (8,9) (9,8)
    };
    \end{axis}
\end{tikzpicture}
\caption{Twelve points in 2D. How many groups do you see?}
\label{fig:intro_points}
\end{figure}

\newthought{How many groups} do you see? Three seems natural, but could you argue for two?
For six?

\newthought{Where would you place} a representative point for each group?
Notice that you already have an intuition for what that point should be.

\newthought{If a new point} arrives at $(3.5, 3.5)$, which group does it join?
How much does it change your picture of the groups?

\newthought{The Learning Objectives} of this lecture:
\begin{itemize}
    \item Understand the K-Means objective and algorithm, including convergence.
    \item Apply initialization strategies and K-selection methods.
    \item Identify when K-Means fails and understand why.
\end{itemize}


% ==============================================================================================
% BLOCK 2: MAIN THEORY
% ==============================================================================================

\newpage

\section{The K-Means Algorithm}\index{K-Means}\index{inertia}

\newthought{A data point}\index{data point} is a vector $x_i \in \mathbb{R}^d$, where $d$ is the number of features and $i = 1, \ldots, n$ indexes the observations.

\newthought{A cluster}\index{cluster} is a non-empty subset $C_k \subseteq \{x_1, \ldots, x_n\}$.
A \textbf{clustering}\index{clustering} of the data is a collection of $K$ clusters $\{C_1, \ldots, C_K\}$ that are disjoint and exhaustive:

\begin{align*}
    C_j \cap C_k &= \emptyset \quad \text{for } j \neq k \\
    \bigcup_{k=1}^{K} C_k &= \{x_1, \ldots, x_n\}
\end{align*}

\marginnote{\newthought{Partition} is the mathematical synonym for clustering. A partition of a set is any division into non-overlapping, exhaustive subsets — every element in exactly one group, no element left out. Think of $K{=}3$: you get three clusters (one group each) and one clustering (all three together). The word partition emphasises the structural guarantee; clustering emphasises the process.

\newthought{Each point belongs to exactly one cluster.} This is the hard assignment assumption. Soft variants (e.g.\ Gaussian mixture models) assign fractional membership probabilities instead.}

\vspace{2.5em}
\newthought{K-Means} finds a clustering of $n$ data points into $K$ clusters by minimizing the total squared distance from each point to its cluster's centroid.
This quantity is called \textbf{inertia} or within-cluster sum of squares (WCSS):

\begin{align}
    J &= \sum_{k=1}^{K} \sum_{x_i \in C_k} \|x_i - \mu_k\|^2
    \label{eq:inertia}
\end{align}

where $\mu_k = \frac{1}{|C_k|} \sum_{x_i \in C_k} x_i$ is the centroid of cluster $C_k$.

\marginnote{\newthought{Minimizing $J$} is NP-hard in general. K-Means finds a local optimum, not necessarily the global one. The result depends on initialization.}

\newthought{The algorithm alternates between two steps} until the assignments no longer change:

\begin{align*}
    \text{Assignment:} \quad c_i &\leftarrow \arg\min_k \|x_i - \mu_k\|^2 \quad \text{for each } x_i \\[6pt]
    \text{Update:} \quad \mu_k &\leftarrow \frac{1}{|C_k|} \sum_{x_i:\, c_i = k} x_i \quad \text{for each } k
\end{align*}

The assignment step asks: which centroid is closest?
The update step asks: given the current assignments, where is the best centroid?

\newthought{A trace through the algorithm} on six coffee samples makes this concrete.
With $K{=}3$ and initial centroids at $\mu_1{=}(1,2)$, $\mu_2{=}(4,5)$, $\mu_3{=}(8,9)$:

\marginnote{Six exercise samples:\\
$A=(1,2)$, $B=(2,1)$\\
$C=(4,5)$, $D=(5,4)$\\
$E=(8,9)$, $F=(9,8)$\\[4pt]
acidity, bitterness (1--10)}

\begin{figure}
\centering
\begin{tikzpicture}[scale=1]
    % Panel 1: Initialization
    \begin{axis}[
        name=init,
        title={Initialization},
        xlabel={acidity}, ylabel={bitterness},
        xmin=0, xmax=10, ymin=0, ymax=10,
        width=0.32\textwidth, height=0.32\textwidth,
        axis line style={draw=none},
        tick style={black, thin},
        xtick={2,5,8}, ytick={2,5,8},
        xticklabel style={font=\tiny},
        yticklabel style={font=\tiny},
        tick align=outside, tick pos=left,
        title style={font=\small},
    ]
    \addplot[only marks, mark=*, mark size=2pt, black!30] coordinates {
        (1,2)(2,1)(4,5)(5,4)(8,9)(9,8)
    };
    \addplot[only marks, mark=x, mark size=6pt, black, thick] coordinates {
        (1,2)(4,5)(8,9)
    };
    \end{axis}

    % Panel 2: After assignment
    \begin{axis}[
        name=assign,
        at={(init.east)}, anchor=west, xshift=0.4cm,
        title={After assignment},
        xlabel={acidity},
        xmin=0, xmax=10, ymin=0, ymax=10,
        width=0.32\textwidth, height=0.32\textwidth,
        axis line style={draw=none},
        tick style={black, thin},
        xtick={2,5,8}, ytick={2,5,8},
        xticklabel style={font=\tiny},
        yticklabel style={font=\tiny},
        tick align=outside, tick pos=left,
        title style={font=\small},
    ]
    \addplot[only marks, mark=*, mark size=2pt, black] coordinates {(1,2)(2,1)};
    \addplot[only marks, mark=*, mark size=2pt, black!55] coordinates {(4,5)(5,4)};
    \addplot[only marks, mark=*, mark size=2pt, black!20] coordinates {(8,9)(9,8)};
    \addplot[only marks, mark=x, mark size=6pt, black, thick] coordinates {(1,2)(4,5)(8,9)};
    \end{axis}

    % Panel 3: After update
    \begin{axis}[
        name=update,
        at={(assign.east)}, anchor=west, xshift=0.4cm,
        title={After update},
        xlabel={acidity},
        xmin=0, xmax=10, ymin=0, ymax=10,
        width=0.32\textwidth, height=0.32\textwidth,
        axis line style={draw=none},
        tick style={black, thin},
        xtick={2,5,8}, ytick={2,5,8},
        xticklabel style={font=\tiny},
        yticklabel style={font=\tiny},
        tick align=outside, tick pos=left,
        title style={font=\small},
    ]
    \addplot[only marks, mark=*, mark size=2pt, black] coordinates {(1,2)(2,1)};
    \addplot[only marks, mark=*, mark size=2pt, black!55] coordinates {(4,5)(5,4)};
    \addplot[only marks, mark=*, mark size=2pt, black!20] coordinates {(8,9)(9,8)};
    \addplot[only marks, mark=x, mark size=6pt, black, thick] coordinates {(1.5,1.5)(4.5,4.5)(8.5,8.5)};
    \end{axis}
\end{tikzpicture}
\caption{One K-Means iteration. Centroids (crosses) are initialized on data points, assignments follow, then centroids move to cluster means. Assignments do not change in the next iteration: converged.}
\label{fig:kmeans_trace}
\end{figure}

\newthought{Convergence} is guaranteed in finite steps.
The argument rests on three observations:
first, each assignment step can only decrease $J$, since every point moves to an equal or closer centroid;
second, each update step can only decrease $J$, since the mean minimizes squared distance to its members;
third, there are at most $K^n$ possible assignment configurations.
Since $J$ is non-increasing and the number of states is finite, the algorithm must reach a fixed assignment and stop.

\marginnote{\newthought{Local optimum.} Convergence does not mean the algorithm found the best clustering. It found \textit{a} clustering. A different initialization may yield a better one.}

Typical convergence takes 10--100 iterations in practice.



\vspace{2.5em}
\newthought{Initialization Matters}\index{K-Means++}

K-Means is sensitive to where centroids start.
If two initial centroids fall inside the same natural cluster, one cluster gets split and another gets merged, and the algorithm may never recover.

\begin{figure}
\centering
\begin{tikzpicture}[scale=1]
    % Good initialization result
    \begin{axis}[
        name=good,
        title={Good initialization: $J=3$},
        xlabel={acidity}, ylabel={bitterness},
        xmin=0, xmax=10, ymin=0, ymax=10,
        width=0.45\textwidth, height=0.45\textwidth,
        axis line style={draw=none},
        tick style={black, thin},
        xtick={2,5,8}, ytick={2,5,8},
        xticklabel style={font=\small},
        yticklabel style={font=\small},
        tick align=outside, tick pos=left,
        title style={font=\small},
    ]
    \addplot[only marks, mark=*, mark size=2pt, black] coordinates {(1,2)(2,1)};
    \addplot[only marks, mark=*, mark size=2pt, black!55] coordinates {(4,5)(5,4)};
    \addplot[only marks, mark=*, mark size=2pt, black!20] coordinates {(8,9)(9,8)};
    \addplot[only marks, mark=x, mark size=6pt, black, thick] coordinates {(1.5,1.5)(4.5,4.5)(8.5,8.5)};
    \end{axis}

    % Bad initialization result
    \begin{axis}[
        name=bad,
        at={(good.east)}, anchor=west, xshift=0.8cm,
        title={Bad initialization: $J=34$},
        xlabel={acidity},
        xmin=0, xmax=10, ymin=0, ymax=10,
        width=0.45\textwidth, height=0.45\textwidth,
        axis line style={draw=none},
        tick style={black, thin},
        xtick={2,5,8}, ytick={2,5,8},
        xticklabel style={font=\small},
        yticklabel style={font=\small},
        tick align=outside, tick pos=left,
        title style={font=\small},
    ]
    % A is its own singleton cluster
    \addplot[only marks, mark=*, mark size=2pt, black] coordinates {(1,2)};
    % B is its own singleton cluster
    \addplot[only marks, mark=*, mark size=2pt, black!55] coordinates {(2,1)};
    % C, D, E, F all merged into one cluster
    \addplot[only marks, mark=*, mark size=2pt, black!20] coordinates {(4,5)(5,4)(8,9)(9,8)};
    \addplot[only marks, mark=x, mark size=6pt, black, thick] coordinates {(1,2)(2,1)(6.5,6.5)};
    \end{axis}
\end{tikzpicture}
\caption{Same data, different initial centroids. Left: one centroid per natural cluster gives $J{=}3$. Right: two centroids inside the cold-brew cluster leave the drip and espresso groups merged, giving $J{=}34$.}
\label{fig:kmeans_init}
\end{figure}

\marginnote{\newthought{Multiple restarts.} A practical fix: run K-Means $n$ times with different random initializations, keep the result with the lowest $J$. Typical default: $n{=}10$.}

\newthought{K-Means++}\cite{ArthurVassilvitskii2007} gives a smarter starting point.
The first centroid is chosen uniformly at random.
Each subsequent centroid is chosen with probability proportional to its squared distance to the nearest already-chosen centroid:

\begin{align*}
    p(x_i) &= \frac{D(x_i)^2}{\sum_j D(x_j)^2}
\end{align*}

where $D(x_i) = \min_{j < k} \|x_i - \mu_j\|$.
Points far from all current centroids are more likely to be chosen next, spreading the initial centroids across the data.
This gives a theoretical guarantee on the expected inertia relative to the global optimum.



\vspace{2.5em}
\newthought{Choosing K}\index{elbow method}\index{silhouette score}

K must be set by the practitioner.
Two complementary heuristics help.

\newthought{The elbow method} plots inertia $J$ against $K$ and looks for a kink: the point where adding one more cluster yields diminishing returns.
For the coffee data, $J$ drops sharply from $K{=}1$ to $K{=}3$, then flattens.

\begin{figure}
\centering
\begin{tikzpicture}[scale=1]
    \begin{axis}[
        xlabel={$K$},
        ylabel={inertia $J$},
        xmin=0.5, xmax=6.5,
        ymin=0, ymax=115,
        width=0.5*\textwidth,
        height=0.4*\textwidth,
        axis line style={draw=none},
        tick style={black, thick},
        xtick={1,2,3,4,5,6},
        ytick={0,20,40,60,80,100},
        xticklabel style={font=\small},
        yticklabel style={font=\small},
        tick align=outside,
        tick pos=left,
    ]
    \addplot[black, thick, mark=*, mark size=2pt] coordinates {
        (1,102) (2,21) (3,3) (4,2) (5,1) (6,0)
    };
    \draw[black!40, dashed] (axis cs:3,3) -- (axis cs:3,0);
    \node[font=\small] at (axis cs:3.6,15) {elbow at $K{=}3$};
    \end{axis}
\end{tikzpicture}
\caption{Elbow plot for the coffee data. The kink at $K{=}3$ matches the three natural groups.}
\label{fig:elbow}
\end{figure}

\marginnote{\newthought{The elbow is not always clear.} On real data, the curve may decrease smoothly with no obvious kink. The elbow is a heuristic, not a guarantee.}

\newthought{The silhouette score}\index{silhouette score} measures how well each point fits its assigned cluster relative to the next-best cluster.
For point $x_i$, let $a(i)$ be its mean distance to points in the same cluster and $b(i)$ its mean distance to points in the nearest other cluster:

\begin{align*}
    s(i) &= \frac{b(i) - a(i)}{\max\bigl(a(i),\, b(i)\bigr)}
\end{align*}

$s(i)$ ranges from $-1$ (point would fit better elsewhere) to $+1$ (point is well-placed).
The mean silhouette over all points is a quality score for the full clustering.
Choose $K$ that maximizes it.

\marginnote{\newthought{For our coffee data} with $K{=}3$, every point has $a(i) {=} \sqrt{2} \approx 1.4$ and $b(i) \approx 4.4$, giving $s(i) \approx 0.68$. A clear, well-separated clustering.}



\vspace{2.5em}
\newthought{When K-Means Fails}\index{failure modes}

K-Means assumes clusters are roughly spherical, equally sized, and equally dense.
When the data violates these assumptions, the Voronoi partition imposed by Euclidean centroids cuts the data in the wrong places.

\begin{figure}
\centering
\begin{tikzpicture}[scale=1]
    % Panel 1: Non-convex (rings)
    \begin{axis}[
        name=rings,
        title={Non-convex},
        xlabel={$x_1$}, ylabel={$x_2$},
        xmin=-3.5, xmax=3.5,
        ymin=-3.5, ymax=3.5,
        width=0.32\textwidth, height=0.32\textwidth,
        axis line style={draw=none},
        tick style={draw=none},
        xtick=\empty, ytick=\empty,
        title style={font=\small},
        axis equal,
    ]
    \addplot[only marks, mark=*, mark size=1.5pt, black,
        domain=0:360, samples=30] ({0.9*cos(x)}, {0.9*sin(x)});
    \addplot[only marks, mark=*, mark size=1.5pt, black!35,
        domain=0:360, samples=45] ({2.5*cos(x)}, {2.5*sin(x)});
    \end{axis}

    % Panel 2: Different sizes
    \begin{axis}[
        name=sizes,
        at={(rings.east)}, anchor=west, xshift=0.4cm,
        title={Different sizes},
        xlabel={$x_1$},
        xmin=-2, xmax=12,
        ymin=-3, ymax=3,
        width=0.32\textwidth, height=0.32\textwidth,
        axis line style={draw=none},
        tick style={draw=none},
        xtick=\empty, ytick=\empty,
        title style={font=\small},
    ]
    \addplot[only marks, mark=*, mark size=1.5pt, black] coordinates {
        (0,0)(0.2,0.3)(-0.1,0.2)(0.3,-0.1)(-0.2,-0.2)(0.1,0.4)
    };
    \addplot[only marks, mark=*, mark size=1.5pt, black!35] coordinates {
        (6,0)(7.5,1)(8,0)(6,-1.5)(7,-2)(9,1)(5,0)(8,-1)(6,2)(9,-1)(7,2)(5,-1)
    };
    \end{axis}

    % Panel 3: Different densities
    \begin{axis}[
        name=densities,
        at={(sizes.east)}, anchor=west, xshift=0.4cm,
        title={Different densities},
        xlabel={$x_1$},
        xmin=-2, xmax=12,
        ymin=-2, ymax=4,
        width=0.32\textwidth, height=0.32\textwidth,
        axis line style={draw=none},
        tick style={draw=none},
        xtick=\empty, ytick=\empty,
        title style={font=\small},
    ]
    \addplot[only marks, mark=*, mark size=1.5pt, black] coordinates {
        (0,1)(0.1,0.9)(0,1.1)(0.05,0.95)(-0.05,1.05)(0.1,1.1)(-0.1,0.9)
    };
    \addplot[only marks, mark=*, mark size=1.5pt, black!35] coordinates {
        (5,0)(8,3)(6,0)(9,1)(5,2)(8,0)(7,3)(6,2)
    };
    \end{axis}
\end{tikzpicture}
\caption{Three scenarios where K-Means fails. Left: concentric rings share a centroid and cannot be separated. Centre: a large and a small cluster pull the boundary toward the small one. Right: a dense and a sparse cluster of equal inertia look equivalent to K-Means.}
\label{fig:kmeans_failures}
\end{figure}

\marginnote{\newthought{K-Medoids} uses actual data points as cluster centers instead of means, making it robust to outliers and compatible with non-Euclidean distances. \newthought{Mini-Batch K-Means} updates centroids on random subsets of the data, scaling to millions of points at the cost of a slightly noisier result.}

\newthought{When K-Means struggles}, the underlying issue is always the same: the Euclidean distance to a centroid is the wrong measure of belonging for that data.
The next chapters address this by changing the notion of neighborhood (hierarchical clustering) and of density (DBSCAN).



% ==============================================================================================
% BLOCK 3: STUDENT ACTIVATION
% ==============================================================================================

\newpage

\section{Examples \& Exercises}\index{examples}\index{exercises}

\marginnote{\newthought{Exercises} are for practice and reinforcing concepts. Try to solve them on your own first,
try things, play with it, discuss; this is not a time trial. And there is no shame in not ending up at the right answer,
in the same sense that uncovering great questions
and tossing them around is usually pretty fruitful in the long run.}

\newthought{Computing by hand} builds the intuition that no amount of library calls can replace.
Step through the arithmetic slowly.

\vspace{1em}

\marginnote{$A=(1,2)$, $B=(2,1)$\\$C=(4,5)$, $D=(5,4)$\\$E=(8,9)$, $F=(9,8)$\\[4pt]acidity, bitterness (1--10)}
\newthought{Given six coffee samples} $A$--$F$, run K-Means with $K{=}3$ and initial centroids $\mu_1{=}A$, $\mu_2{=}C$, $\mu_3{=}E$.
Perform the assignment step, the update step, and compute the inertia after convergence.

\vspace{0.5em}

The assignment step computes the squared distance from each point to each centroid and assigns each point to the nearest one.
The distances are symmetric by construction so we only check whether a point is closest to its natural cluster center:

\begin{align*}
    d(A,\mu_1)^2 &= 0, \quad d(A,\mu_2)^2 = 18, \quad d(A,\mu_3)^2 = 98 \;\to\; C_1 \\
    d(B,\mu_1)^2 &= 2, \quad d(B,\mu_2)^2 = 20, \quad d(B,\mu_3)^2 = 100 \;\to\; C_1 \\
    d(C,\mu_1)^2 &= 18, \quad d(C,\mu_2)^2 = 0, \quad d(C,\mu_3)^2 = 32 \;\to\; C_2 \\
    d(D,\mu_1)^2 &= 20, \quad d(D,\mu_2)^2 = 2, \quad d(D,\mu_3)^2 = 34 \;\to\; C_2 \\
    d(E,\mu_1)^2 &= 98, \quad d(E,\mu_2)^2 = 32, \quad d(E,\mu_3)^2 = 0 \;\to\; C_3 \\
    d(F,\mu_1)^2 &= 100, \quad d(F,\mu_2)^2 = 34, \quad d(F,\mu_3)^2 = 2 \;\to\; C_3
\end{align*}

The update step recomputes each centroid as the mean of its assigned points:

\begin{align*}
    \mu_1' &= \tfrac{1}{2}\bigl((1,2)+(2,1)\bigr) = (1.5,\;1.5) \\
    \mu_2' &= \tfrac{1}{2}\bigl((4,5)+(5,4)\bigr) = (4.5,\;4.5) \\
    \mu_3' &= \tfrac{1}{2}\bigl((8,9)+(9,8)\bigr) = (8.5,\;8.5)
\end{align*}

Running the assignment step again with the new centroids yields the same clusters; the algorithm has converged.
The inertia is:

\begin{align*}
    J &= \|A - \mu_1'\|^2 + \|B - \mu_1'\|^2
       + \|C - \mu_2'\|^2 + \|D - \mu_2'\|^2
       + \|E - \mu_3'\|^2 + \|F - \mu_3'\|^2 \\
      &= 0.5 + 0.5 + 0.5 + 0.5 + 0.5 + 0.5 \\
      &= 3
\end{align*}

Each pair contributes equally; the clustering is geometrically symmetric.


\vspace{2.5em}
\newthought{Now repeat} with bad initial centroids $\mu_1{=}A$, $\mu_2{=}B$, $\mu_3{=}C$ — two centroids inside the cold-brew cluster.

\marginnote{\newthought{What goes wrong.} Both $\mu_1$ and $\mu_2$ compete for cold-brew samples. The algorithm optimizes what it is given; it has no way of knowing that $A$ and $B$ are in the same natural group.}

Assignment step with $\mu_1{=}(1,2)$, $\mu_2{=}(2,1)$, $\mu_3{=}(4,5)$:

\begin{align*}
    A &\to C_1 \quad (\text{distance 0 to } \mu_1) \\
    B &\to C_2 \quad (\text{distance 0 to } \mu_2) \\
    C,\,D,\,E,\,F &\to C_3 \quad (\text{all closer to } \mu_3 \text{ than to } \mu_1 \text{ or } \mu_2)
\end{align*}

After the update: $\mu_1{=}(1,2)$ (singleton), $\mu_2{=}(2,1)$ (singleton), $\mu_3{=}(6.5,6.5)$ (four points merged).
The assignments do not change in the next iteration.
The inertia of this converged solution is:

\begin{align*}
    J_{\text{bad}} &= 0 + 0 + \|C - \mu_3\|^2 + \|D - \mu_3\|^2 + \|E - \mu_3\|^2 + \|F - \mu_3\|^2 \\
                   &= 0 + 0 + 8.5 + 8.5 + 8.5 + 8.5 \;=\; 34
\end{align*}

The bad initialization found a solution more than ten times worse than the good one ($J{=}34$ vs $J{=}3$).
This is why K-Means++ and multiple restarts matter.


\vspace{2.5em}
\newthought{For the elbow method}, run K-Means on the same six points for $K = 1, \ldots, 6$ and record the inertia.

\marginnote{Inertia values:\\[4pt]
\begin{tabular}{cc}
$K$ & $J$ \\
\hline
1 & 102 \\
2 & 21 \\
3 & 3 \\
4 & 2 \\
5 & 1 \\
6 & 0 \\
\end{tabular}}

The drop from $K{=}1$ to $K{=}3$ is large (102 to 3); from $K{=}3$ onward the curve is nearly flat.
The elbow is at $K{=}3$, matching the three natural coffee varieties.
Note that $K{=}6$ always achieves $J{=}0$ by placing one centroid per point; this is not useful.



% ============================= CODE ===============================
\marginnote{\newthought{Code} will be provided as a Python notebook. Use it as a starting point, break things, and observe what changes.}

\marginnote{$n=20$ samples\\
\textit{acidity} (1--10)\\
\textit{bitterness} (1--10)\\[4pt]
Varieties: cold brew, drip, espresso, latte.}
\newthought{A coffee dataset to explore.} Load the dataset and scatter-plot acidity vs.\ bitterness.
Run K-Means with $K{=}3$ on raw features and visualize the cluster assignments.
Try different random seeds; how stable are the results?
Then vary $K$ from 1 to 8, record the inertia and silhouette score for each, and plot both curves.
Finally, deliberately give K-Means a bad initialization and compare the resulting inertia with the K-Means++ result.

\newthought{What to observe.} With a good initialization and $K{=}3$, the three clear clusters emerge immediately.
The elbow and the silhouette peak agree at $K{=}3$.
With a bad initialization, inertia is higher and cluster boundaries shift; the silhouette score drops.
With $K{=}4$ a natural cluster splits in two, and the silhouette score begins to fall.



\section{Self-Reflection and Recap}\index{reflection}\index{recap}

\newthought{Self-Reflection} questions to guide your thinking:
\begin{itemize}
    \item What does the K-Means objective actually minimize, and why is that a reasonable definition of a good clustering?
    \item Why does K-Means always converge, and why does convergence not guarantee the best solution?
    \item What does K-Means++ change about the initialization, and why does spreading centroids out help?
    \item When would you trust the elbow method? When would you not?
    \item How does the silhouette score measure something different from inertia?
    \item What shape of cluster cannot K-Means find, and why?
    \item How does the choice of distance metric interact with the K-Means update step?
    \item Given a new dataset, what would be your first three steps before running K-Means?
\end{itemize}

\newthought{Recap} of Key Concepts:
\begin{itemize}
    \item K-Means minimizes inertia by alternating assignment (points to nearest centroid) and update (centroid to cluster mean)
    \item Convergence is guaranteed; optimality is not, use K-Means++ or multiple restarts
    \item The elbow and silhouette score are complementary heuristics for choosing $K$
    \item K-Means assumes spherical, equally sized clusters; it fails on rings, elongated shapes, and density differences
\end{itemize}

\marginnote{\newthought{Teaser.} K-Means assigns every point to exactly one flat cluster. What if structure exists at multiple scales, or if some points genuinely belong between groups?}
\newthought{So far, we found flat partitions} of the data into $K$ fixed groups.
But natural structure is often hierarchical: coffee varieties cluster by roast, roasts cluster by origin, origins cluster by continent.
And some points genuinely sit at the boundary of two groups, which a hard assignment hides.
The next chapters address both: hierarchical clustering reveals nested structure through dendrograms, and density-based methods find clusters of arbitrary shape without requiring a fixed $K$.
