% !TEX root = ../notes_unsupervisedlearning.tex
% Chapter 4: Density-Based Clustering
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Density-Based Clustering}\index{DBSCAN}\index{density-based clustering}
\newthought{It is all about that space, about that space.}

% ==============================================================================================
% BLOCK 1: INTRODUCTION
% ==============================================================================================

\newthought{The Why}\index{density-based clustering}

The clustering algorithms we have seen so far share a common mechanism of assigning points to clusters based on their distance 
to an other point; a centroid or a linkage point.
Every point, no matter how isolated, is absorbed into some cluster latest at the final merge.
Real data contains noise: GPS measurements with dropouts, sensor readings during interference, 
survey responses that are genuine outliers.

\begin{marginfigure}
\centering
\begin{tikzpicture}
    \begin{axis}[
        xmin=0, xmax=10, ymin=0, ymax=10,
        width=\marginparwidth, height=\marginparwidth,
        axis line style={draw=none},
        tick style={black, thin},
        xtick={2,4,6,8}, ytick={2,4,6,8},
        xticklabel style={font=\tiny},
        yticklabel style={font=\tiny},
        tick align=outside, tick pos=left,
    ]
        % Dense cluster: each point carries its own epsilon-neighbourhood; they overlap
        \draw[black!25, thin, dashed] (axis cs:2,2)     circle [radius=1.0];
        \draw[black!25, thin, dashed] (axis cs:1.5,2.8) circle [radius=1.0];
        \draw[black!25, thin, dashed] (axis cs:2.8,1.6) circle [radius=1.0];
        \draw[black!25, thin, dashed] (axis cs:1.8,1.3) circle [radius=1.0];
        \addplot[only marks, mark=*, mark size=1.5pt, black] coordinates {(2,2)(1.5,2.8)(2.8,1.6)(1.8,1.3)};
        \node[font=\tiny, black!55, anchor=south west, inner sep=1pt] at (axis cs:2.05,2.05)   {4};
        \node[font=\tiny, black!55, anchor=south west, inner sep=1pt] at (axis cs:1.55,2.85)   {2};
        \node[font=\tiny, black!55, anchor=south west, inner sep=1pt] at (axis cs:2.85,1.65)   {2};
        \node[font=\tiny, black!55, anchor=south west, inner sep=1pt] at (axis cs:1.85,1.35)   {2};
        % Isolated p_n: its epsilon-neighbourhood contains no one else
        \draw[black!25, thin, dashed] (axis cs:6.5,6.5) circle [radius=1.0];
        \addplot[only marks, mark=*, mark size=1.5pt, black!55] coordinates {(6.5,6.5)};
        \node[font=\tiny, black!55, anchor=south west, inner sep=1pt] at (axis cs:6.55,6.55)   {1};
        \end{axis}
    \end{tikzpicture}
    \caption{Each point carries its own dashed $\varepsilon$-neighbourhood; the small grey digit next to a point gives the number of points inside that circle, itself included. In the dense cluster at the lower left the four neighbourhoods overlap and chain the points together; the isolated $p_n$ at the upper right has no companions within $\varepsilon$. Density, not distance to a centre, determines membership.}
\end{marginfigure}

\newthought{No centroid required.} A density-based cluster is a space, not a mean and not a link.
Its shape is implicit in the data\index{cluster shape}\index{non-convex clusters}, seeded by the points themselves; nothing forces it to be convex or spherical\index{geometric assumptions};
detects noise\index{noise}, outliers\index{outliers} and sparse bridges naturally, and avoids committing to $K$ in advance. 
This geometric-assumption-free view is both the strength of density-based methods and, as we will see, 
the source of their own limitations.


\newthought{In order to move from pairwise distances to density-aware neighbourhoods} we have good reason to find ways to,
\begin{itemize}
    \item define what it means for a region of space to be dense or sparse.
    \item grow clusters of arbitrary shape, including non-convex shapes, from local density, without assuming convex or spherical structure.
    \item separate signal from noise, so that sparse outliers are labelled as such rather than forced into a cluster.
    \item let the data itself decide how many clusters exist, without committing to $K$ in advance.
\end{itemize}










\newpage
\subsection{Hands On Experience}

\newthought{Consider} the same six canonical points, now with a seventh point $x_7{=}(4,5)$ placed between the two natural groups.

\begin{figure}
\centering
\begin{tikzpicture}[scale=1]
    \begin{axis}[
        xmin=0, xmax=10,
        ymin=0, ymax=10,
        width=0.5\textwidth,
        height=0.5\textwidth,
        axis line style={draw=none},
        tick style={black, thick},
        xtick={1,2,3,4,5,6,7,8,9},
        ytick={1,2,3,4,5,6,7,8,9},
        xticklabel style={font=\small},
        yticklabel style={font=\small},
        tick align=outside,
        tick pos=left,
    ]
    \addplot[only marks, mark=*, mark size=2pt, black] coordinates {
        (1,2) (2,3) (3,1)
    };
    \addplot[only marks, mark=*, mark size=2pt, black!55] coordinates {
        (6,8) (8,7) (8,9)
    };
    \addplot[only marks, mark=*, mark size=2pt, black!30] coordinates {
        (4,5)
    };
    \node[font=\small, anchor=south east] at (axis cs:1,2) {$x_1$};
    \node[font=\small, anchor=south east] at (axis cs:2,3) {$x_2$};
    \node[font=\small, anchor=north]      at (axis cs:3,1) {$x_3$};
    \node[font=\small, anchor=south west] at (axis cs:6,8) {$x_4$};
    \node[font=\small, anchor=north west] at (axis cs:8,7) {$x_5$};
    \node[font=\small, anchor=south west] at (axis cs:8,9) {$x_6$};
    \node[font=\small, anchor=south west] at (axis cs:4,5) {$x_7$};
    \end{axis}
\end{tikzpicture}
\caption{Six canonical points plus $x_7{=}(4,5)$ sitting between the two natural groups. Which points lie in dense neighbourhoods, and which sits in a sparse region?}
\label{fig:dbscan_intro_scatter}
\end{figure}

\marginnote{\newthought{The neighbourhood radius $\varepsilon$} is the distance within which a point looks for companions.
Too small and every point is left alone; too large and everything collapses into one cluster.
The right value lives where the data transitions from dense to sparse, and we will read it off the data itself 
using the $k$-distance plot.}

\newthought{Draw a circle} of radius $\varepsilon{=}2.5$ around each point and count who sits inside.
The three bottom-left points find each other, the three top-right points find each other, and $x_7$ finds empty space; 
read off the data without ever choosing a number of clusters.

\newthought{Our approaches so far would refuse} to let $x_7$ be lonely.
With $K{=}2$ it hands the point to whichever centroid happens to sit slightly closer, 
an assignment driven by initialisation rather than by any structure in the data.
Hierarchical clustering would fold $x_7$ into some merge and bury it inside a cluster it never really joined.

\newthought{Density allows us to do so.}
A point whose $\varepsilon$-neighbourhood is empty is not handed to the nearest cluster; 
it is labelled $p_n$.
Cluster shape is read off local neighbourhoods with a sense for $p_b$'s, the number of clusters emerges 
from where density breaks instead of from a chosen $K$, and sparse points are flagged instead of absorbed.


\begin{marginfigure}
\centering
\begin{tikzpicture}
    \begin{axis}[
        xmin=0, xmax=10, ymin=0, ymax=10,
        width=\marginparwidth, height=\marginparwidth,
        axis line style={draw=none},
        tick style={black, thin},
        xtick={2,4,6,8}, ytick={2,4,6,8},
        xticklabel style={font=\tiny},
        yticklabel style={font=\tiny},
        tick align=outside, tick pos=left,
    ]
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({1 + 2.5*cos(x)}, {2 + 2.5*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({2 + 2.5*cos(x)}, {3 + 2.5*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({3 + 2.5*cos(x)}, {1 + 2.5*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({6 + 2.5*cos(x)}, {8 + 2.5*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({8 + 2.5*cos(x)}, {7 + 2.5*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({8 + 2.5*cos(x)}, {9 + 2.5*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({4 + 2.5*cos(x)}, {5 + 2.5*sin(x)});
    \addplot[only marks, mark=*, mark size=1.5pt, black]    coordinates {(1,2)(2,3)(3,1)};
    \addplot[only marks, mark=*, mark size=1.5pt, black!55] coordinates {(6,8)(8,7)(8,9)};
    \addplot[only marks, mark=*, mark size=1.5pt, black!30] coordinates {(4,5)};
    \node[font=\tiny, black!55, anchor=south west, inner sep=1pt] at (axis cs:1.15,2.15) {3};
    \node[font=\tiny, black!55, anchor=south west, inner sep=1pt] at (axis cs:2.15,3.15) {3};
    \node[font=\tiny, black!55, anchor=south west, inner sep=1pt] at (axis cs:3.15,1.15) {3};
    \node[font=\tiny, black!55, anchor=south west, inner sep=1pt] at (axis cs:6.15,8.15) {3};
    \node[font=\tiny, black!55, anchor=south west, inner sep=1pt] at (axis cs:8.15,7.15) {3};
    \node[font=\tiny, black!55, anchor=south west, inner sep=1pt] at (axis cs:8.15,9.15) {3};
    \node[font=\tiny, black!55, anchor=south west, inner sep=1pt] at (axis cs:4.15,5.15) {1};
    \end{axis}
\end{tikzpicture}
\caption{The seven canonical points with their dashed $\varepsilon{=}2.5$ neighbourhoods. The small grey digit next to each point gives $|N_\varepsilon|$, itself included. Each natural group forms a three-point neighbourhood; $x_7$ sits in a sparse region and finds only itself within $\varepsilon$.}
\label{fig:dbscan_intro_circles}
\end{marginfigure}

\newthought{The Learning Objectives} of this lecture:
\begin{itemize}
    \item Trace a density clustering by hand on a small dataset and learn how to set the neighbourhood radius $\varepsilon$, 
    and the minimum number of points $n_{\min}$ from the data itself.
    \item Understand the concepts of core ($p_c$), border ($p_b$), and noise ($p_n$) points by formalising neighbourhoods.
    \item Recognise when density-based clustering succeeds where K-Means and hierarchical fail, where they share characteristics, and where it fails in 
    turn under varying density, and in high dimensionality.
\end{itemize}








% ==============================================================================================
% BLOCK 2: MAIN THEORY
% ==============================================================================================

\newpage

\subsection{The Density-Based Algorithm}\index{DBSCAN}

\newthought{DB Scan}\cite{Ester1996}\index{DBSCAN} or Density-Based Spatial Clustering of Applications with Noise, 
builds clusters from local density rather than from a chosen number of centroids or a sequence of merges.
Two parameters govern the algorithm: $\varepsilon$, the neighbourhood radius, and $n_{\min}$, 
the minimum number of points required to form a dense region.

\vspace{1.5em}

\newthought{Given data} $X = \{x_1,\ldots,x_n\}$ and parameters $(\varepsilon,\,n_{\min})$, DB Scan returns a labelling
\begin{align*}
    \tilde{y}: X \;\longrightarrow\; \{1,\,2,\,\ldots,\,K,\;p_n\}
\end{align*}
where $K$ is determined by the data rather than fixed in advance. The output is a flat partition\index{partition} plus a $p_n$ set $N$.

\begin{marginfigure}[1em]
\centering
\begin{tikzpicture}
    \draw[black!40, dashed] (0,0) circle (1.0cm);
    \node[font=\tiny, black!40, anchor=south] at (0,1.05) {$\varepsilon$};
    \fill[black]    (0,0)         circle (2.5pt);
    \fill[black!55] ( 0.40, 0.30) circle (2pt);
    \fill[black!55] (-0.30, 0.35) circle (2pt);
    \fill[black!55] ( 0.50,-0.35) circle (2pt);
    \fill[black!55] (-0.35,-0.30) circle (2pt);
    \node[font=\tiny, black!55, anchor=south west, inner sep=1pt] at (0.05,0.05) {5};
\end{tikzpicture}
\vspace{1em}
\caption{\newthought{Core point.} The black point has $|N_\varepsilon|{=}5 \geq n_{\min}{=}4$. 
Four companions sit inside its dashed $\varepsilon$-neighbourhood. The small grey digit at the upper right gives $|N_\varepsilon|$, the point itself included.}
\label{fig:dbscan_core}
\end{marginfigure}



\newthought{Every point is classified} into one of three roles before any cluster is formed.
The $\varepsilon$-neighbourhood\index{epsilon-neighborhood} of a point $p$ is the set of all points within distance $\varepsilon$:
\begin{align*}
    N_\varepsilon(p) &= \{q \in X : d(p,\,q) \leq \varepsilon\}.
\end{align*}

\begin{itemize}
    \item A point $p_c$ is a core point\index{core point} if $|N_\varepsilon(p_c)| \geq n_{\min}$.
    \item A point $p_b$ is a border point\index{border point} if $|N_\varepsilon(p_b)| < n_{\min}$ but $p_b \in N_\varepsilon(p_c)$ for some core point $p_c$.
    \item A point $p_n$ is noise\index{noise point} if it is neither core nor border.
\end{itemize}

\begin{marginfigure}[1em]
\centering
\begin{tikzpicture}
    \draw[black!40, dashed] (0,0)   circle (1.0cm);
    \node[font=\tiny, black!40, anchor=south] at (0,1.05) {$\varepsilon$};
    \draw[black!40, dashed] (0.8,0) circle (1.0cm);
    \node[font=\tiny, black!40, anchor=south] at (0.8,1.05) {$\varepsilon$};
    \fill[black]    (0, 0)       circle (2.5pt);
    \fill[black!55] (-0.30, 0.35) circle (2pt);
    \fill[black!55] (-0.50,-0.25) circle (2pt);
    \fill[black!55] ( 0.30, 0.45) circle (2pt);
    \fill[black!30] (0.8, 0)     circle (2.5pt);
    \node[font=\tiny, black!55, anchor=south west, inner sep=1pt] at (0.05,0.05) {5};
    \node[font=\tiny, black!55, anchor=south west, inner sep=1pt] at (0.85,0.05) {3};
\end{tikzpicture}
\vspace{1em}
\caption{\newthought{Border point $p_b$.} The light grey point on the right has $|N_\varepsilon|{=}3 < n_{\min}{=}4$. It is not a $p_c$ itself, but it falls inside the $\varepsilon$-neighbourhood of the black $p_c$ on the left and is absorbed into that cluster.}
\label{fig:dbscan_border}
\end{marginfigure}




\begin{marginfigure}[1em]
\centering
\begin{tikzpicture}
    \draw[black!40, dashed] (0,0) circle (1.0cm);
    \node[font=\tiny, black!40, anchor=south] at (0,1.05) {$\varepsilon$};
    \fill[black!30] (0,0) circle (2.5pt);
    \node[font=\tiny, black!55, anchor=south west, inner sep=1pt] at (0.05,0.05) {1};
\end{tikzpicture}
\vspace{1em}
\caption{\newthought{Noise point $p_n$.} With $|N_\varepsilon|{=}1$ and no $p_c$ within $\varepsilon$, the point belongs to no cluster. The dashed circle is otherwise empty.}
\label{fig:dbscan_noise}
\end{marginfigure}


\vspace{1.5em}
\newthought{Density reachability} is not symmetric. A $p_b$ is reachable from a
$p_c$ whenever $p_b \in N_\varepsilon(p_c)$, but $p_c$ is not reachable from $p_b$ because $p_b$ is not a $p_c$. Cluster membership propagates from $p_c$'s outward, never in reverse.

\newthought{Density connectivity}\index{density connectivity} extends reachability to full cluster membership.
Point $q$ is density-reachable\index{density reachability} from $p$ if there exists a chain of points $p{=}p_1, p_2, \ldots, p_m{=}q$ such that
\begin{align*}
    p_{i+1} \in N_\varepsilon(p_i) \quad \text{and} \quad |N_\varepsilon(p_i)| \geq n_{\min} \quad \text{for } i=1,\ldots,m-1.
\end{align*}
Two points $p,\,q$ are density-connected\index{density connectivity} if there exists a third point $o_c$ from which both are density-reachable; a cluster is then defined as a maximal set of pairwise density-connected points.


\newpage
\newthought{Trace through our seven points.} Dashed circles are $\varepsilon{=}2.5$ neighbourhoods; 
dark points belong to $C_1$, grey to $C_2$, and an open circle marks $p_n$.

\newthought{Initialization}\index{initialization!DBSCAN}\index{determinism} is trivial as the set of clusters
and their core-memberships is the same regardless of the order in which points are processed.\index{convergence} 
Any unvisited $p_c$ triggers a new cluster; absorbing surrounding $p_b$'s immediately; and growing the cluster by absorbing 
other $p_c$'s and then their $p_b$'s. Only $p_b$'s that happen to lie in the $\varepsilon$-neighbourhood of several $p_c$'s 
can be assigned to different clusters depending on visit order, but keeping their point class $p_c$, $p_b$, or $p_n$.

\vspace{1.5em}
\begin{algorithm}[H]
\caption{Naive Density-Based Clustering (DBSCAN)}
\begin{algorithmic}[1]
\Require Data $X$, neighbourhood radius $\varepsilon$, density threshold $n_{\min}$
\Ensure Cluster label for each point (or $p_n$)
\State Mark all points as unvisited; $C \gets 0$
\For{each unvisited point $p_i \in X$}
    \State Mark $p_i$ as visited
    \State Compute $N \gets N_\varepsilon(p_i)$
    \If{$|N| < n_{\min}$}
        \State Label $p_i$ as $p_n$
    \Else
        \State Label $p_i$ as $p_c$; $C \gets C + 1$; assign $p_i$ to cluster $C$
        \For{each $p_j \in N \setminus \{p_i\}$}
            \If{$p_j$ is unvisited}
                \State Mark $p_j$ as visited
                \State Compute $N' \gets N_\varepsilon(p_j)$
                \If{$|N'| \geq n_{\min}$} label $p_j$ as $p_c$; $N \gets N \cup N'$ \EndIf
            \EndIf
            \If{$p_j$ has no cluster} assign $p_j$ to $C$ (as $p_c$ or $p_b$) \EndIf
        \EndFor
    \EndIf
\EndFor
\end{algorithmic}
\end{algorithm}
\vspace{1.5em}

\marginnote{\newthought{Complexity.}\index{computational complexity!DBSCAN} Each of the $n$ points triggers one $\varepsilon$-neighbourhood query costing $O(n)$, giving $O(n^2)$ overall. A spatial index (k-d or ball tree) reduces each query to $O(\log n)$, yielding $O(n \log n)$ in low dimensions.}

\newthought{Termination and determinism.} The outer loop visits every point at most once and the inner 
loop only grows by adding previously unvisited points, so the algorithm halts after $O(n)$ neighbourhood queries. 
The resulting cluster set is determined uniquely by the data and the pair $(\varepsilon,\,n_{\min})$: $p_c$'s 
are fixed by the $|N_\varepsilon|$ threshold.






%%%% TRACE OF DBSCAN ON THE SEVEN POINTS %%%%

\begin{figure}
\centering
\begin{tikzpicture}
    \begin{axis}[
        xlabel={$x^{(1)}$}, ylabel={$x^{(2)}$},
        xmin=0, xmax=10, ymin=0, ymax=10,
        width=0.5\textwidth, height=0.5\textwidth,
        axis line style={draw=none},
        tick style={black, thin},
        xtick={2,4,6,8}, ytick={2,4,6,8},
        xticklabel style={font=\small},
        yticklabel style={font=\small},
        tick align=outside, tick pos=left,
    ]
    \addplot[only marks, mark=*, mark size=2pt, black!30]
        coordinates {(1,2)(2,3)(3,1)(6,8)(8,7)(8,9)(4,5)};
    \node[font=\tiny, anchor=north west] at (axis cs:1,2) {$x_1$};
    \node[font=\tiny, anchor=south east] at (axis cs:2,3) {$x_2$};
    \node[font=\tiny, anchor=north]      at (axis cs:3,1) {$x_3$};
    \node[font=\tiny, anchor=south west] at (axis cs:6,8) {$x_4$};
    \node[font=\tiny, anchor=north west] at (axis cs:8,7) {$x_5$};
    \node[font=\tiny, anchor=south west] at (axis cs:8,9) {$x_6$};
    \node[font=\tiny, anchor=south west] at (axis cs:4,5) {$x_7$};
    \end{axis}
\end{tikzpicture}
\caption{\newthought{Initialization}: all $n{=}7$ points unvisited. Parameters $\varepsilon{=}2.5$, $n_{\min}{=}2$ are fixed. 
No seed necessary. Randomly picking $x_1$ as the first point to process.}
\label{fig:dbscan_trace_start}
\end{figure}


\begin{figure}
\centering
\begin{tikzpicture}
    \begin{axis}[
        xlabel={$x^{(1)}$}, ylabel={$x^{(2)}$},
        xmin=0, xmax=10, ymin=0, ymax=10,
        width=0.5\textwidth, height=0.5\textwidth,
        axis line style={draw=none},
        tick style={black, thin},
        xtick={2,4,6,8}, ytick={2,4,6,8},
        xticklabel style={font=\small},
        yticklabel style={font=\small},
        tick align=outside, tick pos=left,
    ]
    \addplot[black!35, dashed, thin, domain=0:360, samples=60]
        ({1 + 2.5*cos(x)}, {2 + 2.5*sin(x)});
    \addplot[only marks, mark=*, mark size=2pt, black]
        coordinates {(1,2)(2,3)(3,1)};
    \addplot[only marks, mark=*, mark size=2pt, black!30]
        coordinates {(6,8)(8,7)(8,9)(4,5)};
    \node[font=\tiny, anchor=north west] at (axis cs:1,2) {$x_1, C_1$};
    \node[font=\tiny, anchor=south east] at (axis cs:2,3) {$x_2, C_1$};
    \node[font=\tiny, anchor=north]      at (axis cs:3,1) {$x_3, C_1$};
    \node[font=\tiny, anchor=south west] at (axis cs:6,8) {$x_4$};
    \node[font=\tiny, anchor=north west] at (axis cs:8,7) {$x_5$};
    \node[font=\tiny, anchor=south west] at (axis cs:8,9) {$x_6$};
    \node[font=\tiny, anchor=south west] at (axis cs:4,5) {$x_7$};
    \node[font=\tiny, black!55, anchor=south west, inner sep=1pt] at (axis cs:1.1,2.1) {3};
    \end{axis}
\end{tikzpicture}
\caption{\newthought{Visit $x_1$}: $N_{2.5}(x_1){=}\{x_1,\,x_2,\,x_3\}$, $|N|{=}3 \geq 2$, so $x_1$ is a $p_c$.\\[4pt]
Seed cluster $C_1$; assign $x_1$ to $C_1$ and add $x_2$ and $x_3$ to the expansion queue. Points outside the dashed circle remain unvisited.\\[4pt]
Distances: $d(x_1,x_2){=}\sqrt{2}{\approx}1.41\;\checkmark$,\; $d(x_1,x_3){=}\sqrt{5}{\approx}2.24\;\checkmark$.}
\label{fig:dbscan_trace_x1}
\end{figure}


\begin{figure}
\centering
\begin{tikzpicture}
    \begin{axis}[
        xlabel={$x^{(1)}$}, ylabel={$x^{(2)}$},
        xmin=0, xmax=10, ymin=0, ymax=10,
        width=0.5\textwidth, height=0.5\textwidth,
        axis line style={draw=none},
        tick style={black, thin},
        xtick={2,4,6,8}, ytick={2,4,6,8},
        xticklabel style={font=\small},
        yticklabel style={font=\small},
        tick align=outside, tick pos=left,
    ]
    \addplot[black!35, dashed, thin, domain=0:360, samples=60]
        ({2 + 2.5*cos(x)}, {3 + 2.5*sin(x)});
    \addplot[black!35, dashed, thin, domain=0:360, samples=60]
        ({3 + 2.5*cos(x)}, {1 + 2.5*sin(x)});
    \addplot[only marks, mark=*, mark size=2pt, black]
        coordinates {(1,2)(2,3)(3,1)};
    \addplot[only marks, mark=*, mark size=2pt, black!30]
        coordinates {(6,8)(8,7)(8,9)(4,5)};
    \node[font=\tiny, anchor=north west] at (axis cs:1,2) {$x_1, C_1$};
    \node[font=\tiny, anchor=south east] at (axis cs:2,3) {$x_2, C_1$};
    \node[font=\tiny, anchor=north]      at (axis cs:3,1) {$x_3, C_1$};
    \node[font=\tiny, anchor=south west] at (axis cs:6,8) {$x_4$};
    \node[font=\tiny, anchor=north west] at (axis cs:8,7) {$x_5$};
    \node[font=\tiny, anchor=south west] at (axis cs:8,9) {$x_6$};
    \node[font=\tiny, anchor=south west] at (axis cs:4,5) {$x_7$};
    \node[font=\tiny, black!55, anchor=south west, inner sep=1pt] at (axis cs:2.1,3.1) {3};
    \node[font=\tiny, black!55, anchor=south west, inner sep=1pt] at (axis cs:3.1,1.1) {3};
    \end{axis}
\end{tikzpicture}
\caption{\newthought{Expand $C_1$}: dequeue $x_2$: $N_{2.5}(x_2){=}\{x_1,x_2,x_3\}$, all already assigned to $C_1$, no new points added.\\[4pt]
Dequeue $x_3$: $N_{2.5}(x_3){=}\{x_1,x_2,x_3\}$, same result.\\[4pt]
Expansion queue empty: $C_1{=}\{x_1,\,x_2,\,x_3\}$ is sealed.}
\label{fig:dbscan_trace_expand_c1}
\end{figure}


\begin{figure}
\centering
\begin{tikzpicture}
    \begin{axis}[
        xlabel={$x^{(1)}$}, ylabel={$x^{(2)}$},
        xmin=0, xmax=10, ymin=0, ymax=10,
        width=0.5\textwidth, height=0.5\textwidth,
        axis line style={draw=none},
        tick style={black, thin},
        xtick={2,4,6,8}, ytick={2,4,6,8},
        xticklabel style={font=\small},
        yticklabel style={font=\small},
        tick align=outside, tick pos=left,
    ]
    % x4 neighbourhood only
    \addplot[black!35, dashed, thin, domain=0:360, samples=60]
        ({6 + 2.5*cos(x)}, {8 + 2.5*sin(x)});
    % distance connectors from x4
    \draw[black!40, thin, dotted] (axis cs:6,8) -- (axis cs:8,7);
    \draw[black!40, thin, dotted] (axis cs:6,8) -- (axis cs:8,9);
    % C1 sealed (dark, no labels — already traced)
    \addplot[only marks, mark=*, mark size=2pt, black]
        coordinates {(1,2)(2,3)(3,1)};
    % C2 forming (grey)
    \addplot[only marks, mark=*, mark size=2pt, black!55]
        coordinates {(6,8)(8,7)(8,9)};
    % x7 still unvisited
    \addplot[only marks, mark=*, mark size=2pt, black!30]
        coordinates {(4,5)};
    \node[font=\tiny, anchor=north west] at (axis cs:1,2)   {$x_1$};
    \node[font=\tiny, anchor=south east] at (axis cs:2,3)   {$x_2$};
    \node[font=\tiny, anchor=north]      at (axis cs:3,1)   {$x_3$};
    \node[font=\tiny, anchor=south west] at (axis cs:6,8)   {$x_4, C_2$};
    \node[font=\tiny, anchor=north west] at (axis cs:8,7)   {$x_5, C_2$};
    \node[font=\tiny, anchor=south west] at (axis cs:8,9)   {$x_6, C_2$};
    \node[font=\tiny, anchor=south west] at (axis cs:4,5)   {$x_7$};
    % neighbourhood count
    \node[font=\tiny, black!55, anchor=south east, inner sep=1pt] at (axis cs:5.9,8.1) {3};
    % distance labels
    \node[font=\tiny, black!40, anchor=north west, inner sep=1pt] at (axis cs:7.1,7.4) {$\sqrt{5}$};
    \node[font=\tiny, black!40, anchor=south west, inner sep=1pt] at (axis cs:7.1,8.6) {$\sqrt{5}$};
    \end{axis}
\end{tikzpicture}
\caption{\newthought{Visit $x_4$}: $N_{2.5}(x_4){=}\{x_4,\,x_5,\,x_6\}$, $|N|{=}3 \geq 2$, so $x_4$ is a $p_c$.\\[4pt]
Seed $C_2$; expand from $x_5$ and $x_6$: both are $p_c$'s and their neighbourhoods cover only $\{x_4,x_5,x_6\}$.\\[4pt]
Distances: $d(x_4,x_5){=}\sqrt{5}{\approx}2.24\;\checkmark$,\; $d(x_4,x_6){=}\sqrt{5}{\approx}2.24\;\checkmark$,\; $d(x_4,x_7){=}\sqrt{13}{\approx}3.61\;\times$.\\[4pt]
$C_2{=}\{x_4,\,x_5,\,x_6\}$ sealed. Only $x_7$ remains unvisited.}
\label{fig:dbscan_trace_c2}
\end{figure}


\begin{figure}
\centering
\vspace{0.5em}
\begin{tikzpicture}
    \begin{axis}[
        xlabel={$x^{(1)}$}, ylabel={$x^{(2)}$},
        xmin=0, xmax=10, ymin=0, ymax=10,
        width=0.5\textwidth, height=0.5\textwidth,
        axis line style={draw=none},
        tick style={black, thin},
        xtick={2,4,6,8}, ytick={2,4,6,8},
        xticklabel style={font=\small},
        yticklabel style={font=\small},
        tick align=outside, tick pos=left,
    ]
    % x8 hypothetical neighbourhood — unique anchor for this figure
    \addplot[black!35, dashed, thin, domain=0:360, samples=60]
        ({8 + 2.5*cos(x)}, {5.5 + 2.5*sin(x)});
    % C1 sealed (dark)
    \addplot[only marks, mark=*, mark size=2pt, black]
        coordinates {(1,2)(2,3)(3,1)};
    % C2 forming (grey)
    \addplot[only marks, mark=*, mark size=2pt, black!55]
        coordinates {(6,8)(8,7)(8,9)};
    % x7 still unvisited
    \addplot[only marks, mark=*, mark size=2pt, black!30]
        coordinates {(4,5)};
    % x8 hypothetical point
    \addplot[only marks, mark=*, mark size=2pt, black!55]
        coordinates {(8,5.5)};
    \node[font=\tiny, anchor=north west] at (axis cs:1,2) {$x_1, C_1$};
    \node[font=\tiny, anchor=south east] at (axis cs:2,3) {$x_2, C_1$};
    \node[font=\tiny, anchor=north]      at (axis cs:3,1) {$x_3, C_1$};
    \node[font=\tiny, anchor=south west] at (axis cs:6,8) {$x_4, C_2$};
    \node[font=\tiny, anchor=north west] at (axis cs:8,7) {$x_5, C_2$};
    \node[font=\tiny, anchor=south west] at (axis cs:8,9) {$x_6, C_2$};
    \node[font=\tiny, anchor=south west] at (axis cs:4,5) {$x_7$};
    \node[font=\tiny, anchor=west]       at (axis cs:8.15,5.5) {$x_8, C_2$};
    \node[font=\tiny, black!55, anchor=south east, inner sep=1pt] at (axis cs:5.9,8.1) {3};
    \node[font=\tiny, black!55, anchor=south west, inner sep=1pt] at (axis cs:8.1,7.1) {4};
    \node[font=\tiny, black!55, anchor=south east, inner sep=1pt] at (axis cs:7.9,9.1) {3};
    \node[font=\tiny, black!55, anchor=east, inner sep=1pt]       at (axis cs:7.85,5.5) {2};
    \end{axis}
\end{tikzpicture}
\caption{\newthought{Hypothetical Point $x_8{=}(8,\,5.5)$}: $d(x_5,x_8){=}1.5 \leq \varepsilon$, so $x_5$ absorbs $x_8$ into its neighbourhood.\\[4pt]
$N_{2.5}(x_8){=}\{x_8,\,x_5\}$, so $|N|{=}2 < n_{\min}{=}3$; $x_8$ is a border point.
Because $x_5$ is already a core point of $C_2$, $x_8$ is classified as $C_2$.\\[4pt]}
\label{fig:hypothetical_border_x8}
\end{figure}


\begin{figure}
\centering
\begin{tikzpicture}
    \begin{axis}[
        xlabel={$x^{(1)}$}, ylabel={$x^{(2)}$},
        xmin=0, xmax=10, ymin=0, ymax=10,
        width=0.5\textwidth, height=0.5\textwidth,
        axis line style={draw=none},
        tick style={black, thin},
        xtick={2,4,6,8}, ytick={2,4,6,8},
        xticklabel style={font=\small},
        yticklabel style={font=\small},
        tick align=outside, tick pos=left,
    ]
    % C1 sealed (dark)
    \addplot[only marks, mark=*, mark size=2pt, black]
        coordinates {(1,2)(2,3)(3,1)};
    % C2 sealed (grey)
    \addplot[only marks, mark=*, mark size=2pt, black!55]
        coordinates {(6,8)(8,7)(8,9)};
    % x7 noise
    \addplot[only marks, mark=x, mark size=2pt, black, thick]
        coordinates {(4,5)};
    % epsilon neighbourhood circle around x7 (radius = 2.5 in data coords)
    \addplot[black!40, dashed, thin, domain=0:360, samples=60]
        ({4 + 2.5*cos(x)}, {5 + 2.5*sin(x)});
    % labels
    \node[font=\tiny, anchor=north west] at (axis cs:1,2) {$x_1, C_1$};
    \node[font=\tiny, anchor=south east] at (axis cs:2,3) {$x_2, C_1$};
    \node[font=\tiny, anchor=north]      at (axis cs:3,1) {$x_3, C_1$};
    \node[font=\tiny, anchor=south west] at (axis cs:6,8) {$x_4, C_2$};
    \node[font=\tiny, anchor=north west] at (axis cs:8,7) {$x_5, C_2$};
    \node[font=\tiny, anchor=south west] at (axis cs:8,9) {$x_6, C_2$};
    \node[font=\tiny, anchor=south west] at (axis cs:4,5) {$x_7, p_n$};
    \end{axis}
\end{tikzpicture}
\caption{\newthought{Visit $x_7$}: its two nearest neighbours are $x_2$ at $d{\approx}2.83$ and $x_4$ at $d{\approx}3.61$, both beyond $\varepsilon{=}2.5$. So $N_{2.5}(x_7){=}\{x_7\}$, $|N|{=}1 < 2$, and $x_7$ is labelled $p_n$.\\[4pt]
All points visited. Final result: $C_1{=}\{x_1,x_2,x_3\}$, $C_2{=}\{x_4,x_5,x_6\}$, and $x_7$ left as noise. No $K$ was chosen; the noise point was never forced into a cluster.}
\label{fig:dbscan_trace_noise}
\end{figure}



\subsection{Hyperparameter Selection}\index{DBSCAN!parameter selection}\index{hyperparameter selection}

\newthought{Choosing $n_{\min}$} follows a practical rule of thumb: set $n_{\min} \geq d{+}1$ where $d$ 
is the dimension of the feature space.
For two-dimensional data, $n_{\min}{=}3$ or $4$ is typical; higher dimensions require larger values to keep sparse neighbourhoods 
from being treated as dense. 

\begin{marginfigure}
\centering
\begin{tikzpicture}
    \begin{axis}[
        xlabel={$x^{(1)}$},
        xmin=0, xmax=6,
        ymin=-0.5, ymax=0.5,
        width=\marginparwidth, height=0.5\marginparwidth,
        axis line style={draw=none},
        tick style={black, thin},
        xtick={1,3,5},
        ytick=\empty,
        xticklabel style={font=\tiny},
        tick align=outside, tick pos=left,
    ]
    % epsilon boundaries as vertical dashed lines
    \draw[black!35, dashed, thin] (axis cs:1,-0.4) -- (axis cs:1,0.4);
    \draw[black!35, dashed, thin] (axis cs:5,-0.4) -- (axis cs:5,0.4);
    % three points on the line
    \addplot[only marks, mark=*, mark size=2pt, black]
        coordinates {(1,0)(3,0)(5,0)};
    \node[font=\tiny, anchor=south] at (axis cs:1,0.08) {$x_a$};
    \node[font=\tiny, anchor=south] at (axis cs:3,0.08) {$x_b$};
    \node[font=\tiny, anchor=south] at (axis cs:5,0.08) {$x_c$};
    \end{axis}
\end{tikzpicture}
\caption{How many points do you need at least to fence in a point in 2D? 3D?}
\label{fig:nmin_intuition}
\end{marginfigure}


\newthought{Choosing $\varepsilon$} uses the $k$-distance plot\index{k-distance plot}: for each point, 
compute the distance $d_k(x_i)$ to its $k$-th nearest neighbour at $k{=}n_{\min}{-}1$.

\vspace{1.5em}
\begin{figure}
\centering
\begin{tikzpicture}
    \begin{axis}[
        xlabel={$x^{(1)}$}, ylabel={$x^{(2)}$},
        xmin=0, xmax=10, ymin=0, ymax=10,
        width=0.5\textwidth, height=0.5\textwidth,
        axis line style={draw=none},
        tick style={black, thin},
        xtick={2,4,6,8}, ytick={2,4,6,8},
        xticklabel style={font=\small},
        yticklabel style={font=\small},
        tick align=outside, tick pos=left,
    ]
    % x4 epsilon-neighbourhood circle (radius = 2.5)
    \addplot[black!35, dashed, thin, domain=0:360, samples=60]
        ({6 + 2.5*cos(x)}, {8 + 2.5*sin(x)});
    % connector x4 -- x5
    \draw[black!40, thin, dashed] (axis cs:6,8) -- (axis cs:8,7);
    % C1 (dimmed -- not the focus)
    \addplot[only marks, mark=*, mark size=2pt, black!30]
        coordinates {(1,2)(2,3)(3,1)};
    % C2
    \addplot[only marks, mark=*, mark size=2pt, black!55]
        coordinates {(6,8)(8,7)(8,9)};
    % x7
    \addplot[only marks, mark=*, mark size=2pt, black!30]
        coordinates {(4,5)};
    \node[font=\tiny, anchor=north west] at (axis cs:1,2)   {$x_1$};
    \node[font=\tiny, anchor=south east] at (axis cs:2,3)   {$x_2$};
    \node[font=\tiny, anchor=north]      at (axis cs:3,1)   {$x_3$};
    \node[font=\tiny, anchor=south west] at (axis cs:6,8)   {$x_4$};
    \node[font=\tiny, anchor=north west] at (axis cs:8,7)   {$x_5$};
    \node[font=\tiny, anchor=south west] at (axis cs:8,9)   {$x_6$};
    \node[font=\tiny, anchor=south west] at (axis cs:4,5)   {$x_7$};
    \end{axis}
\end{tikzpicture}
\caption{\newthought{$k$-th nearest neighbour of $x_4$}: $n_{\min}{=}2$, $k{=}n_{\min}{-}1{=}1$. The dashed circle shows the $\varepsilon{=}2.5$ neighbourhood; the nearest neighbour of $x_4$ is $x_5$ at distance $d_k(x_4){=}\sqrt{5}{\approx}2.24$, well inside $\varepsilon$.\\[4pt]}
\label{fig:kdistance_x4}
\end{figure}

\marginnote{\newthought{Sensitivity sweep.} Try a few $n_{\min}$ values ($d{+}1$, $2d$, etc.) and compare the resulting $k$-distance plots. The value that gives the sharpest, most unambiguous elbow is usually the right one.}


\vspace{1.5em}
and sort these distances in ascending order,
\begin{align*}
    d_k^{(1)} \leq d_k^{(2)} \leq \ldots \leq d_k^{(n)}.
\end{align*}

\vspace{1.5em}
The curve rises slowly in dense regions and sharply in sparse ones; the elbow marks the transition and is the recommended $\varepsilon$.
Notice, that the sorting does not reflect cluster membership: points from different clusters can be interleaved in the $k$-distance plot. 


\begin{figure}
\centering
\begin{tikzpicture}
    \begin{axis}[
        xlabel={points (sorted by $d_k$)},
        ylabel={$d_k$ ($k{=}1$)},
        xmin=0.5, xmax=7.5,
        ymin=0, ymax=4.2,
        width=0.5\textwidth, height=0.4\textwidth,
        axis line style={draw=none},
        tick style={black, thin},
        xtick={1,2,3,4,5,6,7},
        xticklabels={$x_1$,$x_2$,$x_5$,$x_6$,$x_3$,$x_4$,$x_7$},
        ytick={0,1,2,3,4},
        xticklabel style={font=\small},
        yticklabel style={font=\small},
        tick align=outside, tick pos=left,
    ]
    % Elbow point (filled): last dense point before the jump
    \addplot[only marks, mark=o, mark size=2pt, black] coordinates {(6,2.24)};
    % Other points (open)
    \addplot[only marks, mark=*, mark size=2pt, black] coordinates {(1,1.41) (2,1.41) (3,2.00) (4,2.00) (5,2.24) (7,3.61)};
    % Line segments with gaps around each dot (wide gaps for Tufte style)
    \draw[black, thin] (axis cs:1.18,1.41) -- (axis cs:1.82,1.41);
    \draw[black, thin] (axis cs:2.18,1.48) -- (axis cs:2.82,1.90);
    \draw[black, thin] (axis cs:3.18,2.00) -- (axis cs:3.82,2.00);
    \draw[black, thin] (axis cs:4.18,2.03) -- (axis cs:4.82,2.20);
    \draw[black, thin] (axis cs:5.18,2.24) -- (axis cs:5.82,2.24);
    \draw[black, thin] (axis cs:6.18,2.38) -- (axis cs:6.82,3.44);
    % chosen epsilon cut line
    \draw[densely dashed, black!60] (axis cs:0.5,2.5) -- (axis cs:7.5,2.5);
    \node[font=\small, anchor=north west] at (axis cs:0.55,3.7) {$\varepsilon{=}2.5$};
    \end{axis}
\end{tikzpicture}
\caption{$k$-distance plot for the seven canonical points with $k{=}n_{\min}{-}1{=}1$. 
Each $d_k^{(i)}$ is the distance from one point to its nearest neighbour, sorted ascending. 
The first six values sit at $\sqrt{2}{\approx}1.41$, $2.00$, or $\sqrt{5}{\approx}2.24$; $d_k^{(7)}$ jumps to $\sqrt{13}{\approx}3.61$ 
(the lone $x_7$). The unfilled dot marks the elbow, and any $\varepsilon$ in the gap, such as $\varepsilon{=}2.5$ (dashed), 
separates the dense six from the sparse outlier.}
\label{fig:kdistance_plot}
\end{figure}








\newpage
\subsection{When DB Scan Fails}\index{failure modes!DBSCAN}

\newthought{DB Scan has its own pathologies.}
Every density-based method sets a single threshold.
The most common arise from the single global $\varepsilon$, which filters or bridges.
Then we have the curse of dimensionality, which makes all points look far apart and destroys the notion of density.


\vspace{2.5em}
\marginnote{\newthought{Varying Density}}

\begin{figure}
\centering
\begin{tikzpicture}
    \begin{axis}[
        xlabel={$x^{(1)}$}, ylabel={$x^{(2)}$},
        xmin=-1, xmax=14, ymin=-1, ymax=7,
        width=0.5\textwidth,
        axis equal,
        axis line style={draw=none},
        tick style={black, thin},
        xtick={0,4,8,12}, ytick={0,3,6},
        xticklabel style={font=\small},
        yticklabel style={font=\small},
        tick align=outside, tick pos=left,
    ]
    % Epsilon circles on left cluster (eps = 2.0) — heavily overlapping
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({0.5+2.0*cos(x)},{1.0+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({1.0+2.0*cos(x)},{0.5+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({1.0+2.0*cos(x)},{1.5+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({1.5+2.0*cos(x)},{1.0+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({0.8+2.0*cos(x)},{1.2+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({0.6+2.0*cos(x)},{0.7+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({1.2+2.0*cos(x)},{0.8+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({1.3+2.0*cos(x)},{1.3+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({0.9+2.0*cos(x)},{1.6+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({0.4+2.0*cos(x)},{1.4+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({1.1+2.0*cos(x)},{1.1+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({0.7+2.0*cos(x)},{0.9+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({1.4+2.0*cos(x)},{0.6+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({0.5+2.0*cos(x)},{1.3+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({1.0+2.0*cos(x)},{1.0+2.0*sin(x)});
    % Epsilon circles on right cluster (eps = 2.0)
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({5.0+2.0*cos(x)},{3.0+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({7.0+2.0*cos(x)},{2.0+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({9.0+2.0*cos(x)},{4.5+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({8.5+2.0*cos(x)},{1.5+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({6.0+2.0*cos(x)},{5.0+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({10.5+2.0*cos(x)},{3.0+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({11.0+2.0*cos(x)},{4.0+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({6.5+2.0*cos(x)},{3.5+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({8.0+2.0*cos(x)},{5.5+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({10.0+2.0*cos(x)},{2.0+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({7.5+2.0*cos(x)},{4.5+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({9.5+2.0*cos(x)},{2.5+2.0*sin(x)});
    % Tight, dense cluster
    \addplot[only marks, mark=*, mark size=2pt, black] coordinates {
        (0.5,1.0)(1.0,0.5)(1.0,1.5)(1.5,1.0)(0.8,1.2)
        (0.6,0.7)(1.2,0.8)(1.3,1.3)(0.9,1.6)(0.4,1.4)
        (1.1,1.1)(0.7,0.9)(1.4,0.6)(0.5,1.3)(1.0,1.0)
    };
    % Spread, sparse cluster
    \addplot[only marks, mark=*, mark size=2pt, black!55] coordinates {
        (5.0,3.0)(7.0,2.0)(9.0,4.5)(8.5,1.5)(6.0,5.0)(10.5,3.0)
        (11.0,4.0)(6.5,3.5)(8.0,5.5)(10.0,2.0)(7.5,4.5)(9.5,2.5)
    };
    % epsilon label
    \node[font=\tiny, black!55, anchor=south west] at (axis cs:1.6,2.8) {$\varepsilon{=}2$};
    \end{axis}
\end{tikzpicture}
\caption{Varying cluster density. An $\varepsilon$ tight enough to resolve the left cluster
leaves the right cluster disconnected; an $\varepsilon$ large enough to connect the right cluster
merges everything on the left into one blob, loosing expressiveness of the clustering. No single global threshold works for both.}
\label{fig:dbscan_fail_density}
\end{figure}

\marginnote{\newthought{Bridging and Chaining}}\index{chaining!DBSCAN}\index{bridging}

\begin{figure}
\centering
\begin{tikzpicture}
    \begin{axis}[
        xlabel={$x^{(1)}$}, ylabel={$x^{(2)}$},
        xmin=-1, xmax=14, ymin=-1, ymax=7,
        width=0.5\textwidth,
        axis equal,
        axis line style={draw=none},
        tick style={black, thin},
        xtick={0,4,8,12}, ytick={0,3,6},
        xticklabel style={font=\small},
        yticklabel style={font=\small},
        tick align=outside, tick pos=left,
    ]
    % Epsilon circles along the bridge chain (eps = 2.0)
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({1.5+2.0*cos(x)},{2.5+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({3.0+2.0*cos(x)},{2.0+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({4.5+2.0*cos(x)},{2.1+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({6.0+2.0*cos(x)},{2.0+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({7.5+2.0*cos(x)},{1.9+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({9.0+2.0*cos(x)},{2.0+2.0*sin(x)});
    \addplot[black!25, dashed, thin, domain=0:360, samples=40] ({10.0+2.0*cos(x)},{2.0+2.0*sin(x)});
    % Left cluster
    \addplot[only marks, mark=*, mark size=2pt, black] coordinates {
        (0.5,2.0)(1.0,1.5)(1.5,2.5)(1.0,2.8)(0.8,2.2)(1.3,1.8)(0.6,2.6)
    };
    % Right cluster
    \addplot[only marks, mark=*, mark size=2pt, black!55] coordinates {
        (10.0,2.0)(10.5,1.5)(11.0,2.5)(10.5,2.8)(10.3,2.2)(10.8,1.8)(10.1,2.6)
    };
    % Bridge: chain of dense intermediates
    \addplot[only marks, mark=*, mark size=2pt, black!30] coordinates {
        (3.0,2.0)(4.5,2.1)(6.0,2.0)(7.5,1.9)(9.0,2.0)
    };
    \end{axis}
\end{tikzpicture}
\caption{Bridging: a thin chain of $p_c$-eligible points (light) spans the gap between
two well-separated clusters. The overlapping $\varepsilon$-circles show how density reachability chains through $p_c$'s,
fusing both into one cluster. Increasing $n_{\min}$ or decreasing $\varepsilon$
breaks the bridge at the cost of shrinking the clusters themselves.}
\label{fig:dbscan_fail_bridging}
\end{figure}

\newthought{OPTICS}\index{OPTICS} or Ordering Points To Identify the Clustering Structure~\cite{Ankerst1999}
introduces two per-point distances. The core distance $\varepsilon_c$ is the per-point analogue of 
the global $\varepsilon$, the smallest radius that captures $n_{\min}$ neighbours:
\begin{align*}
    \varepsilon_c(x_i) &= \min\{r : |N_r(x_i)| \geq n_{\min}\}.
\end{align*}
The reachability distance $r(x_i, x_j)$ is defined for every pair where $x_i$ is an already-visited $p_c$
and $x_j$ is an unvisited point. It measures the cost for $x_i$ to absorb $x_j$,
and is at least $\varepsilon_c(x_i)$, even if $x_j$ sits closer:
\begin{align*}
    r(x_i, x_j) &= \max\!\big\{\varepsilon_c(x_i),\; d(x_i, x_j)\big\}.
\end{align*}

\marginnote{Dense regions have small $\varepsilon_c$, sparse regions large ones.
The $\max$ ensures that a point is never reached at a distance shorter than its $\varepsilon_c$,
so reachability adapts to local density automatically.}






\newthought{HDBSCAN}\index{HDBSCAN} or Hierarchical DBSCAN~\cite{Campello2013} replaces the Euclidean distance with a 
mutual reachability distance:
\begin{align*}
    d_{\mathrm{mreach}}(x_i, x_j) &= \max\!\big\{\varepsilon_c(x_i),\; \varepsilon_c(x_j),\; d(x_i, x_j)\big\}.
\end{align*}
This spreads apart points in sparse regions; exactly those that should not be merged early.
From here HDBSCAN builds a full density hierarchy by applying single-linkage clustering to the mutual reachability distance, and 
potentially condenses it into a flat clustering by selecting the most persistent clusters across all density levels.

\marginnote{\newthought{Density hierarchies and dendrograms.} 
The density language and the hierarchical-clustering language describe the same structure from different angles.}

\newthought{The only parameter} the user must set is $n_{\min}$, and the result is often robust to its choice.




















\vspace{3.5em}


\newthought{The curse of dimensionality}\index{curse of dimensionality}: All distance-based methods share a deeper vulnerability as the number of features $d$ grows,
the volume of the space increases so fast that a fixed number of points $N$ becomes sparse everywhere.\index{high dimensionality}

\begin{marginfigure}
\centering
\begin{tikzpicture}
    \begin{axis}[
        xlabel={$x^{(1)}$},
        xmin=-0.1, xmax=1.1,
        ymin=-0.15, ymax=0.15,
        width=\marginparwidth,
        axis equal,
        axis line style={draw=none},
        tick style={black, thin},
        xtick={0,0.5,1},
        ytick=\empty,
        xticklabel style={font=\tiny},
        tick align=outside, tick pos=left,
    ]
    % unit interval
    \draw[black!20, thin] (axis cs:0,0) -- (axis cs:1,0);
    % 10 uniform points on [0,1]
    \addplot[only marks, mark=*, mark size=1.5pt, black]
        coordinates {(0.05,0)(0.14,0)(0.23,0)(0.35,0)(0.42,0)
                     (0.56,0)(0.67,0)(0.74,0)(0.88,0)(0.97,0)};
    % mean NN spacing circle around central point
    \addplot[black!40, dashed, thin, domain=0:360, samples=60, smooth]
        ({0.42 + 0.1*cos(x)}, {0 + 0.1*sin(x)});
    \end{axis}
\end{tikzpicture}
\caption{$N{=}10$ points in $[0,1]^1$: mean NN spacing ${\approx}10^{-1}{=}0.1$. The line is well covered.}
\label{fig:cod_spacing_1d}
\end{marginfigure}

\marginnote{\newthought{$\propto$} means ``proportional to'': $a \propto b$ says $a = c \cdot b$ for some constant $c$ independent of the variables of interest.}


\begin{marginfigure}[1.5em]
\centering
\begin{tikzpicture}
    \begin{axis}[
        xlabel={$x^{(1)}$}, ylabel={$x^{(2)}$},
        xmin=-0.15, xmax=1.15, ymin=-0.15, ymax=1.15,
        width=\marginparwidth, height=\marginparwidth,
        axis line style={draw=none},
        tick style={black, thin},
        xtick={0,0.5,1}, ytick={0,0.5,1},
        xticklabel style={font=\tiny},
        yticklabel style={font=\tiny},
        tick align=outside, tick pos=left,
    ]
    % 10 uniform points in [0,1]^2
    \addplot[only marks, mark=*, mark size=1.5pt, black]
        coordinates {(0.12,0.08)(0.85,0.15)(0.35,0.42)(0.72,0.68)
                     (0.08,0.75)(0.55,0.22)(0.92,0.88)(0.28,0.92)
                     (0.65,0.45)(0.18,0.38)};
    % mean NN spacing circle around central point
    \addplot[black!40, dashed, thin, domain=0:360, samples=60, smooth]
        ({0.65 + 0.32*cos(x)}, {0.45 + 0.32*sin(x)});
    \end{axis}
\end{tikzpicture}
\caption{Same $N{=}10$ points in $[0,1]^2$: the space has grown from a line to a square, 
and the mean NN spacing rises to $10^{-1/2}{\approx}0.32$.}
\label{fig:cod_spacing_2d}
\end{marginfigure}

\newthought{Consider $N$ points} drawn uniformly in the unit hypercube $[0,1]^d$.
The mean nearest-neighbour distance scales as
\begin{align*}
    d_{\mathrm{NN}} \;\propto\; N^{-1/d}.
\end{align*}
Fix $N{=}10$ and watch the spacing grow with $d$:
\begin{align*}
    d =  1: \quad & 10^{-1/1}  = 0.10 \\
    d =  2: \quad & 10^{-1/2}  \approx 0.32 \\
    d = 3: \quad & 10^{-1/3}  \approx 0.46 \\
    d = 10: \quad & 10^{-1/10} \approx 0.79 \\
    d = 50: \quad & 10^{-1/50} \approx 0.96
\end{align*}
By $d{=}50$ every point is nearly as far from its neighbour as from the boundary of the cube.
The points have not moved apart; the space has grown underneath them.

\vspace{1.5em}
\newthought{For distance-based methods this is fatal.} The $\varepsilon$-neighbourhood of a point captures a hypersphere whose volume
shrinks relative to the cube: $V_{\text{sphere}} / V_{\text{cube}} \to 0$ as $d$ grows.
A radius that includes many neighbours in two dimensions encloses almost nothing in fifty.
Either $\varepsilon$ must grow until every point is a neighbour, collapsing all structure into one cluster, or
it stays small and every point is labelled $p_n$.














% ==============================================================================================
% BLOCK 3: STUDENT ACTIVATION
% ==============================================================================================

\newpage

\section{Examples \& Exercises}\index{examples}\index{exercises}

\newthought{Computing by hand} reveals what the algorithm actually does at each step; run it slowly and feel where density propagates and where it stops.

\vspace{1em}

\begin{marginfigure}
    \centering
    \begin{tikzpicture}
        \begin{axis}[
            xmin=0, xmax=10,
            ymin=0, ymax=10,
            width=\marginparwidth,
            height=\marginparwidth,
            axis line style={draw=none},
            tick style={black, thin},
            xtick={0,3,6,9},
            ytick={0,3,6,9},
            xticklabel style={font=\tiny},
            yticklabel style={font=\tiny},
            tick align=outside,
            tick pos=left,
            xlabel={$x^{(1)}$},
            ylabel={$x^{(2)}$},
        ]
        \addplot[only marks, mark=*, mark size=2pt, black]    coordinates {(1,2)(2,3)(3,1)};
        \addplot[only marks, mark=*, mark size=2pt, black!55] coordinates {(6,8)(8,7)(8,9)};
        \addplot[only marks, mark=*, mark size=2pt, black!30] coordinates {(4,5)};
        \node[font=\tiny, anchor=south east] at (axis cs:1,2) {$x_1$};
        \node[font=\tiny, anchor=south east] at (axis cs:2,3) {$x_2$};
        \node[font=\tiny, anchor=north]      at (axis cs:3,1) {$x_3$};
        \node[font=\tiny, anchor=south west] at (axis cs:6,8) {$x_4$};
        \node[font=\tiny, anchor=north west] at (axis cs:8,7) {$x_5$};
        \node[font=\tiny, anchor=south west] at (axis cs:8,9) {$x_6$};
        \node[font=\tiny, anchor=south west] at (axis cs:4,5) {$x_7$};
        \end{axis}
    \end{tikzpicture}
    \caption{Seven points for the DBSCAN trace. $x_7{=}(4,5)$ sits between the two natural groups.}
    \label{fig:dbscan_exercise_scatter}
\end{marginfigure}

\newthought{Given seven points} $x_1{=}(1,2)$, $x_2{=}(2,3)$, $x_3{=}(3,1)$, $x_4{=}(6,8)$, $x_5{=}(8,7)$, $x_6{=}(8,9)$, $x_7{=}(4,5)$, run DBSCAN with $\varepsilon{=}2.5$ and $n_{\min}{=}2$.
Your tasks: compute the neighbourhood size $|N_{2.5}(x_i)|$ for each point, classify each point as $p_c$, $p_b$, or $p_n$, and identify the clusters.

\vspace{0.5em}

\newthought{Neighbourhood sizes.} For each point $x_i$ count all points within distance 2.5 (including $x_i$ itself).
Here is the full calculation for $x_1$:

\begin{align*}
    d(x_1,\,x_2) &= \sqrt{(2-1)^2 + (3-2)^2} \\
                 &= \sqrt{1 + 1} \\
                 &= \sqrt{2} \\
                 &\approx 1.41 \quad (\leq 2.5,\; \text{counted}) \\[4pt]
    d(x_1,\,x_3) &= \sqrt{(3-1)^2 + (1-2)^2} \\
                 &= \sqrt{4 + 1} \\
                 &= \sqrt{5} \\
                 &\approx 2.24 \quad (\leq 2.5,\; \text{counted}) \\[4pt]
    d(x_1,\,x_7) &= \sqrt{(4-1)^2 + (5-2)^2} \\
                 &= \sqrt{9 + 9} \\
                 &= \sqrt{18} \\
                 &\approx 4.24 \quad (> 2.5,\; \text{not counted})
\end{align*}

So $N_{2.5}(x_1) = \{x_1,\, x_2,\, x_3\}$ and $|N_{2.5}(x_1)| = 3 \geq 2$: $x_1$ is a $p_c$.
Compute the remaining neighbourhood sizes yourself, then verify against the table below.

\begin{table*}[ht]
\centering
\begin{tabular}{p{0.07\textwidth}p{0.12\textwidth}p{0.10\textwidth}p{0.12\textwidth}p{0.30\textwidth}p{0.10\textwidth}}
    \toprule
    \textbf{Point} & \textbf{Coords} & \textbf{$|N_{2.5}|$} & \textbf{Type} & \textbf{Neighbours in $N_{2.5}$} & \textbf{Cluster} \\
    \midrule
    $x_1$ & $(1,2)$ & $3$ & $p_c$  & $\{x_1,\, x_2,\, x_3\}$          & $C_1$ \\
    $x_2$ & $(2,3)$ & $3$ & $p_c$  & $\{x_1,\, x_2,\, x_3\}$          & $C_1$ \\
    $x_3$ & $(3,1)$ & $3$ & $p_c$  & $\{x_1,\, x_2,\, x_3\}$          & $C_1$ \\
    $x_4$ & $(6,8)$ & $3$ & $p_c$  & $\{x_4,\, x_5,\, x_6\}$          & $C_2$ \\
    $x_5$ & $(8,7)$ & $3$ & $p_c$  & $\{x_4,\, x_5,\, x_6\}$          & $C_2$ \\
    $x_6$ & $(8,9)$ & $3$ & $p_c$  & $\{x_4,\, x_5,\, x_6\}$          & $C_2$ \\
    $x_7$ & $(4,5)$ & $1$ & $p_n$  & $\{x_7\}$ only                    & $-$ \\
    \bottomrule
\end{tabular}
\vspace{2em}
\caption{DBSCAN trace for $\varepsilon{=}2.5$, $n_{\min}{=}2$. All six canonical points are $p_c$'s; $x_7$ has no neighbour within radius 2.5 and is labelled $p_n$. The two natural clusters are recovered without specifying $K$ and without assigning $p_n$ to either group.}
\label{tab:dbscan_trace}
\end{table*}





\newthought{Verify $x_7$ is $p_n$.} Compute $d(x_7,\,x_3)$ to confirm that $x_7$ lies outside the $\varepsilon$-neighbourhood of its nearest candidate:
\begin{align*}
    d(x_7,\,x_3) &= \sqrt{(4-3)^2 + (5-1)^2} \\
                 &= \sqrt{1 + 16} \\
                 &= \sqrt{17} \\
                 &\approx 4.12 \quad (> 2.5)
\end{align*}
Compute $d(x_7,\,x_4)$ yourself.
Why is $x_7$ correctly identified as $p_n$ rather than being assigned to whichever cluster is nearer?


\newthought{Adding a border point.} Place an eighth point $x_8{=}(8,\,5.5)$ into the dataset and raise the density threshold to $n_{\min}{=}3$.
Compute $|N_{2.5}(x_8)|$ and classify $x_8$ as $p_c$, $p_b$, or $p_n$.
Which cluster, if any, does $x_8$ join, and why?
Verify that the six original cluster members remain $p_c$'s under the new threshold.

\begin{marginfigure}
    \centering
    \begin{tikzpicture}
        \begin{axis}[
            xmin=0, xmax=10,
            ymin=0, ymax=10,
            width=\marginparwidth,
            height=\marginparwidth,
            axis line style={draw=none},
            tick style={black, thin},
            xtick={0,3,6,9},
            ytick={0,3,6,9},
            xticklabel style={font=\tiny},
            yticklabel style={font=\tiny},
            tick align=outside,
            tick pos=left,
            xlabel={$x^{(1)}$},
            ylabel={$x^{(2)}$},
        ]
        \addplot[only marks, mark=*, mark size=2pt, black]    coordinates {(1,2)(2,3)(3,1)};
        \addplot[only marks, mark=*, mark size=2pt, black!55] coordinates {(6,8)(8,7)(8,9)};
        \addplot[only marks, mark=*, mark size=2pt, black!30] coordinates {(4,5)};
        \addplot[only marks, mark=*, mark size=2pt, black!55] coordinates {(8,5.5)};
        \node[font=\tiny, anchor=south east] at (axis cs:1,2) {$x_1$};
        \node[font=\tiny, anchor=south east] at (axis cs:2,3) {$x_2$};
        \node[font=\tiny, anchor=north]      at (axis cs:3,1) {$x_3$};
        \node[font=\tiny, anchor=south west] at (axis cs:6,8) {$x_4$};
        \node[font=\tiny, anchor=north west] at (axis cs:8,7) {$x_5$};
        \node[font=\tiny, anchor=south west] at (axis cs:8,9) {$x_6$};
        \node[font=\tiny, anchor=south west] at (axis cs:4,5) {$x_7$};
        \node[font=\tiny, anchor=west]       at (axis cs:8.15,5.5) {$x_8$};
        \end{axis}
    \end{tikzpicture}
    \caption{Eight points for the border-point exercise. $x_8{=}(8,\,5.5)$ sits below $C_2$.}
    \label{fig:dbscan_exercise_borderpoint}
\end{marginfigure}

\newthought{Density reachability from $x_8$.} Still using the eight-point dataset with $\varepsilon{=}2.5$ and $n_{\min}{=}3$:
is $x_8$ density-reachable from $x_4$?
If so, write down the chain of points that connects them.
Now reverse the question: is $x_4$ density-reachable from $x_8$? Why or why not?


\newthought{Parameter sensitivity.} Re-run the trace with $\varepsilon{=}4.5$ and $n_{\min}{=}2$.
Does $x_7$ change classification?
At what minimum value of $\varepsilon$ does $x_7$ first enter the neighbourhood of some $p_c$?

\newthought{Building the $k$-distance plot.} Return to the original seven points with $n_{\min}{=}2$, so $k{=}n_{\min}{-}1{=}1$.
For each point $x_i$, compute the distance to its nearest neighbour $d_1(x_i)$.
Here is $x_5$:
\begin{align*}
    d(x_5,\,x_4) &= \sqrt{(8-6)^2 + (7-8)^2} \\
                 &= \sqrt{5} \\
                 &\approx 2.24 \\[4pt]
    d(x_5,\,x_6) &= \sqrt{(8-8)^2 + (9-7)^2} \\
                 &= \sqrt{4} \\
                 &= 2.00 \\[4pt]
    d_1(x_5) &= 2.00 \quad (\text{nearest neighbour is } x_6)
\end{align*}
Compute $d_1(x_i)$ for the remaining six points. Sort all seven values in ascending order and sketch the curve.
Where is the elbow? Does the $\varepsilon{=}2.5$ from the earlier trace sit in the gap?

\newthought{Comparison with K-Means.} Run K-Means with $K{=}2$ on the same seven points.
Which cluster does K-Means assign $x_7$ to?
Is that assignment more or less informative than DBSCAN's $p_n$ label, and why?

\newthought{K-Means baseline result.} After convergence with $K{=}2$, the centroids settle at $\mu_1 \approx (2.0,\,2.0)$ and $\mu_2 \approx (7.3,\,8.0)$.
Point $x_7{=}(4,5)$ is closer to $\mu_1$:
\begin{align*}
    d(x_7,\,\mu_1) &= \sqrt{(4-2)^2 + (5-2)^2} \\
                   &= \sqrt{4+9} \\
                   &= \sqrt{13} \approx 3.61 \\[4pt]
    d(x_7,\,\mu_2) &= \sqrt{(4-7.3)^2 + (5-8.0)^2} \\
                   &= \sqrt{10.89+9} \\
                   &= \sqrt{19.89} \approx 4.46
\end{align*}
K-Means assigns $x_7$ to $C_1$, as shown below.
The assignment is not wrong in a strict sense; it is simply forced.
K-Means has no $p_n$ concept and cannot abstain.

\begin{figure}
\centering
\begin{tikzpicture}
    \begin{axis}[
        xlabel={$x^{(1)}$}, ylabel={$x^{(2)}$},
        xmin=0, xmax=10, ymin=0, ymax=10,
        width=0.5\textwidth, height=0.5\textwidth,
        axis line style={draw=none},
        tick style={black, thin},
        xtick={2,4,6,8}, ytick={2,4,6,8},
        xticklabel style={font=\small},
        yticklabel style={font=\small},
        tick align=outside, tick pos=left,
    ]
    % C1 points (dark) including x7
    \addplot[only marks, mark=*, mark size=2pt, black]
        coordinates {(1,2)(2,3)(3,1)(4,5)};
    % C1 centroid
    \addplot[only marks, mark=o, mark size=2pt, black, thick]
        coordinates {(2.5,2.75)};
    % C2 points (grey)
    \addplot[only marks, mark=*, mark size=2pt, black!55]
        coordinates {(6,8)(8,7)(8,9)};
    % C2 centroid
    \addplot[only marks, mark=o, mark size=2pt, black!55, thick]
        coordinates {(7.33,8)};
    % Labels
    \node[font=\tiny, anchor=north west] at (axis cs:1,2)   {$x_1$};
    \node[font=\tiny, anchor=south east] at (axis cs:2,3)   {$x_2$};
    \node[font=\tiny, anchor=north]      at (axis cs:3,1)   {$x_3$};
    \node[font=\tiny, anchor=south west] at (axis cs:4,5)   {$x_7$};
    \node[font=\tiny, anchor=south west] at (axis cs:6,8)   {$x_4$};
    \node[font=\tiny, anchor=north west] at (axis cs:8,7)   {$x_5$};
    \node[font=\tiny, anchor=south west] at (axis cs:8,9)   {$x_6$};
    \node[font=\tiny, anchor=south east] at (axis cs:2.5,2.75) {$\mu_1$};
    \node[font=\tiny, anchor=south west] at (axis cs:7.33,8)   {$\mu_2$};
    \end{axis}
\end{tikzpicture}
\caption{K-Means baseline with $K{=}2$. o-marks for centroids $\mu_1$ and $\mu_2$. Because K-Means must assign every point, $x_7$ is forced into $C_1$ (dark). Compare with Figure~\ref{fig:dbscan_trace_noise}: DBSCAN labels $x_7$ as $p_n$ and makes no forced assignment.}
\label{fig:kmeans_baseline}
\end{figure}







\newthought{From global to local density.} Return to the eight-point dataset ($x_1,\ldots,x_7$ plus $x_8{=}(8,\,5.5)$) with $n_{\min}{=}3$.
For each point $x_i$, compute the core distance $\varepsilon_c(x_i)$: the smallest radius that captures at least $n_{\min}$ points, itself included.
Here is $x_5$:
\marginnote{The core distance $\varepsilon_c(x_5){=}2.00$ is the radius needed to capture $\{x_8,\, x_6,\, x_5\}$: exactly three points.}
\begin{align*}
    d(x_5,\,x_8) &= 1.50 \\[4pt]
    d(x_5,\,x_6) &= 2.00 \\[4pt]
    d(x_5,\,x_4) &= \sqrt{5} \\
                 &\approx 2.24 \\[4pt]
    \varepsilon_c(x_5) &= 2.00
\end{align*}
Compute $\varepsilon_c$ for $x_1$ through $x_4$, $x_6$, $x_7$, and $x_8$.
Which points have the largest core distance, and what does that tell you about their local density?

Now compute the mutual reachability distance for two pairs:
\begin{align*}
    d_{\mathrm{mreach}}(x_4,\,x_5) &= \max\!\big\{\varepsilon_c(x_4),\;\varepsilon_c(x_5),\;d(x_4,\,x_5)\big\} \\
    d_{\mathrm{mreach}}(x_7,\,x_2) &= \max\!\big\{\varepsilon_c(x_7),\;\varepsilon_c(x_2),\;d(x_7,\,x_2)\big\}
\end{align*}
Compare each to the raw Euclidean distance. For which pair does $d_{\mathrm{mreach}}$ differ from $d$, and why?
Why does HDBSCAN not need a global $\varepsilon$?

\newthought{The curse of dimensionality by the numbers.} The mean nearest-neighbour spacing scales as $d_{\mathrm{NN}} \propto N^{-1/d}$.
Compute $N^{-1/d}$ for $N{=}100$ at $d = 2,\; 10,\; 50,\; 100$.
At what dimension $d$ does the spacing first exceed $0.9$?
If you tried to run DBSCAN at that dimension, what would happen to the $\varepsilon$-neighbourhoods?


% ---------------------------------------------------------------------------
% Exercise 6: Coding exercise
% ---------------------------------------------------------------------------
\vspace{2.5em}
\marginnote{\newthought{Code} will be provided as a Python notebook.
Use it as a starting point, break things, and observe what changes.}

\newthought{Back at the coffeetable code.}
Same 1\,797 aroma fingerprints, two coordinates.
Implement epsilon-neighborhoods and the core-point classifier,
use the $k$-distance plot to pick $\varepsilon$, then compare
DBSCAN with $K$-Means and Ward at the same cluster count.

\newthought{What to observe.}
Core points fill the dense hearts of each variety; noise points land in the
gaps no method should claim.
DBSCAN follows density contours rather than drawing spherical cuts, so it
agrees with $K$-Means in the cores but diverges at irregular boundaries.
A single $\varepsilon$ cannot capture clusters of very different densities. 
In next steps, learn about sensitivity sweeps and HDBSCAN to address this.



\section{Self-Reflection and Recap}\index{reflection}\index{recap}

\newthought{Self-Reflection} questions to guide your thinking:
\begin{itemize}
    \item What does density reachability gain over distance-to-centroid? Which K-Means failures disappear, which new ones appear?
    \item Why is density reachability asymmetric, and what does that imply for the determinism of $p_b$ labels?
    \item A $p_b$ sits inside two $p_c$'s from different clusters. Is its label unique? What does that say about what a DB Scan cluster really is?
    \item Why is bridging the density-based dual of single-linkage chaining, and why does every remedy cost something?
    \item The $k$-distance plot shows a straight line with no elbow. What does this tell you, and which method would you try next?
    \item What is the essential difference between an OPTICS reachability plot and an HDBSCAN density hierarchy?
    \item DB Scan is connected components on the $\varepsilon$-graph. What extra structure does spectral clustering use on top of that graph?
    \item Why do the zero eigenvalues of $L$ count connected components, and why is that alone not enough for real data?
    \item When would you prefer DB Scan over Ward's method even if both produce a sensible partition?
\end{itemize}

\newthought{Recap} of Key Concepts:
\begin{itemize}
    \item DB Scan clusters are maximal sets of density-connected points; no centroid, no pre-specified $K$
    \item Every point is $p_c$ when $|N_\varepsilon| \geq n_{\min}$, $p_b$ when inside a $p_c$'s neighbourhood without being $p_c$ itself, or $p_n$
    \item Density reachability is transitive along $p_c$'s but not symmetric; $p_b$ labels depend on visit order
    \item The partition is determined by $(X, \varepsilon, n_{\min})$ up to shared $p_b$'s; $O(n)$ queries, no restarts
    \item The $k$-distance plot with $k{=}n_{\min}{-}1$, sorted ascending, chooses $\varepsilon$ at the elbow
    \item Failure modes: varying density, where no global $\varepsilon$ fits two scales, and high dimensionality, where distances concentrate
    \item Bridging fuses clusters through chains of $p_c$'s, the density dual of single-linkage chaining
    \item OPTICS exposes all density scales via a reachability plot; HDBSCAN extracts the most persistent clusters from a full density hierarchy
    \item DB Scan is connected components on the $\varepsilon$-graph; spectral clustering\index{spectral clustering} replaces hard connectivity with the NCut objective, solved through the eigenvectors of $L_{\mathrm{sym}}$\index{Laplacian} followed by K-Means
    \item The eigengap heuristic\index{eigengap heuristic} picks $K$ at the largest jump between consecutive Laplacian eigenvalues
\end{itemize}

\newthought{Two ways out of the curse of dimensionality.} Either we collect exponentially more data~\cite{Halevy2009}, which is rarely possible, or we reduce the dimension so that the samples we already have sit densely in the space we actually care about.
The observation that makes the second route tractable is that real datasets almost never use their ambient dimensions fully: the meaningful variation lives on a much lower-dimensional subset, and the rest is redundancy, correlation, or noise.
Finding that signal-carrying subset is what the next part of these notes is about.
\marginnote{\newthought{Teaser.} The curse of dimensionality breaks every method in this chapter, yet real data rarely uses all its ambient dimensions. Can we find the directions that carry the most signal and build a lower-dimensional representation around them, so that the clustering methods of Part I regain their footing?}
\marginnote{\newthought{Feedback.}}
\newthought{Up next.} Part II, Representing Pattern, leaves the original feature space behind and learns to construct new coordinates.
Chapter 5 opens with principal component analysis as the linear baseline.
Only once a dataset is described in the dimensions that carry signal do the clustering methods of Part I regain their footing.

