% !TEX root = ../notes_unsupervisedlearning.tex
% Chapter 14: World Models and Vision-Language-Action Models
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{World Models and Vision-Language-Action Models}\index{World Models}\index{VLA}\index{Embodied AI}
\newthought{Simulate, imagine, then act in the real world.}

%═══════════════════════════════════════════════════════════════════════
% BLOCK 1: INTRODUCTION
%═══════════════════════════════════════════════════════════════════════

\section{The Why}

\marginnote{Humans imagine consequences before acting; can machines learn to do the same?}

Model-free RL learns from experience, but it is sample-inefficient, requiring millions of environment interactions; unsafe, because learning by trial-and-error can be dangerous; and non-transferable, because the agent must relearn for each new task.

\newthought{Model-based RL}\index{model-based RL} learns a model of the world, enabling planning by simulating actions before taking them, imagination by generating synthetic experience for training, and sample efficiency by learning from fewer real interactions.

\newthought{Vision-Language-Action models}\index{VLA} represent the frontier: systems that see, understand language, and act in the physical world.

\newthought{In order to move from trial-and-error to planning in imagination} we have good reason to find ways to,
\begin{itemize}
    \item learn dynamics models that predict future states and rewards from current states and actions, enabling mental simulation.
    \item drive exploration via intrinsic motivation so that the agent seeks novel states even when extrinsic rewards are sparse.
    \item unify vision, language, and action in a single architecture so that web-scale knowledge transfers to robotic control.
\end{itemize}

\subsection{Hands On Experience}

\begin{marginfigure}
\begin{tikzpicture}[scale=0.7]
    % World model
    \node[draw, rectangle, minimum width=2cm] (wm) at (0,0) {World Model};
    % Input
    \node[draw, rectangle] (s) at (-3,0) {$s_t, a_t$};
    % Output
    \node[draw, rectangle] (sp) at (3,0.5) {$\hat{s}_{t+1}$};
    \node[draw, rectangle] (r) at (3,-0.5) {$\hat{r}_t$};

    \draw[->] (s) -- (wm);
    \draw[->] (wm) -- (sp);
    \draw[->] (wm) -- (r);

    % Planning loop
    \draw[->, dashed] (sp.east) to[bend left] (4.5,0) to[bend left] (wm.north);
    \node at (5,0) {\small Plan};
\end{tikzpicture}
\caption{World model predicts next state and reward, enabling mental simulation.}
\label{fig:world_model_intro}
\end{marginfigure}

Imagine you are learning to play a video game:
\begin{enumerate}
    \item After playing briefly, you can imagine what happens if you jump
    \item You simulate different strategies in your head
    \item You execute the best plan in the real game
\end{enumerate}

This is planning with a learned model: act in imagination, then act in reality.

\newthought{The Learning Objectives} of this lecture:
\begin{itemize}
    \item Explain model-based RL, describe world model architectures such as Dreamer, and identify how model errors compound over long planning horizons.
    \item Implement curiosity-driven exploration and distinguish intrinsic from extrinsic reward signals.
    \item Describe VLA architectures and articulate how web-scale vision-language knowledge transfers to robotic control.
\end{itemize}

%═══════════════════════════════════════════════════════════════════════
% BLOCK 2: MAIN THEORY
%═══════════════════════════════════════════════════════════════════════

\newpage
\section{Model-Based Reinforcement Learning}\index{model-based RL}

\marginnote{Learn a model, then plan with it.}

\subsection{The Dynamics Model}

\newthought{Learn $f_\theta$} to predict:
\begin{equation}
    \hat{s}_{t+1}, \hat{r}_t = f_\theta(s_t, a_t)
\end{equation}

Training is supervised learning on collected transitions $(s, a, r, s')$.

\subsection{Planning with the Model}

\newthought{Model Predictive Control}\index{MPC}:
\begin{enumerate}
    \item At state $s$, simulate many action sequences
    \item Evaluate cumulative reward for each
    \item Execute first action of best sequence
    \item Repeat at next state
\end{enumerate}

\begin{algorithm}[H]
\caption{MPC with Learned Model}
\begin{algorithmic}[1]
\Require Current state $s$, model $f_\theta$, horizon $H$, samples $N$
\For{$i = 1$ to $N$}
    \State Sample action sequence $a_1^i, \ldots, a_H^i$
    \State $\hat{s} \gets s$
    \State $R_i \gets 0$
    \For{$t = 1$ to $H$}
        \State $\hat{s}, \hat{r} \gets f_\theta(\hat{s}, a_t^i)$
        \State $R_i \gets R_i + \gamma^{t-1} \hat{r}$
    \EndFor
\EndFor
\State \Return $a_1^{i^*}$ where $i^* = \arg\max_i R_i$
\end{algorithmic}
\end{algorithm}

\subsection{Model Error and Compounding}\index{model error}

\marginnote{Model errors compound over long horizons.}

\newthought{Small prediction errors accumulate:}
\begin{equation}
    \text{error at step } H \approx H \cdot \epsilon_{\text{single-step}}
\end{equation}

Mitigation strategies include short planning horizons, ensemble models, and frequent re-planning.


\section{World Models}\index{world model}

\marginnote{World models: compress observations into latent space, predict in latent space.}

\subsection{Ha and Schmidhuber (2018)}\index{Ha-Schmidhuber world model}

\newthought{The architecture} has three components~\cite{Ha2018WorldModels}:
\begin{enumerate}
    \item Vision (V): a VAE encodes observations to latent $z$
    \item Memory (M): an RNN predicts future latents
    \item Controller (C): a simple policy on $(z, h)$
\end{enumerate}

\begin{figure}
\centering
\begin{tikzpicture}[scale=0.8]
    \node[draw, rectangle] (obs) at (0,0) {Observation $o_t$};
    \node[draw, rectangle] (vae) at (3,0) {VAE (V)};
    \node[draw, rectangle] (z) at (6,0) {Latent $z_t$};
    \node[draw, rectangle] (rnn) at (6,-2) {RNN (M)};
    \node[draw, rectangle] (ctrl) at (9,-1) {Controller (C)};
    \node[draw, rectangle] (act) at (12,-1) {Action $a_t$};

    \draw[->] (obs) -- (vae);
    \draw[->] (vae) -- (z);
    \draw[->] (z) -- (rnn);
    \draw[->] (z) -- (ctrl);
    \draw[->] (rnn) -- (ctrl);
    \draw[->] (ctrl) -- (act);
    \draw[->, dashed] (rnn.west) to[bend right] node[left] {$h_t$} (rnn.south west);
\end{tikzpicture}
\caption{World Models architecture: V compresses, M predicts, C acts.}
\label{fig:world_models}
\end{figure}

\newthought{The key insight:} train V and M on real experience, then train C entirely in imagination.

\subsection{Dreamer}\index{Dreamer}

\marginnote{\newthought{Dreamer.} State-of-the-art latent world model~\cite{Hafner2020Dreamer}.}

\newthought{Dreamer learns} four components:
\begin{itemize}
    \item Representation model: $p(s_t | s_{t-1}, a_{t-1}, o_t)$
    \item Transition model: $p(s_t | s_{t-1}, a_{t-1})$
    \item Reward model: $p(r_t | s_t)$
    \item Decoder: $p(o_t | s_t)$
\end{itemize}

\newthought{Actor-critic in imagination:} imagine trajectories using the transition model, estimate values using imagined rewards, and update the policy to maximise imagined return.


\section{Exploration and Curiosity}\index{curiosity}\index{intrinsic motivation}

\marginnote{Sparse rewards require intrinsic motivation to explore.}

\subsection{Intrinsic Motivation}

\newthought{Add intrinsic reward} for visiting novel states:
\begin{equation}
    r_{\text{total}} = r_{\text{extrinsic}} + \beta \cdot r_{\text{intrinsic}}
\end{equation}

\subsection{Curiosity-Driven Exploration}\index{ICM}

\newthought{The Intrinsic Curiosity Module}~\cite{Pathak2017Curiosity}:
\begin{equation}
    r_{\text{intrinsic}} = \|f_\theta(s_t, a_t) - s_{t+1}\|^2
\end{equation}

High prediction error indicates a novel state, which yields high reward.

\newthought{The noisy TV problem:} random noise is unpredictable but not interesting. The agent wastes time on inherently stochastic signals.

\subsection{Random Network Distillation}\index{RND}

\newthought{RND}~\cite{Burda2019RND} uses a fixed random network $f$ that maps states to features and a learnable network $\hat{f}$ that tries to match $f$. The intrinsic reward is $\|f(s) - \hat{f}(s)\|^2$. Novel states are hard to match because $\hat{f}$ has not seen them.


\section{Vision-Language Models for Robotics}\index{vision-language model}

\marginnote{VLMs: understand images and text together.}

\subsection{From CLIP to Robotics}

\newthought{Foundation models provide} visual understanding to recognise objects, scenes, and attributes; language grounding to connect words to visual concepts; and common sense as implicit knowledge about the world.

\newthought{The key question:} how to connect perception and language to action?

\subsection{PaLM-E: Embodied Multimodal Reasoning}\index{PaLM-E}

\newthought{PaLM-E} injects visual tokens into a large language model~\cite{Driess2023PaLME}. The model can reason about images and plan actions, outputting natural language plans or action tokens.


\section{Vision-Language-Action Models}\index{VLA}

\marginnote{VLAs: the next step in embodied AI.}

\subsection{RT-1: Robotics Transformer}\index{RT-1}

\newthought{RT-1} takes an image and a language instruction as input and outputs discretised robot actions~\cite{Brohan2022RT1}. It was trained on 130k real robot demonstrations.

\begin{figure}
\centering
\begin{tikzpicture}[scale=0.8]
    \node[draw, rectangle] (img) at (0,1) {Image};
    \node[draw, rectangle] (lang) at (0,-1) {Language};
    \node[draw, rectangle, minimum width=2cm] (enc) at (3,0) {Transformer};
    \node[draw, rectangle] (act) at (6,0) {Actions};

    \draw[->] (img) -- (enc);
    \draw[->] (lang) -- (enc);
    \draw[->] (enc) -- (act);

    \node at (0,2) {\small ``Pick up the apple''};
    \node at (6,1) {\small $(\Delta x, \Delta y, \Delta z, \ldots)$};
\end{tikzpicture}
\caption{VLA architecture: image + language $\to$ transformer $\to$ robot actions.}
\label{fig:vla}
\end{figure}

\subsection{RT-2: VLM as Robot Policy}\index{RT-2}

\newthought{RT-2} fine-tunes a vision-language model for action prediction~\cite{Brohan2023RT2}. Actions are encoded as text tokens, and the model inherits the VLM's reasoning and generalisation.

\subsection{OpenVLA: Open-Source VLA}\index{OpenVLA}

\newthought{OpenVLA} is a 7B parameter model trained on the Open X-Embodiment dataset~\cite{Kim2024OpenVLA}. It supports multiple robot embodiments and provides open weights for research.


\section{Towards Embodied AI}\index{embodied AI}

\marginnote{The ultimate test of intelligence: acting in the physical world.}

\subsection{Sim-to-Real Transfer}\index{sim-to-real}

\newthought{Train in simulation, deploy on a real robot:} domain randomisation varies simulation parameters, system identification matches simulation to reality, and real-world fine-tuning adapts with real data.

\subsection{Open Challenges}

\newthought{Key challenges remain:} safety, because robots can cause harm; generalisation across environments; data scarcity, because robot data is expensive to collect; and multimodality, because touch, sound, and proprioception remain underexplored.


\section{Course Synthesis}

\marginnote{The full arc: from clustering to embodied agents.}

\begin{figure}
\centering
\begin{tikzpicture}[node distance=0.8cm, scale=0.7]
    \node[draw, rectangle] (struct) at (0,0) {Discover Structure};
    \node[draw, rectangle, right of=struct, xshift=2cm] (repr) {Learn Representations};
    \node[draw, rectangle, right of=repr, xshift=2cm] (self) {Self-Supervision};
    \node[draw, rectangle, right of=self, xshift=2cm] (weak) {Weak Supervision};
    \node[draw, rectangle, right of=weak, xshift=2cm] (interact) {Interaction};
    \node[draw, rectangle, right of=interact, xshift=2cm] (embody) {Embodiment};

    \draw[->] (struct) -- (repr);
    \draw[->] (repr) -- (self);
    \draw[->] (self) -- (weak);
    \draw[->] (weak) -- (interact);
    \draw[->] (interact) -- (embody);
\end{tikzpicture}
\caption{The course arc: from structure to embodied intelligence.}
\label{fig:course_arc}
\end{figure}


%═══════════════════════════════════════════════════════════════════════
% BLOCK 3: STUDENT ACTIVATION
%═══════════════════════════════════════════════════════════════════════

\newpage
\section{Examples \& Exercises}

\subsection{Exercise 1: World Model Design}

You want to learn a world model for a driving simulator. Design the architecture:
\begin{enumerate}
    \item What is the input?
    \item What should it predict?
    \item How would you train it?
\end{enumerate}

\subsection{Exercise 2: Curiosity Trade-offs}

A robot explores a room. The TV shows random static. Why might curiosity-driven exploration spend too much time watching the TV?

\subsection{Exercise 3: VLA Advantages}

Why might a VLA trained on internet data generalise better than a robot-only model?


\section{Self-Reflection and Recap}

\newthought{Self-Reflection.}

\begin{enumerate}
    \item Why is model-based RL more sample-efficient but potentially less accurate than model-free RL?
    \item How do VLAs leverage pre-trained vision-language knowledge for robotic control?
    \item What failure modes arise when the world model is inaccurate over long horizons?
    \item How does RND avoid the noisy TV problem that plagues prediction-error curiosity?
    \item Why does Dreamer train the policy entirely in imagination, and what are the risks?
    \item What role does sim-to-real transfer play in making model-based approaches practical?
    \item Looking across the entire course, what is the common thread that connects clustering, VAEs, self-supervised learning, and world models?
\end{enumerate}

\vspace{1.5em}

\newthought{Recap.}

\begin{itemize}
    \item Model-based RL learns a dynamics model and plans in latent imagination; Dreamer achieves state-of-the-art sample efficiency by training actor-critic entirely on imagined trajectories.
    \item Curiosity-driven exploration adds intrinsic reward for visiting novel states; RND avoids the noisy TV problem by measuring novelty through network distillation rather than prediction error.
    \item Vision-Language-Action models unify perception, language understanding, and motor control; RT-2 and OpenVLA inherit web-scale knowledge for robotic generalisation.
    \item The course arc, from discovering structure in data through learned representations, self-supervision, weak supervision, interaction, and finally embodiment, traces a single theme: pattern emerges from data.
\end{itemize}

\vspace{2.5em}

\newthought{We began with the question:} how do we find structure in data without labels? We explored classical clustering and dimensionality reduction, then moved to learned representations with VAEs and self-supervised learning. We saw how weak supervision, self-training, and human feedback can bootstrap learning. Finally, we arrived at reinforcement learning and embodied agents, systems that learn to perceive, reason, and act in the world.

\marginnote{The future belongs to systems that can learn from experience.}

\newthought{The common thread throughout:} structure emerges from data. Whether we discover it through clustering, learn it through reconstruction, or refine it through interaction, unsupervised and self-supervised learning form the foundation of modern AI.

\marginnote{\newthought{feedback}}
