% !TEX root = ../notes_unsupervisedlearning.tex
% Chapter 10: Weak Supervision and Labeling Functions
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Weak Supervision and Labeling Functions}\index{Weak Supervision}\index{Snorkel}
\newthought{What if labels are noisy, partial, or heuristic?}

%═══════════════════════════════════════════════════════════════════════
% BLOCK 1: INTRODUCTION
%═══════════════════════════════════════════════════════════════════════

\section{The Why}

\marginnote{Labels are random variables, not ground truth facts.}

The assumption of clean labels is a fiction: crowd workers make mistakes with 5--20\% error rates, experts disagree on ambiguous cases, heuristics provide cheap but noisy labels, and legacy data carries inconsistent labelling standards.

\newthought{Rather than pretend labels are perfect,} weak supervision\index{weak supervision} embraces noise by modelling label uncertainty explicitly, combining multiple noisy sources, and learning despite imperfect supervision.

\newthought{In order to move from the fiction of perfect labels to principled learning under noise} we have good reason to find ways to,
\begin{itemize}
    \item encode domain knowledge as programmatic labelling functions that trade coverage for precision.
    \item aggregate multiple noisy label sources into a single probabilistic label, accounting for accuracy and correlation.
    \item train discriminative models on probabilistic labels so they generalise beyond the rules that produced them.
\end{itemize}

\subsection{Hands On Experience}

\begin{marginfigure}
\begin{tikzpicture}[scale=0.7]
    \begin{axis}[
        ybar,
        xlabel={Labeling Function},
        ylabel={Votes},
        ymin=0, ymax=10,
        xtick={1,2,3},
        xticklabels={LF1, LF2, LF3},
        width=\textwidth,
        height=0.8\textwidth,
        bar width=0.4cm,
    ]
    \addplot[fill=black!30] coordinates {(1,7) (2,5) (3,8)};
    \addplot[fill=black!60] coordinates {(1,3) (2,5) (3,2)};
    \legend{Positive, Negative}
    \end{axis}
\end{tikzpicture}
\caption{Three labelling functions voting on examples. They disagree.}
\label{fig:lf_votes}
\end{marginfigure}

You want to classify emails as spam or not-spam. You write three rules:
\begin{enumerate}
    \item LF1: Contains ``free money'' $\to$ spam
    \item LF2: From known contact $\to$ not spam
    \item LF3: Contains link + urgency words $\to$ spam
\end{enumerate}

For an email from a known contact about a ``free trial offer'':
\begin{itemize}
    \item LF1 votes: spam, matches ``free''
    \item LF2 votes: not spam, known contact
    \item LF3 votes: abstain, no link
\end{itemize}

How do you combine these conflicting votes?

\newthought{The Learning Objectives} of this lecture:
\begin{itemize}
    \item Design labelling functions that encode domain knowledge and characterise their coverage-accuracy trade-off.
    \item Apply the Snorkel framework to aggregate noisy label sources into probabilistic training labels via a learned label model.
    \item Handle remaining label noise through loss correction, sample weighting, and co-teaching.
\end{itemize}

%═══════════════════════════════════════════════════════════════════════
% BLOCK 2: MAIN THEORY
%═══════════════════════════════════════════════════════════════════════

\newpage
\section{Sources of Weak Labels}\index{weak labels}

\marginnote{Weak labels come from many sources, each with different characteristics.}

\begin{table}
\centering
\begin{tabular}{lcc}
\toprule
\textbf{Source} & \textbf{Coverage} & \textbf{Accuracy} \\
\midrule
Expert annotation & High & High (95\%+) \\
Crowd workers & High & Medium (70--90\%) \\
Heuristic rules & Low--Medium & Variable \\
Distant supervision & High & Low--Medium \\
Pre-trained models & High & Medium--High \\
\bottomrule
\end{tabular}
\caption{Common weak supervision sources.}
\label{tab:weak_sources}
\end{table}

\newthought{Noisy crowd labels} come from multiple annotators with varying quality. Distant supervision\index{distant supervision} derives labels from knowledge bases, for example labelling all sentences mentioning two entities in a relation as expressing that relation. Heuristic rules encode domain knowledge as functions. Data augmentation lets transformed examples inherit parent labels. Auxiliary signals such as metadata, user behaviour, and sensor data provide additional evidence.


\section{Labeling Functions}\index{labeling function}

\marginnote{A labelling function is a noisy voter that may abstain.}

\newthought{A labelling function} $\lambda: X \to Y \cup \{\emptyset\}$ maps inputs to a label or abstention $\emptyset$.

\newthought{Examples for text classification:}
\begin{itemize}
    \item Keyword matching: contains(``urgent'') $\to$ spam
    \item Pattern matching: regex for phone numbers $\to$ contact info
    \item External models: sentiment classifier output
    \item Heuristics: short emails $\to$ not spam
\end{itemize}

\subsection{Properties of Good Labeling Functions}

\newthought{Good labelling functions} have high precision, so that when they vote they are usually right; reasonable coverage, so that they vote on enough examples; diversity, so that different LFs capture different patterns; and low correlation, so that their errors are independent.

\begin{figure}
\centering
\begin{tikzpicture}[scale=0.8]
    \begin{axis}[
        xlabel={Coverage},
        ylabel={Accuracy},
        xmin=0, xmax=1,
        ymin=0.5, ymax=1,
        width=0.6\textwidth,
        axis line style={draw=none},
        tick style={black, thin},
        tick align=outside, tick pos=left,
    ]
    \addplot[only marks, mark=*, mark size=4pt] coordinates {
        (0.1, 0.95) (0.3, 0.85) (0.5, 0.75) (0.8, 0.65)
    };
    \node at (axis cs:0.15, 0.91) {\small LF1};
    \node at (axis cs:0.35, 0.81) {\small LF2};
    \node at (axis cs:0.55, 0.71) {\small LF3};
    \node at (axis cs:0.85, 0.61) {\small LF4};
    \draw[dashed, black!40] (axis cs:0,0.8) -- (axis cs:1,0.8);
    \node[black!55] at (axis cs:0.9, 0.82) {\small threshold};
    \end{axis}
\end{tikzpicture}
\caption{Labelling functions trade off coverage and accuracy.}
\label{fig:lf_tradeoff}
\end{figure}


\section{Label Modeling}\index{label model}

\marginnote{The label model estimates true labels from noisy votes.}

\subsection{Majority Vote}

\newthought{The simplest aggregation} is majority vote:
\begin{equation}
    \hat{y} = \text{mode}(\{\lambda_i(x) : \lambda_i(x) \neq \emptyset\})
\end{equation}

This assumes equal accuracy and ignores correlations.

\subsection{Modeling Labeler Accuracy}

\newthought{Treating each LF} as having accuracy $\alpha_i = P(\lambda_i(x) = y | \lambda_i(x) \neq \emptyset)$ gives a weighted vote:

\begin{equation}
    P(y | \lambda_1, \ldots, \lambda_m) \propto P(y) \prod_{i: \lambda_i \neq \emptyset} P(\lambda_i | y)
\end{equation}

\subsection{Modeling Correlations}

\newthought{Labelling functions may have correlated errors,} for example two keyword-based LFs that fail on the same examples.

A factor graph model captures pairwise correlations:
\begin{equation}
    P(y, \lambda_1, \ldots, \lambda_m) = \frac{1}{Z} \phi_Y(y) \prod_i \phi_i(y, \lambda_i) \prod_{i,j} \phi_{ij}(\lambda_i, \lambda_j)
\end{equation}

where $\phi_{ij}$ captures the pairwise dependency structure.


\section{The Snorkel Framework}\index{Snorkel}

\marginnote{\newthought{Snorkel.} End-to-end weak supervision system~\cite{Ratner2017Snorkel}.}

\begin{figure}
\centering
\begin{tikzpicture}[node distance=1.5cm]
    \node[draw, rectangle] (data) {Unlabeled Data};
    \node[draw, rectangle, right of=data, xshift=1.5cm] (lfs) {Labeling Functions};
    \node[draw, rectangle, right of=lfs, xshift=1.5cm] (lm) {Label Model};
    \node[draw, rectangle, right of=lm, xshift=1.5cm] (prob) {Probabilistic Labels};
    \node[draw, rectangle, right of=prob, xshift=1.5cm] (dm) {Discriminative Model};

    \draw[->] (data) -- (lfs);
    \draw[->] (lfs) -- (lm);
    \draw[->] (lm) -- (prob);
    \draw[->] (prob) -- (dm);
\end{tikzpicture}
\caption{Snorkel pipeline: LFs $\to$ Label Model $\to$ Probabilistic Labels $\to$ Train End Model.}
\label{fig:snorkel_pipeline}
\end{figure}

\begin{algorithm}[H]
\caption{Snorkel Workflow}
\begin{algorithmic}[1]
\State \textbf{Write} labeling functions encoding domain knowledge
\State \textbf{Apply} LFs to unlabeled data $\to$ label matrix $\Lambda$
\State \textbf{Train} label model on $\Lambda$ to estimate LF accuracies and correlations
\State \textbf{Generate} probabilistic labels $\tilde{y} = P(y | \Lambda)$
\State \textbf{Train} discriminative model on $(X, \tilde{y})$
\end{algorithmic}
\end{algorithm}

\newthought{The key insight} is that the discriminative model generalises beyond the labelling functions; it learns features, not just rules.


\section{Learning with Label Noise}\index{label noise}

\marginnote{Even with probabilistic labels, training can be noisy.}

\subsection{Forward and Backward Loss Correction}

\newthought{If noise is class-conditional} with known transition matrix $T$ where $T_{ij} = P(\tilde{y} = j | y = i)$:

Forward correction:
\begin{equation}
    \tilde{\ell}(f(x), \tilde{y}) = \ell(f(x), T^T f(x))
\end{equation}

Backward correction:
\begin{equation}
    \tilde{\ell}(f(x), \tilde{y}) = T^{-1} \ell(f(x), \tilde{y})
\end{equation}

\subsection{Sample Weighting}

\newthought{Down-weighting likely-mislabelled examples} gives:
\begin{equation}
    \mathcal{L} = \sum_i w_i \ell(f(x_i), \tilde{y}_i)
\end{equation}

where $w_i$ is based on loss value, prediction confidence, or label model confidence.

\subsection{Co-Teaching}\index{co-teaching}

\newthought{Co-teaching} trains two networks~\cite{Han2018CoTeaching}; each selects small-loss samples for the other:
\begin{itemize}
    \item Network A selects clean samples for Network B
    \item Network B selects clean samples for Network A
    \item Both avoid memorising the same noisy labels
\end{itemize}


\section{Association Rules as Labeling Functions}\index{association rules}

\marginnote{From Apriori to labelling: mine rules from data structure.}

\newthought{Association rule learning} and labelling functions share a common structure: association rules $\{A, B\} \to \{C\}$ are scored by support and confidence, while labelling functions map $\text{condition}(x) \to \text{label}$.

Mining frequent patterns can automatically generate labelling functions:
\begin{enumerate}
    \item Find frequent itemsets in feature space
    \item Convert to rules with confidence threshold
    \item Use as labelling functions
\end{enumerate}


%═══════════════════════════════════════════════════════════════════════
% BLOCK 3: STUDENT ACTIVATION
%═══════════════════════════════════════════════════════════════════════

\newpage
\section{Examples \& Exercises}

\subsection{Exercise 1: Write Labeling Functions}

Design 5 labelling functions for classifying product reviews as positive or negative.

\subsection{Exercise 2: Label Aggregation}

Three LFs vote on an example:
\begin{itemize}
    \item LF1: positive, accuracy 0.9
    \item LF2: negative, accuracy 0.8
    \item LF3: positive, accuracy 0.7
\end{itemize}

Assuming independent LFs and uniform prior, compute $P(y = \text{positive})$.

\subsection{Exercise 3: LF Correlation}

Two LFs both use keyword matching. Why might their errors be correlated? How does this affect the label model?


\section{Self-Reflection and Recap}

\newthought{Self-Reflection.}

\begin{enumerate}
    \item When is weak supervision preferable to collecting clean labels?
    \item How do you know if your labelling functions are good enough?
    \item What happens to the label model when all labelling functions are correlated?
    \item Why does the discriminative model in Snorkel generalise beyond the labelling functions themselves?
    \item How does co-teaching avoid the confirmation bias that a single network would accumulate?
    \item What are the risks of distant supervision, and how would you mitigate them?
    \item Could you combine weak supervision with active learning, and if so, how?
\end{enumerate}

\vspace{1.5em}

\newthought{Recap.}

\begin{itemize}
    \item Weak supervision embraces label noise by modelling labeller accuracy and correlation rather than treating annotations as ground truth.
    \item Labelling functions encode domain knowledge as noisy voters; the coverage-accuracy trade-off determines their utility in the label model.
    \item Snorkel aggregates multiple noisy sources into probabilistic labels and trains a discriminative model that generalises beyond the rules.
    \item Noise-robust training via loss correction, sample weighting, and co-teaching handles residual label errors after aggregation.
\end{itemize}

\vspace{2.5em}

\marginnote{\newthought{Teaser.} What if the model itself becomes the labeller, using its own confident predictions as training signal?}

\newthought{We have used heuristics and noisy external sources for labels.} But labelling functions must be written by hand, and their coverage is inherently limited. A natural next step is to let the model label the data it is most confident about, then retrain on those self-generated labels.

\newthought{The next chapter explores self-training,} where the model's own predictions become pseudo-labels, and consistency regularisation forces stable predictions under augmentation.

\marginnote{\newthought{feedback}}
