%% meta.tex -- Shared metadata and templates for lecture notes
%% Author: Dr.-Ing. Mark Schutera

\InputIfFileExists{version.tex}{}{}
\providecommand{\docversionmajor}{0}
\providecommand{\docversionminor}{0}
\providecommand{\docversionpatch}{0}
\newcommand{\docversion}{\docversionmajor.\docversionminor.\docversionpatch}

%%
% Default author information
%%
\newcommand{\defaultauthor}{Prof. Dr.-Ing. Mark Schutera}
\newcommand{\defaultpublisher}{Unfinished Lecture Notes}

%%
% Copyright page template
%%
\newcommand{\makecopyright}[1][]{%
  \newpage
  \begin{fullwidth}
  ~\vfill
  \thispagestyle{empty}
  \setlength{\parindent}{0pt}
  \setlength{\parskip}{\baselineskip}
  Copyright \copyright\ \the\year\ \thanklessauthor
  
  \par\smallcaps{Published by \thanklesspublisher}
    
\par Combobulated with the help of multiple large language model driven tools. Licensed under the Creative Commons Attribution-NonCommercial 4.0 International License (``CC BY-NC-SA 4.0''). 
You may not use this file for commercial purposes. If you remix, transform, or build upon the material, you must distribute your contributions under the same license as the original
You must obtain explicit permission from the author for uses beyond those permitted by this license. 
To view a copy of this license, visit \url{https://creativecommons.org/licenses/by-nc-sa/4.0/}. 
Unless required by applicable law or agreed to in writing, distributed material is provided on an 
\smallcaps{``AS IS'' BASIS, WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND}, either express or implied. 
See the license for details.\index{license}

  \par\textit{Last updated, \monthyear\ --- v\docversion}
  \end{fullwidth}
}

%%
% Standard front matter structure
%%
\newcommand{\makestandardfrontmatter}{%
  \frontmatter
  \maketitle
  \makecopyright
}

%%
% Standard dedication page template
%%
\newcommand{\makededication}[1]{%
  \cleardoublepage
  ~\vfill
  \begin{doublespace}
  \noindent\fontsize{18}{22}\selectfont\itshape
  \nohyphenation
  #1
  \end{doublespace}
  \vfill
  \vfill
}

%%
% Standard introduction template
%%
\newcommand{\makeintroduction}[1]{%
  \cleardoublepage
  \chapter*{Introduction}
  #1
}

%%
% Philosophy of education
%%
\newcommand{\makephilosophy}{%
  \newpage
  \chapter*{A Philosophy of Education}
  \newthought{The beautiful risk of education} is that it is an encounter.
  I am trying to model a way of thinking, and reasoning, then stepping back.
  The rest is yours.

  \vspace{2.5em}



  \begin{itemize}
    \item \textbf{Start with why.} A method understood only procedurally is a method you will misapply when it matters.
    Before learning how something works, understand why it was built, what it replaced, and where it fails.
    The mechanics then follow from necessity, not memorization.
    \item \textbf{Build taste alongside competence.} You need a sense of what beautiful is,
    you need to train your aesthetics before you can reliably produce it.
    This develops by studying what others have built, making it your own, then making your own, and honestly evaluating the gap between them.
    Taste is the compass. Skill is the engine. You need both, in that order.
    \item \textbf{Failure is the designed medium.} Each attempt teaches what a clean run cannot.
    The insight compounds across tries, not from a single correct execution.
    Those who never start failing will fail.
    \item \textbf{Work in the open.} Showing half-finished thinking, naming mistakes as they happen, is a professional skill.
    We practice it here. It is how knowledge actually moves between people. And it is how knowledge moves towards you, 
    at unprecedented speed.
    \item \textbf{The fallacy of mediocracy.} Iteration and failure are the method, not the excuse for settling.
    Do not confuse velocity with quality. Whatever you learn, whatever you build: make it good.
    \item \textbf{Outcomes emerge. They are not delivered.} Judgment, taste, the capacity to sit with uncertainty:
    these are not course outputs. They are what you build from the encounter.
    I can create the conditions. The rest depends on what you do with them.
  \end{itemize}
}

%%
% Page header: show author name on every page (verso and recto).
% Must patch \mainmatter because tufte-book resets \fancyhead inside it.
%%
\let\tufte@orig@mainmatter\mainmatter
\renewcommand{\mainmatter}{%
  \tufte@orig@mainmatter
  \fancyhead[LE]{\thepage\quad\smallcaps{\newlinetospace{\plaintitle}}\quad\smallcaps{\defaultauthor}}%
  \fancyhead[RO]{\smallcaps{\defaultauthor}\quad\smallcaps{\newlinetospace{\plaintitle}}\quad\thepage}%
}

%%
% Font options template
%%
\newcommand{\loadoptionalfonts}{%
  % Font options (uncomment if available)
  %\IfFileExists{bergamo.sty}{\usepackage[osf]{bergamo}}{}% Bembo
  %\IfFileExists{chantill.sty}{\usepackage{chantill}}{}% Gill Sans
  %\usepackage{microtype}
}