%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% How to Write a Scientific Thesis
% Author: Dr.-Ing. Mark Schutera
% Based on Tufte-style layout
%\makeindex % disabled to avoid requiring makeindex/.ind during quick builds
% \index{} entries are present throughout all chapters and are ready to use;
% re-enable by uncommenting the line above and running makeindex.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass{../sharedAssets/tufte}

%%
% Book metadata
\title{Philosophy of Artificial Intelligence}
\author[\defaultauthor]{\defaultauthor}
\publisher[\defaultpublisher]{\defaultpublisher}

% Load optional fonts
\loadoptionalfonts

% Additional packages
\usepackage{pifont}
\usepackage{amsmath}
\usepackage{soul}
\usepackage{textcase}
\usepackage{enumitem}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{float}
\usepackage{qrcode}
\usepackage{algorithm}
\usepackage{algpseudocode}

\extrafloats{100}

\begin{document}



\makestandardfrontmatter
% Dedication
% Alternative quotes:
\makededication{\openepigraph{“The Answer to the Great Question... Of Life, the Universe and Everything... Is... Forty-two,' said Deep Thought, with infinite majesty and calm.”}{The Hitchhiker's Guide to the Galaxy, Douglas Adams}}
\tableofcontents


% Introduction
\makeintroduction{%
  Artificial intelligence is no longer a future concern, it is a present societal,
  political and above all one for human intelligence. Algorithms decide who receives credit, who is flagged at
  a border, who gets a job interview, which cornflakes you buy and so increasingly every decision is touched by the invisible hand of prediction. Yet the concepts we use to govern these
  systems, responsibility, personhood, consent, democratic legitimacy were built for a world of human actors and institutions. This course asks what happens when those concepts collide with machines
  that appear to think, feel, and decide. 

  \newthought{The course name} places Philosophy in front, reflecting our approach to the subject under the impression of the love of wisdom. 
  We are not here to advocate for a particular position or policy outcome, but to equip ourselves with the tools of critical thinking, 
  argumentation, and empathy that will allow us to navigate the complex landscape of AI and its futures. We approach the question not
  through technical optimism or dystopian panic, but through the oldest
  tools available to us: philosophy, fiction, and argument.


  \newthought{The course} is structured around four themes. \textbf{Anthropomorphism}
  asks why we make machines in our image, why we tend to anthropomorphize them, and what obligations that
  creates. \textbf{Accountability} asks who answers when an autonomous
  system causes harm, or for that matter is beneficial. \textbf{Embodiment} asks whether a physical
  presence changes the ethical stakes and responsibilities. \textbf{Agency} asks
  about the role and implications of autonomous decision-making, the concept of alignment and human oversight.

  


  \newthought{Each session} pairs a canonical literary text: Hoffmann, Kafka,
  Kleist, \v{C}apek, Goethe, with a real contemporary question of AI policy and ideology. 
  These texts are thought-experiments mendeled out over centuries of readership, and they surface
  intuitions and a timeliness in their thought leadership that temporary texts barely can reach.
  From these texts we draw out the philosophical questions and ethical dilemmas that are still relevant today, 
  and we use them as a springboard for discussion and debate. The goal is not consensus. It is the capacity to establish a position, hold a position,
  defend a position, give up a position, listen to opposing views, and update your views and positions accordingly. 
  That is what a present with infinitely many futures demands and requires of its humans and humanity.\footnote{These notes are, by their very nature, unfinished. The field moves fast,
  and the arguments deserve to be sharpened by more voices than one. If you spot an error, disagree with a framing, or want to add a source, a question, or a debate position, open a pull request and share your thoughts at \url{https://github.com/Quillstacks/LectureMaterial/tree/main/lecturenotes/notes_philosophyofai}. Every contribution is welcome.}

  \marginnote{%
    \begin{flushright}
    {\color{black}\qrcode[height=1.2cm]{https://github.com/Quillstacks/LectureMaterial/tree/main/lecturenotes/notes_philosophyofai}}
    \end{flushright}%
  }

  \newthought{Lecture Outline} 
  \begin{itemize}
    \item 40 minutes - Literature introduction and guided reading sheet 
    \item 40 minutes - Debate with position exposure
    \item 10 minutes - Reflection on the discussion and key-learnings
    \item Teaser for next session.
  \end{itemize}
}





\mainmatter
\authorinrectoheader

\input{chapters/01_Anthropomorphism.tex}
\newpage
\input{chapters/02_Accountability.tex}
\newpage
\input{chapters/03_Embodiment.tex}
\newpage
\input{chapters/04_Agency.tex}
\newpage

% Appendix
\appendix

\input{chapters/aa_PreparationGuidance.tex} % → Appendix A
\newpage
\input{chapters/ac_DebateProtocol.tex}   % → Appendix B


\bibliographystyle{abbrvnat}
\bibliography{philosophyofai_references}



% Hey Mark this is for you, when thinking about where to take this next.
% Think Ethics for Nerds
% Think Book on The Practice of Philosophy a handbook for beginners.
% Think about inviting guest speakers to talk about their work and how it connects to the themes of the course.

\end{document}
